### 2024-04-05 R/eda-code/EDA_unified.qmd
replacing corrupt file

---
title: "Class Project"
author: Andrew Ruiz
date: "2024-03-13"
output: html_document
---

# load needed packages. make sure they are installed.
```{r}
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(readr)
library(forecast)
library(tsibble) # For time series data wrangling
library(feasts) # For decomposing and visualizing time series
library(RColorBrewer)
```

# This file will contain the EDA code for the class project.
### As the project progresses and the models finalized, the EDA QMD will be updated accordingly.

## MIR heatmap
### This heat map illustrates the timing and intensity of mosquito infection rates (MIR) over the weeks of the year.
#### At least one previous study that a non-zero MIR at or before week 31 is common in years when human cases of EEE are reported. 
```{r}
# Load the saved function
create_heatmap <- readRDS(here("data", "processed-data", "rds","create_heatmap_function.rds"))

# Assuming MIR_by_week is loaded and prepared in this QMD as well
# Now use the loaded function to create the heatmap
MIR_by_week_heatmap <- create_heatmap(MIR_by_week)

# Display the heatmap
MIR_by_week_heatmap
```

## This is a scatter plot of the average number of mosquitoes that were captured in light traps and the mosquito infection rate (MIR).
```{r}
# Load the function from the RDS file
MIRMEL_scatter_plot <- readRDS(here("data", "processed-data", "RDS", "create_scatter_plot_function.rds"))

# Assuming combined_data_mir_all is available in your current session
# Generate the plot
scatter_plot <- create_scatter_plot(combined_data_mir_all)

# Display the plot
print(scatter_plot)

```

### 2024-04-05 R/processing-code/MADAproject_pt2_ruiz.qmd
replacing corrupt file

---
title: "MADA Project Part 2"
subtitle: "loading, cleaning, exploring the data"
author: "Andrew Ruiz"
date: 2024-02-23
format: html
editor: source
---

# Introduction

## This file contains the process for loading, cleaning and exploring the datasets.

### Use this code to ensure you have all the packages installed before proceeding.
#### This code will check if the packages are installed and install them if they are not.
```{r}
# List of packages required for the project
required_packages <- c(
  "readxl", "httr", "dplyr", "tidyr", "skimr", "here", "readr",
  "naniar", "ggplot2", "sf", "topicmodels", "openxlsx", "lubridate",
  "knitr", "kableExtra", "webshot2", "magick", "leaflet", "htmlwidgets",
  "suncalc", "stringr", "ISOweek"
)

# Function to check and install missing packages
install_missing_packages <- function(packages) {
  for (pkg in packages) {
    if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
      install.packages(pkg)
      library(pkg, character.only = TRUE)
    }
  }
}

# Install missing packages
install_missing_packages(required_packages)
```

# Loading libraries
## This list contains the specific versions used in the script.
```{r, message=FALSE, warning=FALSE}
library(readxl) # for loading Excel files (version 1.4.0)
library(httr) # for getting data from the web (version 1.4.2)
library(dplyr) # for data processing/cleaning (version 1.0.7)
library(tidyr) # for data processing/cleaning (version 1.1.3)
library(skimr) # for visualizing data (version 2.1.3)
library(here) # allows relative pathways for data (version 1.0.1)
library(readr) # reading and writing data to and from CSV, TSV, and fixed-width files (version 2.0.1)
library(naniar) # tools to work with missing data (version 0.6.1)
library(ggplot2) # plot creation (version 3.3.5)
library(sf) # standardized way to encode spatial data (version 1.0-2)
library(topicmodels) # tools for topic modeling on text data (version 0.2-12)
library(openxlsx) # Facilitates reading from and writing to .xlsx files (version 4.2.4)
library(lubridate) # Simplifies the work with dates and times (version 1.7.10)
library(knitr) # Allows for dynamic report generation in R (version 1.33)
library(kableExtra) # generates simple HTML or LaTeX tables (version 1.3.4)
library(webshot2) # tools to take screenshots of web pages (version 0.0.2)
library(magick) # advanced image processing in R (version 2.7.3)
library(leaflet) # Enables the creation of interactive maps (version 2.0.4.1)
library(htmlwidgets) # Facilitates the creation of R bindings to JavaScript libraries (version 1.5.4)
library(suncalc) # Calculates sunset and sunrise times based on location (version 0.5.1)
library(stringr) # Makes string manipulation easier and more consistent (version 1.5.1)
library(ISOweek) # Calculates week number according to ISO 8601 standards (version 0.6-2)
library(purrr)
```

# Loading data
## Mosquito trap data
### This dataset contains trap counts by species for all mosquito traps set in the Bristol County from 2007-2023 this dataset will undergo extensive processing below.
#### The dataset was obtained though and open records request with the Commonwealth of MA.
```{r trap-data-import}
# Construct the path to the Excel file
#file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")
file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")
file_exists_check <- file.exists(file_path_mos)
file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")

#file_exists_check <- file.exists("/Users/andrewruiz/MADA_course/RUIZ-MADA-project/data/raw-data/MOSQ_MADA.xlsx")
print(file_exists_check)  # Should return TRUE if the path is correct
print(file_path_mos)

print(file_exists_check)  # This should return TRUE if the file path is correct
# import the data to R and store it in a dataframe named "mosquito"
mosquito_raw <- read_excel(file_path_mos)

#load the first few rows of data to verify the data loaded as expected
head(mosquito_raw)
```

## Virus testing of the trap count data
### This dataset is identical to MOSQ_MADA except that it contains virus test results.
```{r}
# Read the dataset
pcr_data <- read_excel(here("data", "raw-data", "virus_iso.xlsx"))

# Add week number based on the collection data
# This code ensures that the ISO 8601 standard is applied to the week number calculation
# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
# Weeks start on Monday and end on Sunday.
# The last week of the year, week 52 or 53, is the one that contains December 28th.
# we will use week_num to join tables
# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08
pcr_data <- pcr_data %>%
  mutate(
    # Directly format the collection date into "YYYY-WW"
    week_num = paste0(year(`Collection Date`), "-", 
                      sprintf("%02d", isoweek(`Collection Date`)))
  )

# Correctly expand "Both" in Test Type, then adjust to "EEE" and "WNV"
pcr_data_expanded <- pcr_data %>%
  mutate(Test_Type_Expanded = case_when(
    `Test Type` == "Both" ~ "EEE&WNV", # Temporarily mark "Both" for special handling
    TRUE ~ `Test Type`
  )) %>%
  separate_rows(Test_Type_Expanded, sep = "&") %>%
  mutate(
    Year = year(`Collection Date`),
    Week = week(`Collection Date`),
    Submitted_for_Testing = `Submitted for Testing` == "Y",
    # Assuming positive results are marked as "Positive" in `Result`
    Is_Positive = Result == "Positive"
  )

# Save as RDS
saveRDS(pcr_data_expanded, here("data", "processed-data", "rds", "pcr_data_expanded.rds"))

# Save as CSV
write_csv(pcr_data_expanded, here("data", "processed-data", "pcr_data_expanded.csv"))

```

## Mosquito trap sites
### Location for the mosquito trap sites
```{r}
# Construct the path to the file
file_path_trap = here("data", "processed-data", "BCMCP_trap_sites_2021.xlsx")

# import the data to R and store it in a dataframe named "trap_location"
trap_location = read_excel(file_path_trap)

#load the first few rows of data to verify the data loaded as expected
head(trap_location)

#The last 3 columns have missing data. However, they are not important for the purposes of this project
# there is not a need to delete the records
# this is a free text field that provides notes about the location.
# they are not needed so we will delete them
trap_location_selected <- trap_location %>%
  select(`Trap Site Address`, MCD, Location, DecLat, DecLong)
trap_location_selected

# Define the path for the trap locations selected dataset RDS
trap_location_selected_path_rds <- here("data", "processed-data", "rds", "trap_sites_selected.rds")

# Save the trap_location_selected dataframe as an RDS file
saveRDS(trap_location_selected, trap_location_selected_path_rds)
```

## Mosquito virus data

### The data includes all the positive PCR results for EEE and WNV for all mosquitoes submitted for testing

### These files were obtained from: https://www.mass.gov/lists/arbovirus-surveillance-plan-and-historical-data

### The virus isolation data is stored in separate spreadsheets by year from 2014 to 2020

### This code will load all the files, combine them, and then save the combined file in the "raw data" folder

```{r}
#Define the path to the folder with Excel files
folder_path <- here("data", "processed-data", "virus_data")

#List all Excel files (.xls and .xlsx)
file_paths <- list.files(path = folder_path, pattern = "\\.xls[x]*$", full.names = TRUE)

#Read each file into a list of data frames
data_list <- lapply(file_paths, read_excel)

#Check the names of the files read
print(basename(file_paths))

#Combine the datasets
combined_virus_data <- bind_rows(data_list)

# Define the path for the combined dataset
combined_path <- here("data", "processed-data", "combined_virus_data.csv")

# Save the combined data frame as a CSV file
write.csv(combined_virus_data, combined_path, row.names = FALSE)

# Define the path for the combined dataset RDS
combined_path_rds <- here("data", "processed-data", "rds", "combined_virus_data.rds")

# Save the combined data frame as an RDS file
saveRDS(combined_virus_data, combined_path_rds)

# View the first few rows to confirm it's loaded correctly
head(combined_virus_data)
```

## Weather data
### temperature and precipitation are important factors in mosquito population dynamics
### This dataset contains daily weather summaries from the Taunton, MA Airport from 2007-2023
#### The data was obtained from the NOAA National Centers for Environmental Information
```{r}
# Construct the path to the file
file_path_wx = here("data", "processed-data", "weather_airport.csv")

# import the data to R and store it in a dataframe named "wx"
wx = read.csv(file_path_wx)

wx <- wx %>%
  mutate(
    DATE = mdy(DATE),  # Parse the DATE column from "month/day/year"
    week_num = paste0(year(DATE), "-", sprintf("%02d", isoweek(DATE)))  # Create week_num
  )
#load the first few rows of data to verify the data loaded as expected
head(wx)

#the columns below are the only ones needed for this project. 
wx_selected <- wx %>%
  dplyr::select(STATION, NAME, DATE, PRCP, TMAX, week_num)

# Define the path for the RDS file
output_path_wx <- here("data", "processed-data", "rds", "wx_selected.rds")
output_csv_wx <- here("data", "processed-data", "wx_selected.csv")
# Save the wx_selected dataframe as an RDS file
saveRDS(wx_selected, output_path_wx)
write.csv(wx_selected, output_csv_wx, row.names = FALSE)
head(wx_selected)
```

# hourly weather
## This dataset contains hourly weather data from the Taunton, MA Airport from 2007-2023
### The data was obtained from the NOAA National Centers for Environmental Information
```{r}
hourly_wx_path <- here("data", "raw-data", "hourly_wx.csv")

# Ensure hourly_wx is read if not already done
hourly_wx <- read_csv(hourly_wx_path, show_col_types = FALSE)

hourly_wx_clean <- hourly_wx %>%
  mutate(Year = year(DATE),
         ISO_Week = isoweek(DATE),
         week_num = sprintf("%d-%02d", Year, ISO_Week)) %>%
  filter(REPORT_TYPE...3 != "SOD", REPORT_TYPE...3 != "SOM", REPORT_TYPE...3 != "AUTO") %>%
  select(STATION, DATE, HourlyDryBulbTemperature, HourlyRelativeHumidity, HourlyWetBulbTemperature, HourlyWindSpeed, week_num)

str(hourly_wx_clean)

# Define the path for the RDS and CSV files
rds_path_hourly_wx <- here("data", "processed-data", "rds", "hourly_wx_clean.rds")
csv_path_hourly_wx <- here("data", "processed-data", "hourly_wx_clean.csv")

# Save the cleaned dataframe as RDS and CSV files
saveRDS(hourly_wx_clean, rds_path_hourly_wx)  # Save as RDS
write.csv(hourly_wx_clean, csv_path_hourly_wx, row.names = FALSE) # Save as CSV

```
## Drought data
### These datasets were obtained from the US Drought Monitor
#### The U.S. Drought Monitor is produced through a partnership between the National Drought Mitigation Center at the University of Nebraska-Lincoln, the United States Department of Agriculture and the National Oceanic and Atmospheric Administration
##### This data is organized by geographic area and week
##### It includes 4 categories. D4 - Exceptional Drought (0-2 percentile), D3 - Extreme Drought (2-5 percentile), D2 - Severe Drought (5-10 percentile), D1 - Moderate Drought (10-20 percentile), D0 - Abnormally Dry (20-30 percentile), None or 0 - Neither Wet or Dry
##### It provides the percent area covered by each category. 
```{r}
# read file with drought data by county for MA, RI, CT for the years 2007-2023
drought_ma_ri_ct = read_csv(here("data", "raw-data", "comp_stat_county_ma_ri_ct.csv"), show_col_types = FALSE)

# read file with drought data by county for FEMA Region 4 for the years 2007-2023
drought_fema4 = read_csv(here("data", "raw-data", "comp_stat_fema4.csv"), show_col_types = FALSE)

# Add week number based on MapDate for both files
# This code ensures that the ISO 8601 standard is applied to the week number calculation
# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
# Weeks start on Monday and end on Sunday.
# The last week of the year, week 52 or 53, is the one that contains December 28th.
# we will use week_num to join tables
# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08
drought_ma_ri_ct <- drought_ma_ri_ct %>%
  mutate(
    # First, ensure MapDate is in a proper date format
    MapDate = as.Date(as.character(MapDate), format = "%Y%m%d"),
    # Then, format the collection date into "YYYY-WW"
    week_num = paste0(year(MapDate), "-", 
                      sprintf("%02d", isoweek(MapDate)))
  )

drought_fema4 <- drought_fema4 %>%
  mutate(
    # First, ensure MapDate is in a proper date format
    MapDate = as.Date(as.character(MapDate), format = "%Y%m%d"),
    # Then, format the collection date into "YYYY-WW"
    week_num = paste0(year(MapDate), "-", 
                      sprintf("%02d", isoweek(MapDate)))
  )

# Define the path for the RDS and CSV files
rds_path_drought_ma_ri_ct<- here("data", "processed-data", "rds", "drought_ma_ri_ct.rds")
csv_path_drought_ma_ri_ct <- here("data", "processed-data", "drought_ma_ri_ct.csv")
# Save the wx_selected dataframe as an RDS and CSV files
saveRDS(drought_ma_ri_ct, rds_path_drought_ma_ri_ct)
write.csv(drought_ma_ri_ct, csv_path_drought_ma_ri_ct, row.names = FALSE)

# Define the path for the RDS and CSV files
rds_path_drought_fema4<- here("data", "processed-data", "rds", "drought_fema4.rds")
csv_path_drought_fema4 <- here("data", "processed-data", "drought_fema4.csv")
# Save the wx_selected dataframe as an RDS and CSV files
saveRDS(drought_fema4, rds_path_drought_fema4)
write.csv(drought_fema4, csv_path_drought_fema4, row.names = FALSE)
```

## Human and veterinary cases of eastern equine encephalitis and West Nile virus

### these were extracted from annual mosquito reports publised here: https://www.mass.gov/lists/arbovirus-surveillance-plan-and-historical-data

### human cases are listed by county only. animal cases are listed by town/city (with the exception of 1 year)

```{r}
# Construct the path to the human case data file
file_path_human <- here("data", "processed-data", "human_cases.xlsx")

# import the data to R and store it in a dataframe named "cases_human"
cases_human <- read_excel(file_path_human)

# Save 'cases_human' as an RDS file
output_path_rds_human <- here("data", "processed-data", "rds", "human_cases_processed.rds")
saveRDS(cases_human, output_path_rds_human)

#load the first few rows of data to verify the data loaded as expected
head(cases_human)

# Construct the path to the veterinary case data file
file_path_vet <- here("data", "processed-data", "animal_cases.xlsx")

# import the data to R and store it in a dataframe named "cases_human"
cases_vet <- read_excel(file_path_vet)

# Save 'cases_vet' as an RDS file
output_path_rds_vet <- here("data", "processed-data", "rds", "vet_cases_processed.rds")
saveRDS(cases_vet, output_path_rds_vet)

#load the first few rows of data to verify the data loaded as expected
head(cases_vet)
#date of onset is missing for some records, however, they are not needed for this project. 
```

# Processing and cleaning the Mosquito trap data

### Examine the variables and change all column names to lower case

```{r}
str(mosquito_raw)

# Set column names to lowercase
names(mosquito_raw) <- tolower(names(mosquito_raw))

# Check the first few rows to confirm changes
head(mosquito_raw)
```

## we will examine the dataset before making changes.

### first we will Identify missing data

```{r}
sapply(mosquito_raw, function(x) sum(is.na(x)))
```

#### The majority of the missing data is in the "Submitted for Testing" variable. This is not a crucial variable, so we can keep those records. Missing data in that variable generally means 0 mosquitoes were captured in a trap. The other variable are important and so we will remove those records with the other missing data.

### The Massachussetts Department of Public Health adpoted a taxonomy change before it was fully vetted. The change divided the genus *Aedes* into a new genus: *Ochlerotatus*. To align with accepted standards, I will change them back to *Aedes*.

```{r}
# Get unique values using dplyr
unique_genus_values <- mosquito_raw %>%
  distinct(genus) %>%
  pull(genus)

# Print the unique genus values
print(unique_genus_values)

# Change records where genus is "Ochlerotatus" to "Aedes"
mosquito_raw$genus[mosquito_raw$genus == "Ochlerotatus"] <- "Aedes"

# Check the first few rows to confirm changes, or use unique() to see the change in genus values
unique(mosquito_raw$genus)
```

### Execute proposed changes.

```{r}
# Create a copy of the dataframe to avoid modifying the original
mosquito_modified <- mosquito_raw

# Change records where genus is "Ochlerotatus" to "Aedes" to adhere to current taxonomy standards
mosquito_modified$genus[mosquito_modified$genus == "Ochlerotatus"] <- "Aedes"

# Ensure Collection Date is in Date format 
mosquito_modified$`collection date` <- as.Date(mosquito_modified$`collection date`)

# Convert all species code values to uppercase
# This will minimize issues were data as entered incorrectly
# for example Mel, mel, and MEL
mosquito_modified <- mosquito_modified %>%
  mutate(`species code` = toupper(`species code`))

# Remove rows with NA values in the remaining columns of the modified dataframe, 
# except for "Submitted for Testing" which we'll handle separately
# when Submitted for testing is NA or missing, it indicates that there were no mosquitoes captured during a trap event.
# This is common at the beginning and end of season and also when a trap malfunctions
temp_columns <- names(mosquito_modified)
temp_columns <- temp_columns[temp_columns != "submitted for testing"]
mosquito_modified <- na.omit(mosquito_modified, cols = temp_columns)

# Add "week_num" column
# This code ensures that the ISO 8601 standard is applied to the week number calculation
# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
# Weeks start on Monday and end on Sunday.
# The last week of the year, week 52 or 53, is the one that contains December 28th.
# we will use week_num to join tables
# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08
mosquito_modified <- mosquito_modified %>%
  mutate(
    week_num = paste0(year(`collection date`), "-", sprintf("%02d", isoweek(`collection date`)))
  )
# Check the structure of the final cleaned dataframe
str(mosquito_modified)

# before filtering the dates to have a cleaned dataset with all the records.
write_csv(mosquito_modified, here("data", "processed-data", "mosquito_clean_all.csv"))
saveRDS(mosquito_modified, here("data", "processed-data", "rds", "mosquito_clean_all.rds"))
```

## Reformat the collection data

### The Massachusetts Department of Public Health and US CDC require that all mosquito control districts submit their colelction data in a specific format. This format divides each trap event into separate rows based on mosquito species. However, this format is not useful for certain calculations where a zero count for a species in not explicitly recorded. Even if you sort by species and calculate average count per trap, the average could be inflated since the records will not include trap events with a zero count of that species.

### In order to correct this, I will pivot the data and create a column for every unique "species code" and sum the "pool size" for each species code. The resulting table will have one row for every trap event -a trap event is when "town", "date of collection", "trap type" are all the same.

### The resulting data frame will rectangle in shape and contain 1 row per trap event with all possible species as a separate column.

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)

# Step 1: Summarize the data
mosquito_summarized <- mosquito_modified %>%
  group_by(`collection date`, town, `trap type`, `species code`, mcd, `submitted for testing`) %>%
  summarise(pool_size_sum = sum(`pool size`, na.rm = TRUE), .groups = 'drop')

# Step 2: Pivot to wide format
wide_data <- mosquito_summarized %>%
  pivot_wider(
    names_from = `species code`,
    values_from = pool_size_sum,
    id_cols = c(`collection date`, town, `trap type`, mcd)
  )

# Define species_columns after pivoting
species_columns <- setdiff(names(wide_data), c("collection date", "town", "trap type", "mcd"))

# Ensure there's no duplication in the conversion attempt
# Step 3: Convert list columns to numeric, replacing NULL or list elements without values with 0
# Utilizing sapply for a direct approach to handle potential list-columns
wide_data <- wide_data %>%
  mutate(across(.cols = all_of(species_columns), 
                .fns = ~ map_dbl(.x, function(item) {
                  # Handle NULL items directly
                  if (is.null(item)) {
                    return(0)
                  # Handle vectors: convert NA to 0, and take the first item if it's a vector
                  } else if (length(item) > 1 || is.na(item)) {
                    item[is.na(item)] <- 0
                    return(as.numeric(item[1]))
                  } else {
                    return(as.numeric(item))
                  }
                })))
# Step 4: Add week_num column
wide_data <- wide_data %>%
  mutate(
    `collection date` = as.Date(`collection date`),
    week_num = paste0(year(`collection date`), "-", sprintf("%02d", isoweek(`collection date`)))
  )

# Output the transformed data to inspect its structure
head(wide_data)

# Save the new file
# Specify the file path using here()
file_path_wide <- here("data", "processed-data", "mosquito_wide_all.csv")

# Save the wide-format data to the specified path
write_csv(wide_data, file_path_wide)

# Define the path for the wide-format dataset RDS
file_path_wide_rds <- here("data", "processed-data", "rds", "mosquito_wide_all.rds")

# Save the wide-format data as an RDS file
saveRDS(wide_data, file_path_wide_rds)

# Now we will filter the the mosquito data to match the time period of the virus data. 
# Since mosquito populations can be affected by the previous year's populations, we will filter for 2011-

# Filter wide_data for the specified date range
filtered_mosquito_data <- wide_data %>%
  filter(`collection date` >= as.Date("2007-01-01") & `collection date` <= as.Date("2023-12-31"))

# Define the path for the CSV file
file_path_filtered_csv <- here("data", "processed-data", "mosquito_wide_filtered.csv")

# Save the filtered wide-format data to the specified CSV path
write_csv(filtered_mosquito_data, file_path_filtered_csv)

# Define the path for the RDS file
file_path_filtered_rds <- here("data", "processed-data", "rds", "mosquito_wide_filtered.rds")

# Save the filtered wide-format data as an RDS file
saveRDS(filtered_mosquito_data, file_path_filtered_rds)

```

# Exploring the Data

##Study location Below is a map of the mosquito trap sites. This is an interactive map and will only work on web or other html-friendly platforms.

```{r}
# Create the map
map_study <- leaflet(trap_location) %>%
  addTiles() %>%  # This adds the default OpenStreetMap tiles
  addMarkers(lng = ~DecLong, lat = ~DecLat, popup = ~Location)  # Customize the popup content as needed

# Display the map
map_study

# Specify the path and name of your HTML file
file_path_map <- here("results", "figures", "map_study.html")

# Save the map
saveWidget(map_study, file_path_map, selfcontained = TRUE)
```

## EEE in mosquitoes and mammals

### To give an idea of when EEE virus spilled over from mosquitoes and birds and into mammal populations, lets take a look at the data. Let's see when the first EEE isolations were found in mosquitoes each year. Let's also look at the years that human and veterinary cases occured.

### From the table, we see that human or othe mammal cases were recorded in 2014, 2018, 2019, and 2020. 2019 had the highest case counts for both humans and other mammals. **Because of this, we will pay close attention to 2019 through out the exploration.**

```{r}

# virus_data, cases_human, and cases_vet data frames 
# Prepare virus_data: Find the first positive EEE test date by year
first_positive_eee_test_by_year <- combined_virus_data %>%
  filter(Virus == "EEE") %>%
  mutate(Year = year(as.Date(`Collection Date`))) %>%
  group_by(Year) %>%
  summarise(First_Positive_Test_Date = min(`Collection Date`)) %>%
  ungroup() %>%
  arrange(Year)

# Prepare cases_human: Summarize human EEE cases by year
human_cases_by_year <- cases_human %>%
  filter(`Virus Result` == "EEE") %>%
  mutate(Year = year(`Onset Date`)) %>%
  group_by(Year) %>%
  summarise(Human_Cases = n()) %>%
  ungroup() %>%
  arrange(Year)

# Prepare cases_vet: Summarize animal EEE cases by year
animal_cases_by_year <- cases_vet %>%
  filter(virus == "eee") %>%
  group_by(year) %>%
  summarise(Animal_Cases = n()) %>%
  ungroup() %>%
  arrange(year)

# Ensure 'Year' column is numeric across all data frames for compatibility
# Note: This step might already be covered by the mutations above but included here for clarity
first_positive_eee_test_by_year$Year <- as.numeric(first_positive_eee_test_by_year$Year)
animal_cases_by_year$year <- as.numeric(animal_cases_by_year$year) # Make sure year is numeric and named consistently

# Merge the summaries into one combined data frame
combined_eee_data <- first_positive_eee_test_by_year %>%
  left_join(human_cases_by_year, by = "Year") %>%
  left_join(animal_cases_by_year, by = c("Year" = "year")) # Ensure correct column names are used for joining

# Replace NA values with 0 for Human_Cases and Animal_Cases
combined_eee_data$Human_Cases[is.na(combined_eee_data$Human_Cases)] <- 0
combined_eee_data$Animal_Cases[is.na(combined_eee_data$Animal_Cases)] <- 0

# Creating the table with centered headings and data
kable_table_isolation_case <- kable(combined_eee_data, "html", 
                     col.names = c("Year", "First Positive EEE Test Date", "Human EEE Cases", "Animal EEE Cases"),
                     caption = "Annual Summary of EEE Virus Activity",
                     align = c('c','c','c','c')) %>% # This aligns all columns' data to center
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = F, position = "center") %>%
  add_header_above(c(" " = 1, "EEE Virus Detection and Cases" = 3)) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, width = "8em") %>% # Adjusted width without altering alignment
  column_spec(3, width = "5em", bold = TRUE, color = "red") %>%
  column_spec(4, width = "5em", bold = TRUE, color = "blue")

# Displaying the table
kable_table_isolation_case

# Define the path where you want 
#output_path_quarto <- here("results", "tables", "eee_isolation_cases")
output_path_png <- here("results", "tables", "eee_isolation_cases.png")

#save_kable(kable_table_isolation_case, file = output_path_quarto)
save_kable(kable_table_isolation_case, file = output_path_png, zoom = 4)

# Define the paths for the RDS files
first_positive_eee_test_by_year_path <- here("data", "processed-data", "rds", "first_positive_eee_test_by_year.rds")
human_cases_by_year_path <- here("data", "processed-data", "rds", "human_cases_by_year.rds")
animal_cases_by_year_path <- here("data", "processed-data", "rds", "animal_cases_by_year.rds")
combined_eee_data_path <- here("data", "processed-data", "rds", "combined_eee_data.rds")

# Save the data frames as RDS files
saveRDS(first_positive_eee_test_by_year, first_positive_eee_test_by_year_path)
saveRDS(human_cases_by_year, human_cases_by_year_path)
saveRDS(animal_cases_by_year, animal_cases_by_year_path)
saveRDS(combined_eee_data, combined_eee_data_path)
```

## Which county has the most EEE positive mosquitoes of those tested?

```{r}
eee_positive_by_county <- combined_virus_data %>%
  filter(Virus == "EEE") %>%  # Filter for rows where Virus is EEE
  group_by(County) %>%
  summarise(EEE_Positive_Tests = n()) %>%  # Count the number of EEE positive tests in each group
  arrange(desc(EEE_Positive_Tests))  # Order the results by EEE_Positive_Tests in descending order

# Create the table with kable and style it with kableExtra
eee_positive_table <- kable(eee_positive_by_county, "html", 
                            col.names = c("County", "EEE Positive Tests"),
                            caption = "EEE Positive Tests by County",
                            align = c('c','c')) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = FALSE, position = "center") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, bold = TRUE, color = "red")

# Display the table
eee_positive_table

# Define the path where you want to save the HTML file
#output_path_qmd <- here("results", "tables", "eee_isolation_county.html")
output_path_county <- here("results", "tables", "eee_isolation_county.png")

# Save the kableExtra table as an HTML file
#save_kable(eee_positive_table, file = output_path_qmd)
save_kable(eee_positive_table, file = output_path_county, zoom = 6)
```

## *Culiseta melanura* (MEL) is the main enzootic vector of EEE. Some entomologist believe that high levels of MEL traps indicate higher risk for EEE transmission. Using the new mosquito collection table created from the pivot, let's calculate the proportion of MEL captured compared to all other mosquito vector species and create a plot to visualize it.

```{r}

# Define path to the data file
trap_wide_path <- here("data", "processed-data", "mosquito_wide.csv")

# Read the data
trap_proportion <- read_csv(trap_wide_path) %>%
  mutate(Year = year(as.Date(`collection date`, format = "%Y-%m-%d")))

# Calculate the proportion of MEL out of the total for all species
proportion_mel_by_year <- trap_proportion %>%
  group_by(Year) %>%
  summarise(Total_MEL = sum(MEL, na.rm = TRUE),
            Total_All = sum(MEL + PER + PIP + RES + SAL + TRI + SAP, na.rm = TRUE),
            MEL_Prop = Total_MEL / Total_All)

# Calculate the average number of MEL by year
average_mel_by_year <- trap_proportion %>%
  group_by(Year) %>%
  summarise(Average_MEL = mean(MEL, na.rm = TRUE))

# Save the processed data as RDS and CSV
saveRDS(proportion_mel_by_year, here("data", "processed-data", "rds", "yearly_mel_proportion.rds"))
write.csv(proportion_mel_by_year, here("data", "processed-data", "yearly_mel_proportion.csv"), row.names = FALSE)

saveRDS(average_mel_by_year, here("data", "processed-data", "rds", "yearly_mel_average.rds"))
write.csv(average_mel_by_year, here("data", "processed-data", "yearly_mel_average.csv"), row.names = FALSE)

# Plot for the proportion of MEL by year

mel_prop = ggplot(proportion_mel_by_year, aes(x = factor(Year), y = MEL_Prop)) +
  geom_col(fill = "steelblue") +
  labs(title = "Proportion of Culiseta melanura (MEL) by Year",
       x = "Year",
       y = "Proportion of MEL") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(here("results", "figures", "mel_proportion_by_year.png"), width = 8, height = 6, dpi = 300)
mel_prop

# Plot for the average count of MEL by year
mel_ave = ggplot(average_mel_by_year, aes(x = factor(Year), y = Average_MEL, fill = "Year")) +
  geom_col() +
  labs(title = "Average Count of Culiseta melanura (MEL) by Year",
       x = "Year",
       y = "Average Count of MEL") +
  scale_fill_viridis_d() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

mel_ave

ggsave(here("results", "figures", "average_mel_by_year.png"), plot = mel_ave, width = 8, height = 6, dpi = 300)

```
## Average MEL by week
### IN addition to the average MEL by year, we will also look at the average MEL by week. This will give us a more granular view of the MEL population dynamics throughout the mosquito season.
#### This will be used in the analysis phase
```{r}
# Calculate the average number of MEL by week for all trap types
# Define path to the data file
trap_wide_path_all <- here("data", "processed-data", "mosquito_wide_all.csv")

# Read the data
trap_avg_week <- read_csv(trap_wide_path_all)

# Calculate the average number of MEL by week for all trap types
average_mel_week_all <- trap_avg_week %>%
  group_by(week_num, `trap type`) %>%
  summarise(Average_MEL = mean(MEL, na.rm = TRUE), .groups = 'drop')

# Examine the resulting data frame for all trap types
str(average_mel_week_all)

# Save the csv file with averages for all trap types
write_csv(average_mel_week_all, here("data", "processed-data", "average_mel_week_all.csv"))

# Now, filter data for specific CO2 baited light traps
trap_types_to_include <- c("500cc CO2 - CDC Miniature Light Trap", "Dry Ice CO2 - CDC Miniature Light Trap with CO2")
filtered_data <- trap_avg_week %>%
  filter(`trap type` %in% trap_types_to_include)

# Calculate the average number of MEL by week for the filtered CO2 baited traps
average_mel_week_co2 <- filtered_data %>%
  group_by(week_num, `trap type`) %>%
  summarise(Average_MEL = mean(MEL, na.rm = TRUE), .groups = 'drop')

# Examine the resulting data frame for CO2 baited traps
str(average_mel_week_co2)

# Save the csv file with averages for the CO2 baited traps
write_csv(average_mel_week_co2, here("data", "processed-data", "average_mel_week_co2.csv"))
```

## Mosquito abundance is affected by weather. Mosquitoes depend on standing water to complete their larval development. Let's take a look at the annual precipitation using the wx data frame.

```{r annual precip}
# Convert DATE from character to Date type using as.Date and specifying the format
wx$DATE <- as.Date(wx$DATE, format="%m/%d/%y")

# The as.Date function treats years in the format "yy" as follows:
# Years 00-68 are treated as 2000-2068, and years 69-99 are treated as 1969-1999.
# Adjust this according to your data's actual year range if needed.

# Extract the year from each DATE
wx$Year <- format(wx$DATE, "%Y") %>% as.numeric()

# Now that we have the years as numeric values, let's filter for 2007 to 2020
filtered_wx <- wx %>%
  filter(Year >= 2007 & Year <= 2020)

# Save the filtered weather data as an RDS file
filtered_wx_path_rds <- here("data", "processed-data", "rds", "filtered_wx_2007_2020.rds")
saveRDS(filtered_wx, filtered_wx_path_rds)

# Sum PRCP by Year for the filtered data
summed_prcp_by_year_filtered <- filtered_wx %>%
  group_by(Year) %>%
  summarise(Total_PRCP = sum(PRCP, na.rm = TRUE))

# Save the summarized precipitation data as an RDS file
summed_prcp_by_year_path_rds <- here("data", "processed-data", "rds", "summed_prcp_by_year_filtered_2013_2020.rds")
saveRDS(summed_prcp_by_year_filtered, summed_prcp_by_year_path_rds)

# View the result
print(summed_prcp_by_year_filtered)

# Create the ggplot object and assign it to a variable
precip_plot <- ggplot(summed_prcp_by_year_filtered, aes(x = as.factor(Year), y = Total_PRCP)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Total Precipitation by Year (2013-2020)",
       x = "Year",
       y = "Total Precipitation (inches)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

precip_plot

# Specify the file path where you want to save the plot
file_path_precip <- here("results", "figures", "annual_precip.png")

# Use ggsave to save the plot, make sure to use the plot object
ggsave(file_path_precip, plot = precip_plot, width = 8, height = 6, dpi = 300)
```

### Now I want to plot the ave mel by year with the total precip by year

```{r}
# Load the datasets
yearly_mel_ave <- readRDS(here("data", "processed-data", "rds", "yearly_mel_ave.rds"))
summed_prcp_by_year_filtered <- readRDS(here("data", "processed-data", "rds", "summed_prcp_by_year_filtered_2013_2020.rds"))

# Combine the datasets
combined_data <- left_join(yearly_mel_ave, summed_prcp_by_year_filtered, by = "Year")

# Define the path for the RDS file with a descriptive name
combined_data_path_rds <- here("data", "processed-data", "rds", "combined_mel_avg_and_total_precip_2013_2020.rds")

# Save the combined data as an RDS file
saveRDS(combined_data, combined_data_path_rds)

plot_mel_precip = ggplot(data = combined_data) +
  geom_col(aes(x = factor(Year), y = Average_MEL), fill = "blue") +
  geom_line(aes(x = factor(Year), y = Total_PRCP, group = 1), color = "red", linetype = "dashed") +
  geom_point(aes(x = factor(Year), y = Total_PRCP), color = "red", size = 3) +
  geom_hline(yintercept = 45.6, linetype = "dotted", color = "darkgreen", size = 1) +
  annotate("text", x = Inf, y = 45.6, label = "Avg Precip: 45.6 inches", 
           hjust = 3.2, vjust = -1.1, color = "darkgreen", size = 3.5, fontface = "bold", angle = 0) +
  scale_y_continuous(name = "Average Count of MEL", 
                     sec.axis = sec_axis(~ ., name = "Total Precipitation (inches)")) +
  labs(title = "Average Count of MEL Mosquitoes and Total Precipitation by Year",
       x = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.title.x = element_text(face = "bold"))
plot_mel_precip

# Save the plot
ggsave(filename = here("results", "figures", "combined_trends_mel_precip.png"), plot = last_plot(), width = 10, height = 6, dpi = 300)
```

### This will load sunset and sunrise times.
#### the amount of sunlight affects mosquitoes activities and provides cues for them to enter diapause.
##### The suncalc package will be used to calculate the sunrise and sunset times for Bristol County, MA, which is the study location. The times will be calculated for each day from 2007 to 2023. The data will be saved as a CSV file and an RDS file for future use.
```{r}
# Ensure the suncalc library is loaded
library(suncalc)
get_sun_times <- function(date, latitude, longitude) {
  # Get the sunrise and sunset times from the suncalc package
  times <- getSunlightTimes(date = date, lat = latitude, lon = longitude, tz = "UTC")

  # Convert sunrise and sunset times to POSIXct directly without checking class
  # Assume times are returned in a standard format that can be converted directly
  sunrise <- as.POSIXct(times$sunrise, format = "%Y-%m-%dT%H:%M:%S", tz = "UTC")
  sunset <- as.POSIXct(times$sunset, format = "%Y-%m-%dT%H:%M:%S", tz = "UTC")

  # Format sunrise and sunset times within the data frame creation step
  # Create a one-row data frame for each date
  data.frame(
    Date = as.character(date),
    Sunrise = format(sunrise, "%H:%M:%S", tz="UTC"),
    Sunset = format(sunset, "%H:%M:%S", tz="UTC")
  )
}

# Define the geographical coordinates for Bristol County, MA (latitude and longitude)
latitude <- 41.8744
longitude <- -71.0166

# Generate a sequence of dates from 2007-01-01 to 2023-12-31
dates <- seq(as.Date("2007-01-01"), as.Date("2023-12-31"), by = "day")

# Calculate sunrise and sunset times for each date
sun_times <- lapply(dates, function(date) {
  get_sun_times(date, latitude, longitude)
})

# Convert the list to a data frame
# this is a very large amount of data so it takes time to load. 
sun_times_df <- do.call(rbind, sun_times)
colnames(sun_times_df) <- c("Date", "Sunrise", "Sunset")

# Write the data frame to a CSV file
write.csv(sun_times_df, file = "sunrise_sunset_times_bristol_county_ma.csv", row.names = FALSE)

# Define the file path where you want to save the sunrise and sunset times data frame
sunrise_sunset_times_path <- here("data", "processed-data", "rds", "sunrise_sunset_times_bristol_county_ma.rds")

# Save the sun_times_df data frame as an RDS file
saveRDS(sun_times_df, sunrise_sunset_times_path)

# Output to confirm file saving
cat("Sunrise and sunset times saved to:", sunrise_sunset_times_path, "\n")

#Average daylight by week_num
sun_times_df <- sun_times_df %>%
  mutate(
    # Combine Date with Sunrise and Sunset times to create full POSIXct datetime objects
    Sunrise_dt = as.POSIXct(paste(Date, Sunrise), format="%Y-%m-%d %H:%M:%S", tz="UTC"),
    Sunset_dt = as.POSIXct(paste(Date, Sunset), format="%Y-%m-%d %H:%M:%S", tz="UTC"),
    # Now calculate the difference in hours between Sunset and Sunrise
    daylight_hours = as.numeric(difftime(Sunset_dt, Sunrise_dt, units = "hours"))
  )

# Aggregate data by week and calculate average daylight hours
average_daylight_by_week <- sun_times_df %>%
  group_by(week_num) %>%
  summarise(average_daylight_hours = mean(daylight_hours))

# Print first few rows of the resulting data frame
head(average_daylight_by_week)

# Define the path for the CSV file
avg_day_csv <- here("data", "processed-data", "average_daylight_by_week.csv")

# Write the data frame to a CSV file
write.csv(average_daylight_by_week, file = avg_day_csv, row.names = FALSE)

# Define the file path where you want to save the sunrise and sunset times data frame
average_daylight_by_week_path <- here("data", "processed-data", "rds", "average_daylight_by_week.rds")

# Save the sun_times_df data frame as an RDS file
saveRDS(sun_times_df, average_daylight_by_week_path)
```

### Calcualte the average moon phase by week_num.
#### some resaecrh suggests that moon phase can affect mosquito activity.
##### The moon phase is calculated using the getMoonIllumination function from the lunar package. The function returns the percentage of the moon's illuminated surface area and the moon's phase angle. The phase angle is used to determine the moon phase, which can be one of the following: "New Moon", "First Quarter", "Full Moon", or "Last Quarter". The phase angle is converted to a value between 0 and 1, where 0 corresponds to a New Moon, 0.5 corresponds to a Full Moon, and 1 corresponds to a New Moon. The average moon phase for each week is calculated by taking the mean of the moon phase values for each day in the week.

```{r}
# Generate a sequence of dates from 2007-01-01 to 2023-12-31
dates <- seq(as.Date("2007-01-01"), as.Date("2023-12-31"), by = "day")

# Function to get moon phase for a given date
get_moon_phase <- function(date) {
  illumination <- getMoonIllumination(date)
  phase <- illumination$phase
  return(phase)
}

# Calculate moon phase for each date
moon_phases <- sapply(dates, get_moon_phase)

# Create a data frame with dates and their corresponding moon phases
moon_phases_df <- data.frame(Date = dates, MoonPhase = moon_phases)

# View the first few rows of the data frame
str(moon_phases_df)

# Add week_num column to moon_phases_df
moon_phases_df <- moon_phases_df %>%
  mutate(
    week_num = paste0(year(Date), "-", sprintf("%02d", isoweek(Date)))
  )

# Calculate average moon phase by week_num
average_moon_phase_by_week <- moon_phases_df %>%
  group_by(week_num) %>%
  summarise(AverageMoonPhase = mean(MoonPhase, na.rm = TRUE))

# View the results
head(average_moon_phase_by_week)

```

### 2024-04-05 products/manuscript/Manuscript.qmd
replacing corrupt file

---
title: "Factors Influencing Abundance & Virus Isolation in Mosquitoes"
subtitle: "Which environmental, ecological, and temporal factors impact mosquito populations"
author: Andrew Ruiz
date: "`r Sys.Date()`"
md_number: true
format:
  docx:
    toc: false
    highlight-style: github
    reference_docx: path/to/your/reference/docx/file.docx
editor: 
  markdown: 
    wrap: 72
bibliography: 
  - ../../assets/project_mada.bib
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(readxl)
library(dplyr)
library(kableExtra)
library(dplyr)
library(readr)

here::set_here()

```

::: {#summary .unnumbered}
# Summary/Abstract

*This project will inform the development of my dissertation prospectus
by identifying significant relationships between variables and guide the
methods development.* Eastern equine encephalitis (EEE) is caused by an
Alphavirus transmitted to humans by the bite of an infected mosquito.
Human infections are rare, but serious. About 30-50% of symptomatic
cases lead to death and those who survive are left with life-long
disability. The transmission cycle for EEE includes two distinct cycles:
an amplifying enzootic cycle in which the virus is transmitted between
the *Culiseta melanura* mosquito and birds and an epizootic cycle in
which bridge vector mosquitoes transmit the EEE virus (EEEv) from birds
to humans and other mammals [@goZoonoticEncephalitidesCaused2014].
Because C. melanura feeds almost exclusively on birds, it is not
considered a direct human threat. However, EEEv isolations in *C.
melanura* are an early warning sign that EEEv is circulating in the
ecosystem. Isolations in bridge vector mosquito species indicate
heightened transmission risk to humans.
:::

# Problem statement

Which environmental,ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts and which
models best describe the relationship between these factors?

**This project will lay the foundation for dissertation research that
will begin in June, 2024.**

# Introduction

## Background

Eastern equine encephalitis (EEE) is caused by an Alphavirus transmitted
to humans by the bite of an infected mosquito. Human infections are
rare, but serious [@lindseyEasternEquineEncephalitis2018]. Humans and
other mammals are dead-end hosts for EEE. About 94% of human EEE
infections are asymptomatic. While symptomatic cases are rare, the
consequences can be severe. Mortality rates associated with symptomatic
infections range from 30 to 50% [@morensEasternEquineEncephalitis2019].
Those who survive often experience lifelong disability that includes
debilitating neurological damage
[@massachusettsdepartmentofpublichealthArbovirusSurveillancePlan2024].
On average, there are 11 cases reported in the US annually. Most occur
in Massachusetts and Florida [@vanderkelenRiskIndexModel2014]. Until
recently, nearly all human EEE cases in Massachusetts occurred in two
counties: Bristol and Plymouth.

Since 2000, there have been 45 human cases recorded in Massachusetts,
resulting in 22 deaths
[@massachusettsdepartmentofpublichealthArbovirusSurveillancePlan2024].
There are no human vaccines for EEE, and treatment consists of
palliative care only. Prevention measures include personal behaviors to
avoid mosquito bites and decreasing the mosquito population through
pesticide use and environmental modifications like removing standing
water
[@centersfordiseasecontrolandpreventionPreventionEasternEquine2022].

The Bristol County Mosquito Control Project (BCMCP) coordinates mosquito
surveillance and testing in the county from June to October. BCMCP has
used the same sentinel collection sites and trapping methods for over 40
years. Once trapped, mosquitoes are sorted and counted by species, and
vector species are submitted to the Massachusetts Department of Health’s
(MDPH) State Lab for PCR virus testing. Results are available within 24
hours of submission. When rates of EEE mosquito infections are above a
defined threshold, MDPH notifies the local boards of health and
recommends preventive measures. Recommended measures include outreach
and education to increase personal protective practices . Additionally,
MDPH may recommend that cities and towns in the affected areas cancel
evening outdoor events and discourage outdoor activities when mosquito
vectors are most active (dawn and dusk). When infection rates are high
enough to indicate imminent human transmission, the Commonwealth of
Massachusetts will recommend and fund aerial pesticide applications over
the affected areas.

## Study area

Bristol County, Massachusetts covers an area of 691 square miles with a
population density of 962 people per square mile. It borders Rhode
Island to the west, Plymouth County, MA to the east, and Norfolk County,
MA to the north. Naushon Island and Martha's Vineyard are across
Buzzard's Bay to the Southeast.

Atlantic White Cedar (AWC) swamps span over 4,300 acres within
Massachusetts, primarily concentrated in Plymouth and Bristol counties.
These swamps serve as critical breeding habitats for *Culiseta melanura*
(MEL)
[@commonwealthofmassachusettsdivisionoffisheries-wildlifeNaturalCommunityFact2007],
which is recognized as the primary enzootic vector for Eastern Equine
Encephalitis (EEE). AWC trees have large root systems that form "crypts"
and hold water throughout the year. These crypts serve as a main habitat
for MEL larval growth and provide shelter during winter months.

Between 2003 and 2022 Plymouth and Bristol counties were the epicenters
for EEE transmission in Massachusetts, accounting for 24 out of the
state's 41 human EEE cases.The Bristol County Mosquito Control Project
(BCMCP) conducts mosquito surveillance and abatement activities for the
county and often coordinates activities with Plymouth County Mosquito
Control Project.

## Questions/Hypotheses to be addressed

*This project will lay the foundation for my dissertation research.*
**The goal for this phase of the project is to answer which
environmental, ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts.** From prior
research we know that the mosquito life cyle is influenced by a number
of factors. Temperature plays a major role in the mosquito life cyle
virus transmission dynamics. As temperature increases, the rate of
larval development also increases and the time between when a mosquito
ingests virus in a bloodmeal and when it can transmit it decreases. The
result is that the mosquito population can grow at faster rate and they
will have more time to transmit the virus. However, MEL larvae develop
in well insulated crypts that are less sensitive to ambient air
temperature changes, so the effect of temperature changes on MEL
development speed may be less pronounced than other vector species that
develop in open, standing water, like ditches and containers.
Precipitation increases the amount of standing water and increases the
chances of larval survival during the mosquito season and also during
the winter months for those species that overwinter in the larval stage.
For EEE, the main enzootic vector, MEL, overwinters as larvae.

While these relationships are well known, they are not well described
for EEE, especially in terms of quantifying the role these two factors
play in EEE transmission. Specifically, there's a gap in our
understanding of how these factors influence the number of mosquitoes
caught in traps and the rate of virus infection among those mosquitoes.
This gap is more pronounced when distinguishing the impacts on enzootic
(mosquitoes that primarily feed on birds) versus bridge vector
(mosquitoes that can transmit the virus from birds to humans) mosquito
species. Even less well described is the effect the factors have on
enzootic vs. bridge vector mosquito species.

Moreover, the influence of other variables remains even less explored.
Birds, especially passerine species, serve as amplifying hosts for EEEV
and WNV, potentially affecting the level of virus present in an area.
Despite their importance in the virus transmission cycle, the effect of
fluctuations in bird populations on the prevalence of these viruses
throughout a season has received minimal attention. Finally, given the
migratory nature of many passerine species, it's worth exploring whether
EEEV activity in the warmer regions of the United States, during the
months preceding mosquito season in Massachusetts, could serve as an
early indicator of virus activity in the state. Understanding these
dynamics is crucial for predicting outbreaks and implementing effective
control measures.

While the following questions will not be addressed in this project, it
will hopefully inform research that will answer the following questions
in dissertation research beginning this summer:

1\. Can an AI-driven model that leverages historic mosquito
surveillance, ecological, and environmental data accurately quantify the
risk of human EEEv infection in southeastern Massachusetts?

2\. What is the potential of machine learning algorithms to identify
early warning signals for EEE outbreaks, enabling timely public health
interventions in Massachusetts to prevent human infections?

3\. How can machine learning models use real-time data on mosquito
activity, environmental factors, and weather to accurately predict when
the risk of EEE transmission to humans is low enough to lift evening
outdoor activity restrictions? \*

## Data collection and processing

### Mosquito surveillance data

The mosquito surveillance dataset encompasses counts by species from all
mosquito traps deployed in Bristol County, MA, from 2007 to 2023. The
raw dataset includes over 45,000 records and includes temporal and
spatial variable. The BCMCP, perform mosquito surveillance from June
through October annually.

During this interval, CDC light traps baited with carbon dioxide are
placed at various sentinel sites each week. Designed to attract
mosquitoes in search of blood meals, light traps serve as a reliable
measure of the actively feeding mosquito population, thereby
establishing them as the standard method for EEE surveillance. Some of
the sentinel site have been used for over 50 years.

Additionally, the county employs gravid traps, which are specifically
aimed at attracting and capturing gravid *Culex* mosquitoes looking for
egg-laying sites. While these traps are primarily used for West Nile
Virus (WNV) surveillance, they hold significant value for EEE monitoring
as well. Given the role of *Culex* species as bridge vectors for EEE,
their capture in gravid traps provides critical insight into the risk of
virus transmission.

The traps are installed in the afternoon and retrieved the following
morning. After the traps are collected, the next are stored in a freezer
to preserve and kill the mosquitoes. The mosquitoes are sorted by
species using a dissecting scope. Counts for each species are recorded
in the database and a portion of the vector species are driven to the
Massachusetts State Laboratory in Boston. The lab tests each mosquito
pool (or batch) for WNV and EEEv using PCR. Results are shared with
BCMCP on the same day of testing. Virus isolation results are recorded
in the collection records.

In addition to mosquito counts by species, the dataset includes the town
or city in which they were collected, the collection date, and whether
the mosquitoes were submitted for virus testing.

### Minimun infection rate (MIR)

One indicator of the level of virus activity present in the mosquito
population can be calculated from the information contained in dataset.
The minimum infection rate (MIR) can be calculated at various temporal
and spatial scales from trap site over one day to all sites over an
entire season. MIR is calculated as: $$
MIR = \left( \frac{\text{Number of positive mosquito pools}}{\text{Total number of mosquitoes tested}} \right) \times 1000
$$ MIR provides an estimate of the prevalence of a specific pathogen
within a mosquito population. While it does not directly measure the
risk to humans, a higher MIR in mosquitoes that commonly bite humans can
indicate a higher risk of transmission. A study pubulihsed in 2007
indicated that in years with human cases in Massachusetts between 1979
and 2004, MIR was ≥1.0 by week 31 [@hachiyaHumanEasternEquine2007]. MIR
allows for the comparison of infection rates across different
geographical areas or time periods. This is crucial for identifying
areas or times of higher risk and for evaluating the effectiveness of
control measures.

![Minimum infection rate by
week](images/MIR_by_week_heatmap.png){fig-align="center"}

The surveillance dataset is structured to conform with state and federal
reporting requirements. As a result, the format of the data must be
transformed to in order to make it ready for analysis. Once the data
structure is corrected, it can be joined to other relevant datasets
using spatial or temporal attributes.

### Weather data

*weather data is currently being reprocessed because I have decided to
switch from daily summaries to hourly summaries. new analysis will be
added in version 5* Weather data played a significant role in this phase
of the analysis. Hourly weather summaries were downloaded from NOAA's
Climate Data Portal. The dataset covers the years 2007-2023 at the
Taunton Airport in Bristol County.

The hourly summaries were used to calculate weather variables, such as
the daily maximum temperature and precipitation. While daily summaries
provide this information will less processing, the hourly summaries
provide more granular data that can be used to calculate additional
variables, such as the number of hours above a certain temperature
threshold or degree-days. Degree-days is a measure of the amount of heat
accumulated over a period of time. It is calculated by summing the
difference between the average daily temperature and a base temperature.
Degree-days are used to estimate the rate of development of insects and
other organisms that are sensitive to temperature. For mosquitoes,
degree-days can be used to estimate the rate of larval development and
the time between when a mosquito ingests a virus and when it can
transmit it [@guStatisticalEstimationDegree2006].

In addition to the weather summaries, drought data will be included in
the models. Drought data was obtained from the US Drought Monitor and
the Federal Emergency Management Agency (FEMA). The data includes weekly
drought severity ratings for the entire United States. The data is
available at the county level and will be used to calculate the number
of weeks in drought conditions and the severity of the drought. For this
project, drought data for Bristol County, MA, will be used and
incorporated with regional drought indicators.

### Human and animal case data

While the goal of this project is to set a foundation for models that
predict human cases of EEE later this year. Human case data will not be
an outcome in the models for this project. However, the case data will
incorporated into the models. Some have hypothesized that EEE activity
in the southern United States could serve as an early warning signal for
EEE activity in Massachusetts. The rationale is that birds that migrate
from the south to Massachusetts could bring the virus with them. In
active EEE years, more infected birds may arrive in Massachusetts and
spread the virus to local mosquitoes at a higher rate. Florida is a
known hotspot for EEE activity and is the only state that reports more
human cases than Massachusetts. The Florida Department of Health reports
EEE activity in animals and humans annually. In addition to the wild and
domestic animal case reports, many mosquito control programs in Florida
maintain flocks of sentinel chickens that are tested for EEE and WNV.
The sentinel chicken data is used to monitor virus activity in the
mosquito population.

### Bird observation data

The North American Breeding Bird Survey (BBS) is conducted annually in
late spring or early summer. The United State Geological Survey (USGS)
maintains the BBS database. The BBS is a long-term, volunteer-driven
bird monitoring program that collects data on bird populations across
North America. The data is used to track the status and trends of bird
populations over time. The BBS data includes information on the number
of birds observed, the location of the observation, and the date of the
observation. The data is collected by trained volunteers who conduct
roadside surveys along predetermined routes. The data is used to
estimate the population size and distribution of bird species across the
continent. The BBS data will be used to estimate the population size and
distribution of bird species in Massachusetts.

Since birds are the amplifying hosts for EEE, their population size and
distribution can have a significant impact on the prevalence of the
virus in the environment. Additionally, the migratory patterns of birds
can influence the spread of the virus across regions. By incorporating
bird observation data into the models, we can better understand the
relationship between bird populations and EEE activity in Massachusetts.

### Temporal scale

For most models in this project, the outcome, MIR, will be calculated by
week. While the main temporal scale for this project is the week, the
data will be aggregated to other temporal scales. The week is the most
common temporal scale for mosquito surveillance data. The week number
will be calculated according to ISO 8601 standards, and stored in a
YYYY-WW format. For example, week 8 in 2019 will be displayed as
2019-08. The week number will provide a common field by which all
datasets with a temporal attribute can be joined. It will also allow for
grouping records and making calculations by week number. This format
helps to standardize temporal scales and make comparisons between years
easier.

While mosquito surveillance and weather data can be aggregated to the
week, other datasets, such as the Annual BBS cannot appropriately be
aggregated by week. Additionally, mosquito surveillance in Massachusetts
is conducted from June to October and EEE activity in Florida occurs
year-round. For these predictors, the temporal scale will be the year.

## Data aquisition

With the exception of the mosquito collection data, all other datasets
are publicly available for use without restrictions.

## Data import and cleaning

Datasets will be loaded into R in a Quarto document to document and the
processing methods will be documented.

The Massachusetts Department of Public Health requires that all mosquito
control districts submit their collection data in a specific format.
This format divides each trap event into separate rows based on mosquito
species and whether they were submitted for virus testing. However, this
format is not useful for certain calculations where a zero count for a
species is not explicitly recorded. For example, even when sorted by
species to calculate the average count per trap, the average could be
inflated since the records will not include trap events with a zero
count of that species. In order to correct this, the data was pivoted to
create a column for every unique "species code" then the "pool size" for
each species code is summed for each trap event. A trap event is when
"town", "date of collection", "trap type" are all the same. For species
codes not included in the orginal table for a trap event, the field will
be assigned a zero value. The resulting table will have one row for
every trap event.

# Methods

*This section is under development and will be completed after analysis
is complete. The following is a general outline of the methods.*

Symptomatic EEE, and to a lesser extent WNV, human cases are rare. On
average, the United States records only about 9 cases of EEE annually.
Notably, Massachusetts reports more cases than any other state except
Florida, highlighting regional disparities in disease occurrence.
However, the incidence rate for neuroinvasive EEE cases in Bristol
County, Massachusetts, is estimated to be only around 0.06 per 100,000
people
[@centersfordiseasecontrolandpreventionPreventionEasternEquine2022]. The
low incidence of human cases poses significant challenges for
traditional statistical models, making it difficult to accurately
estimate human transmission risk. This difficulty is largely due to the
sparse nature of the data and the variability in transmission patterns,
which complicates efforts to predict outbreaks and necessitates
alternative modeling approaches.

As stated earlier, MIR will be the main outcome of interest for this
project.

### Schematic of workflow

*Schematic will be refined in later steps. The doc will also be
reformatted to move the schematic to an appendix* *The image should
appear upon rendering*

```{mermaid}
flowchart TB
    id1["Questions"] --> id2["Data Collection"]
    id2 --> id3["Data Loading and Cleaning"]
    id3 --> id4["Data Cleaning"]
    id4 --> id5["Data Transformation"]
    id5 --> id6["Data Exploration & Visualization"]
    id6 --> id7["Data Modeling"]
    id7 --> id8["Evaluation"]

    id8 -.-> id1
    id8 --> id9[Interpretation]
    id9 --> id10[Communication]

    classDef mainProcess fill:#fff,stroke:#333,stroke-width:2px;

    class id1,id2,id3,id4,id5,id6,id7,id8,id9,id10 mainProcess
```

## Exploratory/Descriptive analysis

It is important to know where the mosquitoes infected with EEE are
captured. Below is a table showing the virus isolations for EEE in
trapped mosquitoes by county between 2014 and 2020.

![](images/eee_isolation_county-04.png){fig-align="center" width="192"}

One factor to consider is if early virus isolations are associated with
cases in humans and other mammals. The table below shows the date of the
first EEE virus isolation in mosquitoes and the number of human of other
animal cases by year.

![](images/eee_isolation_cases-04.png){fig-align="center" width="251"}

Unfortunately, the time horizon is limited by data availability.
Additionally, EEE human infections are rare, so traditional statistical
methods may be inadequate to suggest any relationships. What is clear is
that 2019 was an exceptional year for EEE.

*Culiseta melanura* (MEL) plays an important role in EEE virus
amplification in the enzootic cycle. Understanding the population
dynamics might provide more insight into what is driving the 2019 spike.

The graph below shows the proportion of MEL among all mosquito vector
species trapped in Bristol County, MA.

![](images/mel_proportion.png){fig-align="center"}

From this we see a large proportion of MEL in the traps for 2019. In
fact, they comprise more than half of all vector species for that year.
Another interesting aspect is the high proportion in 2018. MEL
overwinters as larvae and as long as there is sufficient standing water
to last the winter, MEL hatched in the previous fall would survive to
emerge the next year. This suggests it might be important to look at
conditions in the previous year or season for a better understanding.

Since water plays a significant role in the mosquito life cycle, let's
look at the annual precipitation.

![](images/annual_precip-02.png){fig-align="center"}

The graph indicates that the annual precipitation total for 2018 is
higher than the other years. This suggest that there was enough water to
support large MEL larval populations though out the winter.

Another view plots the average annual MEL catch against total
precipitation.

![](images/combined_trends_mel_precip.png){fig-align="center"}

## Basic statistical analysis

The initial tests have been basic regression models.

The first models explored the relationship between virus isolation in
mosquitoes and weather conditions (average temperature by week number
and total precipitation by week). The outcome was whether there was at
least 1 positive pool in a week. The predictor in the first model was
the average maximum daily temperature the week before the pool week. The
logistic model suggest that there was a relationship between the two.
(The R code for the model can be found in the EDA-Code file:
logistic\_.pos_temp_precip.R) The average temperature in the week before
is positively associated with the likelihood of a positive result, with
an odds ratio of approximately 1.069 for each unit increase in
temperature (Estimate = 0.06695, SE = 0.02033, p \< 0.001), indicating
that higher temperatures significantly increase the probability of a
positive outcome. The model's improvement is highlighted by a decrease
in residual deviance to 405.79 from the null deviance of 417.24, upon
incorporating Lagged_Avg_TMAX as a predictor, demonstrating a better fit
to the observed data. However, with a residual deviance of 405.79 on 299
degrees of freedom, it is evident that not all variability in the
outcome is explained, suggesting the presence of other influential
factors. The model's AIC of 409.79 indicates a balance between model
complexity and predictive power, supporting its relative utility in
understanding the data. However, this relationship should be examined
more closely as temperature is likely auto correlated with progressing
weeks during summer months. While precipitation has been demonstrated to
influence the population dynamics it did not show to have any effect on
a positive result.

Another model assessed whether the temperature precipitation in April,
May, and June of each year effected the date of the first positive pool
test. The date of the first positive pool was identified and assigned a
day number from January 1st of that year. The predictors in this linear
model were the average maximum temperature for April, May, and June and
the sum of precipitation for the same time period. None of these
predictors were significant.

There was one other significant result. When MIR is set as the
dependent, continuous variable in a linear regression and the average
maximum temperature for the previous week is used as the predictor, the model suggests there
is a significant inverse relationship. As average temperature increases,
the MIR decreases (model_mir_temp.R). This relationship is
counterintutive, especially given the result of the logistic regression
and requires more examination.

More models assessing the effects of weather and other factors on
positive pools by mosquito species and MIR values will be developed in
Part 4 of the assignment. Future models will examine the the effects of
weather conditions on mosquito population dynamics.

## Statistical analysis

*UPDATE for Pt 4* *a better source for weather data was found and it
being processed.* *the analysis will be updated in the next version of
the document*

![](images/correlation_plot.png){fig-align="center"}

# Results

*updated in later phases of the project*

## Full analysis

*These will be generated in Pt 5* 
# Discussion 
*this section will be
completed when analysis is done*

## Strengths and Limitations

*this section will be updated after analysis is complete* The accuracy
and reliability of the mosquito collection data is excellent. The same
procedures and sentinel sites have were in place for the entire
analytical horizon. Further, there were only two people responsible for
sorting, identifying and counting species over the entire period.

MIR is not a perfect measure of virus activity in mosquitoes. It tends
to underestimate the level of virus circulating in the environment.
However, other measures require equal pool numbers for testing. There is
a correction factor for variable pool numbers and I will explore this
option in later phases of this project.

The spatial scale is an issue with the human and veterinary cases data.
Human case data is released only on a county scale to ensure patient
privacy. Some veterinary case data is available on the town-level, but
most remains only on the county-level. These limitations are barriers to
accurate spatial analysis. The project will rely on virus isolation in
mosquitoes to model risk. The available bird data relies heavily on
crowd-sourced reports. There is one standardized survey (Breeding Bird
Survey) that seems to be more accurate and reliable. However, this data
is not available for every year.

## Conclusions *this section will be completed when analysis is done*

# References

### 2024-04-05 R/eda-code/Model_mir_temp.R

@@ -1,0 +1,175 @@
+# Load necessary libraries for data manipulation, date handling, and file path management
+library(lubridate)
+library(here)
+library(dplyr)
+library(ggplot2)
+
+
+# Load the datasets:
+# MIR_by_week: Contains information on mosquito pool tests, including the week number and MIR.
+# wx_selected: Weather data with daily maximum temperature (TMAX) and precipitation (PRCP).
+# mosquito_wide: Another dataset related to mosquito data, although it's loaded, it's not used in the subsequent analysis.
+MIR_by_week_co2 <- read_csv(here("data", "processed-data", "MIR_by_week_co2.csv"))
+wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
+mosquito_wide <- read_csv(here("data", "processed-data", "mosquito_wide.csv"))
+drought_fema4 <- read_csv(here("data", "processed-data", "drought_fema4.csv"))
+drought_ma_ri_ct <- read_csv(here("data", "processed-data", "drought_ma_ri_ct.csv"))
+*average_moon_phase_by_week <- read_csv(here("data", "processed-data", "average_moon_phase_by_week.csv"))
+average_daylight_by_week <- read_csv(here("data", "processed-data", "average_daylight_by_week.csv"))
+average_mel_weekco2 <- read_csv(here("data", "processed-data", "average_mel_week_co2.csv"))
+wx_hourly_clean <- read_csv(here("data", "processed-data", "hourly_wx_clean.csv"))
+
+# Aggregate weather data by week:
+# Calculate the average maximum temperature (Avg_TMAX) and total precipitation (Total_PRCP) for each week.
+# This simplification allows for weekly comparisons with the MIR data.
+wx_aggregated <- wx_selected %>%
+  group_by(week_num) %>%
+  summarise(Avg_TMAX = mean(TMAX, na.rm = TRUE),
+            Total_PRCP = sum(PRCP, na.rm = TRUE),
+            .groups = 'drop')
+
+# Adjust for a one-week lag in weather data:
+# Create a new column (Lagged_week_num) that represents each week's preceding week. 
+# This adjustment accounts for the delayed effect weather might have on MIR.
+wx_aggregated <- wx_aggregated %>%
+  mutate(
+    # Split week_num to get year and week as numeric
[…]

### 2024-04-05 R/eda-code/logistic_.pos_temp_precip.R

@@ -1,0 +1,67 @@
+library(here)
+library(dplyr)
+library(lubridate)
+library(readr)
+
+
+# Load the PCR data set which includes information on mosquito pools tested for viruses, 
+# results of those tests, and the date of collection.
+pcr <- read_csv(here("data", "processed-data", "pcr_data_expanded_filtered.csv"))
+
+# Load the weather data set which includes daily maximum temperatures (TMAX) and other weather variables.
+wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
+
+# Aggregate the weather data by week to calculate the average maximum temperature for each week.
+# This simplifies the daily temperature data into a more manageable weekly summary,
+# making it easier to correlate with weekly mosquito test results.
+wx_selected <- wx_selected %>%
+  group_by(week_num) %>%
+  summarise(Avg_TMAX = mean(TMAX, na.rm = TRUE)) %>%
+  ungroup()
+wx_selected <- wx_selected %>%
+  group_by(week_num) %>%
+  summarise(
+    Avg_TMAX = mean(TMAX, na.rm = TRUE),
+    Total_PRCP = sum(PRCP, na.rm = TRUE)
+  ) %>%
+  ungroup()
+
+# Create a lagged version of the average weekly maximum temperature to explore the potential delayed effect of temperature on mosquito test results.
+# The lag function shifts the temperature data by one week, assuming that the impact of temperature on virus transmission might not be immediate.
+weather_weekly <- wx_selected %>%
+  arrange(week_num) %>%
+  mutate(Lagged_Avg_TMAX = lag(Avg_TMAX, 1))
+
+weather_weekly <- wx_selected %>%
+  arrange(week_num) %>%
+  mutate(
+    Lagged_Avg_TMAX = lag(Avg_TMAX, 1),
+    Lagged_Total_PRCP = lag(Total_PRCP, 1)
+  )
+
+weather_weekly <- wx_selected %>%
+  arrange(week_num) %>%
+  mutate(
+    Lagged_Avg_TMAX = lag(Avg_TMAX, 1),
+    Lagged_Total_PRCP = lag(Total_PRCP, 1),
+    Binary_PRCP = if_else(Lagged_Total_PRCP > 0, 1, 0)
+  )
+# Aggregate the PCR data by week to determine if there was at least one positive result in each week.
+# This transformation reduces the detailed test results into a weekly summary,
[…]

### 2024-04-21 R/analysis-code/analysis.qmd
Part 5
Analysis is complete. Readme files and removal of outdated datasets and codes will be removed throughout the week. Additionally, the manuscript will be finalized this weel

---
title: "statistical-analysis"
author: "Andrew Ruiz"
format: html
editor: source
---

# Load the necessary libraries for the analysis

```{r}
library(here)
library(dplyr)
library(lubridate)
library(readr)
library(stargazer)
library(mgcv)
library(tidyr)
library(caret)
library(randomForest)
library(tidymodels)
library(ranger)
library(dials)
library(vip)
library(ggplot2)
library(boot)
library(yardstick)
library(glmnet) # For LASSO
library(dials)
library(mlr3)
library(data.table)
library(Matrix)
library(gbm)
```

## First logistic model

#### Effect of temperature and precipitation the week prior to mosquito collection on the likelihood of observing at least one positive mosquito test result in a week

```{r}
# Load the PCR data set which includes information on mosquito pools tested for viruses, 
# results of those tests, and the date of collection.
pcr <- read_csv(here("data", "processed-data", "pcr_data_expanded_filtered.csv"))

# Load the weather data set which includes daily maximum temperatures (TMAX) and other weather variables.
wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))

# Aggregate the weather data by week to calculate the average maximum temperature for each week.
# This simplifies the daily temperature data into a more manageable weekly summary,
# making it easier to correlate with weekly mosquito test results.
wx_selected <- wx_selected %>%
  group_by(week_num) %>%
  summarise(
    Avg_TMAX = mean(TMAX, na.rm = TRUE),
    Total_PRCP = sum(PRCP, na.rm = TRUE)
  ) %>%
  ungroup()

# View the first few rows to confirm it worked
head(wx_selected)

# Create a lagged version of the average weekly maximum temperature to explore the potential delayed effect of temperature on mosquito test results.
# The lag function shifts the temperature data by one week, assuming that the impact of temperature on virus transmission might not be immediate.
weather_weekly <- wx_selected %>%
  arrange(week_num) %>%
  mutate(Lagged_Avg_TMAX = lag(Avg_TMAX, 1))

weather_weekly <- wx_selected %>%
  arrange(week_num) %>%
  mutate(
    Lagged_Avg_TMAX = lag(Avg_TMAX, 1),
    Lagged_Total_PRCP = lag(Total_PRCP, 1)
  )

weather_weekly <- wx_selected %>%
  arrange(week_num) %>%
  mutate(
    Lagged_Avg_TMAX = lag(Avg_TMAX, 1),
    Lagged_Total_PRCP = lag(Total_PRCP, 1),
    Binary_PRCP = if_else(Lagged_Total_PRCP > 0, 1, 0)
  )
# Aggregate the PCR data by week to determine if there was at least one positive result in each week.
# This transformation reduces the detailed test results into a weekly summary,
# indicating the presence or absence of positive tests each week.
pcr_weekly <- pcr %>%
  group_by(week_num) %>%
  summarise(Positive_Result = max(as.integer(Is_Positive), na.rm = TRUE))

# Merge the aggregated PCR data with the lagged weather data by matching their week numbers.
# This combined data set enables the analysis of the relationship between weekly weather conditions and mosquito test results.
combined_wx_pcr <- merge(pcr_weekly, weather_weekly, by = "week_num")

# Fit a logistic regression model to examine the influence of lagged average weekly maximum temperature on the likelihood of observing at least one positive mosquito test result in a week.
# The binomial family specifies that the dependent variable (Positive_Result) is binary,
# making logistic regression an appropriate modeling choice for this binary outcome.
logistic_model_temp <- glm(Positive_Result ~ Lagged_Avg_TMAX + Binary_PRCP, family = binomial, data = combined_wx_pcr)

# Output the summary of the logistic regression model to review the estimated coefficients, their statistical significance, and the overall model fit.
# This summary provides insights into the relationship between temperature and the probability of positive mosquito test results.
summary(logistic_model_temp)

# Output a LaTeX-formatted table to the console
stargazer(logistic_model_temp, type = "text")

par(mfrow = c(2, 2))
plot(logistic_model_temp)

```

```{r}
pcr <- read_csv(here("data", "processed-data", "pcr_data_expanded_filtered.csv"))
wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))

# Step 2: Calculate the average TMAX for April, May, and June separately
avg_temp_by_month <- wx_selected %>%
  filter(month(DATE) %in% c(4, 5, 6)) %>%
  group_by(Year = year(DATE), Month = month(DATE)) %>%
  summarise(Avg_TMAX = mean(TMAX, na.rm = TRUE), .groups = 'drop')

# Pivot the average temperatures into separate columns for each month
avg_temp_by_month_wide <- avg_temp_by_month %>%
  pivot_wider(names_from = Month, values_from = Avg_TMAX, names_prefix = "Avg_TMAX_M") %>%
  # Ensure all years are present even if some data might be missing
  complete(Year, fill = list(Avg_TMAX_M4 = NA, Avg_TMAX_M5 = NA, Avg_TMAX_M6 = NA))

# Step 4: Identify the first positive test date each year in mosquito_data
first_positive_dates <- pcr %>%
  filter(Result == "Positive") %>%
  group_by(Year = year(`Collection Date`)) %>%
  summarise(First_Positive_Date = min(`Collection Date`), .groups = 'drop')

# Step 5: Merge the datasets on Year
combined_data_firstpos <- left_join(first_positive_dates, avg_temp_by_month_wide, by = "Year")

# Step 6: Convert First_Positive_Date to a numerical value (e.g., day of the year)
combined_data_firstpos <- combined_data_firstpos %>%
  mutate(First_Positive_DOY = yday(First_Positive_Date))

# Explore the relationship via linear regression with Avg_TMAX_M4, Avg_TMAX_M5, Avg_TMAX_M6 as predictors
model2 <- lm(First_Positive_DOY ~ Avg_TMAX_M4 + Avg_TMAX_M5 + Avg_TMAX_M6, data = combined_data_firstpos)
summary(model2)

par(mfrow = c(2, 2))
plot(model2)

```

# Does the total precipitation in April, May, and June influence the date of the first positive test?

```{r}
wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))

# Calculate average TMAX and total PRCP for April, May, and June separately
avg_temp_and_prcp <- wx_selected %>%
  filter(month(DATE) %in% c(4, 5, 6)) %>%
  group_by(Year = year(DATE), Month = month(DATE)) %>%
  summarise(
    Avg_TMAX = mean(TMAX, na.rm = TRUE),
    Total_PRCP = sum(PRCP, na.rm = TRUE),
    .groups = 'drop'
  )

# Continue from the summarise step
wide_data <- avg_temp_and_prcp %>%
  # Convert to a long format to distinguish between Avg_TMAX and Total_PRCP
  pivot_longer(cols = c(Avg_TMAX, Total_PRCP), names_to = "Measure", values_to = "Value") %>%
  # Create a combined name for later pivot_wider
  unite("NewName", Measure, Month, sep = "_M") %>%
  # Pivot wider with a single names_prefix
  pivot_wider(names_from = NewName, values_from = Value) %>%
  complete(Year, fill = list(Avg_TMAX_M4 = NA, Avg_TMAX_M5 = NA, Avg_TMAX_M6 = NA, Total_PRCP_M4 = NA, Total_PRCP_M5 = NA, Total_PRCP_M6 = NA))

# Load and prepare mosquito data
first_positive_dates <- pcr_data_expanded_filtered %>%
  filter(Result == "Positive") %>%
  group_by(Year = year(`Collection Date`)) %>%
  summarise(First_Positive_Date = min(`Collection Date`), .groups = 'drop')

# Merge the datasets on Year
combined_data <- left_join(first_positive_dates, wide_data, by = "Year")

# Convert First_Positive_Date to a numerical value (e.g., day of the year)
combined_data <- combined_data %>%
  mutate(First_Positive_DOY = yday(First_Positive_Date))

# Explore the relationship via linear regression
model <- lm(First_Positive_DOY ~ Total_PRCP_M4 + Total_PRCP_M5 + Total_PRCP_M6, data = combined_data)
summary(model)

```

## Two-stage modeling approach using Generalized Additive Models (GAMs) to understand the factors influencing Mosquito Infection Rate (MIR).

### First stage: zero inflation model

#### This model predicts whether MIR is greater than zero using smoothed functions of several predictors, such as the Drought Severity Composite Index for the area (DSCI_ma), average mosquito larvae counts (Average_MEL), and the average current hourly wind speed (Avg_HourlyWindSpeed_current).

##### *The package required for this is mgcv*

```{r}
# load the dataset with a unique dataframe name
mir_2part <- read_csv(here("data", "processed-data", "joined_data_mir_co2.csv"))

#define the model
zero_gam_model <- gam(I(MIR > 0) ~ s(DSCI_ma) + s(Average_MEL) + s(Avg_HourlyWindSpeed_current),
                      family = binomial(link = "logit"), 
                      data = mir_2part)

summary(zero_gam_model)
plot(zero_gam_model)

# Define the null model with only an intercept
null_gam_model <- gam(I(MIR > 0) ~ 1,
                      family = binomial(link = "logit"), 
                      data = mir_2part)

summary(null_gam_model)

# Calculate the AIC for the null model
aic_null_gam_model <- AIC(null_gam_model)
aic_null_gam_model

aic_zero_gam_model <- AIC(zero_gam_model)
aic_zero_gam_model

# # load the dataset with a unique dataframe name
# mir_2part_all <- read_csv(here("data", "processed-data", "all_mir_clean_joined.csv"))

#define the model
zero_gam_model <- gam(I(MIR > 0) ~ s(DSCI_ma) + s(Average_MEL) + s(Avg_HourlyWindSpeed_current),
                      family = binomial(link = "logit"), 
                      data = mir_2part)

summary(zero_gam_model)
plot(zero_gam_model)

# Define the null model with only an intercept
null_gam_model <- gam(I(MIR > 0) ~ 1,
                      family = binomial(link = "logit"), 
                      data = mir_2part)

summary(null_gam_model)

# Calculate the AIC for the null model
aic_null_gam_model <- AIC(null_gam_model)
aic_null_gam_model

aic_zero_gam_model <- AIC(zero_gam_model)
aic_zero_gam_model
```

```{r}

# Response variable
response_var <- "MIR"

# Specified predictors for modeling
specified_predictors <- c("Avg_TMAX", "Total_PRCP", "Hrs_Above_80_lag", "Hrs_Above_75_lag",
                          "Hrs_Below_50_lag", "Hrs_Between_50_86_lag", "Avg_HourlyDryBulbTemp_lag",
                          "Avg_HourlyWetBulbTemp_lag", "Hrs_Below_50_current", "DSCI_fema",
                          "Avg_HourlyRelativeHumidity_current", "Avg_HourlyWetBulbTemperature_current", 
                          "Avg_HourlyWindSpeed_current", "None.x", "DSCI_fema", "average_daylight_hours", "Average_MEL", "MEL_Prop")

# Initiate lists to store models and summaries
gam_models_zero <- list()
gam_models_positive <- list()
summary_models_zero <- list()
summary_models_positive <- list()

# Loop through each specified predictor
for (predictor in specified_predictors) {
  # Check if predictor exists in the dataset
  if (!predictor %in% names(mir_2part)) {
    cat("Skipping", predictor, "because it is not in the dataset.\n")
    next
  }

  # Zero-inflation part
  formula_zero <- as.formula(paste0("I(", response_var, " > 0) ~ s(", predictor, ")"))
  zero_model <- gam(formula_zero, data = mir_2part, family = binomial(link = "logit"))
  gam_models_zero[[predictor]] <- zero_model
  summary_models_zero[[predictor]] <- summary(zero_model)

  cat("Zero-inflation model for", predictor, "\n")

  # Positive counts part
  if (any(mir_2part[[response_var]] > 0)) {
    positive_data <- mir_2part[mir_2part[[response_var]] > 0, ]
    formula_positive <- as.formula(paste0(response_var, " ~ s(", predictor, ")"))
    positive_model <- gam(formula_positive, data = positive_data, family = gaussian())
    gam_models_positive[[predictor]] <- positive_model
    summary_models_positive[[predictor]] <- summary(positive_model)

    cat("Positive model for", predictor, "\n")
  }

  # print or plot the summaries here...
  print(summary(zero_model))
  print(summary(positive_model))
  plot(zero_model)
  plot(positive_model)
}

### ALL MIR###
# Response variable
response_var <- "MIR"

# Specified predictors for modeling
specified_predictors <- c("Avg_TMAX", "Total_PRCP", "Hrs_Above_80_lag", "Hrs_Above_75_lag",
                          "Hrs_Below_50_lag", "Hrs_Between_50_86_lag", "Avg_HourlyDryBulbTemp_lag",
                          "Avg_HourlyWetBulbTemp_lag", "Hrs_Below_50_current", "DSCI_fema",
                          "Avg_HourlyRelativeHumidity_current", "Avg_HourlyWetBulbTemperature_current", 
                          "Avg_HourlyWindSpeed_current", "None.x", "DSCI_fema", "average_daylight_hours", "Average_MEL", "MEL_Prop")

# Initiate lists to store models and summaries
gam_models_zero <- list()
gam_models_positive <- list()
summary_models_zero <- list()
summary_models_positive <- list()

# Loop through each specified predictor
for (predictor in specified_predictors) {
  # Check if predictor exists in the dataset
  if (!predictor %in% names(mir_2part_all)) {
    cat("Skipping", predictor, "because it is not in the dataset.\n")
    next
  }

  # Zero-inflation part
  formula_zero <- as.formula(paste0("I(", response_var, " > 0) ~ s(", predictor, ")"))
  zero_model <- gam(formula_zero, data = mir_2part_all, family = binomial(link = "logit"))
  gam_models_zero[[predictor]] <- zero_model
  summary_models_zero[[predictor]] <- summary(zero_model)

  cat("Zero-inflation model for", predictor, "\n")

  # Positive counts part
  if (any(mir_2part_all[[response_var]] > 0)) {
    positive_data <- mir_2part_all[mir_2part_all[[response_var]] > 0, ]
    formula_positive <- as.formula(paste0(response_var, " ~ s(", predictor, ")"))
    positive_model <- gam(formula_positive, data = positive_data, family = gaussian())
    gam_models_positive[[predictor]] <- positive_model
    summary_models_positive[[predictor]] <- summary(positive_model)

    cat("Positive model for", predictor, "\n")
  }

  # print or plot the summaries here...
  print(summary(zero_model))
  print(summary(positive_model))
  plot(zero_model)
  plot(positive_model)
}
```

#### Based on the results above, the following have been identified as potential predictors of MIR:

##### This shows the model and the p-value
###### MIR ~ s(Hrs_Above_80_lag) 0.0398
###### MIR ~ s(Hrs_Above_75_lag) 0.0213
###### MIR ~ s(Hrs_Below_50_lag) 0.00489
###### MIR ~ s(Avg_HourlyDryBulbTemp_lag) 0.0301
###### MIR ~ s(Avg_HourlyWetBulbTemp_lag) 0.00928
###### MIR ~ s(Hrs_Below_50_current) 0.000593
###### MIR ~ s(Avg_HourlyWetBulbTemperature_current) 0.028
###### MIR ~ s(Avg_HourlyWindSpeed_current) 0.0369
###### MIR ~ s(MEL_Prop) 0.00867

###### I(MIR > 0) ~ s(Avg_HourlyWindSpeed_current) 0.00483
###### I(MIR > 0) ~ s(None.x) 0.000935
###### I(MIR > 0) ~ s(average_daylight_hours) 1.15e-06
######I(MIR > 0) ~ s(Average_MEL) 4.75e-05
######I(MIR > 0) ~ s(MEL_Prop) 5.11e-05
###### I(MIR > 0) ~ s(DSCI_ma) 0.0174

### There is some concern that dividing the data into two parts (observations >0 for part 2) may reduce the power of the analysis.

### Now use a GAM with all records, where MIR >= 0
```{r}
# Response variable
response_var <- "MIR"

# Specified predictors for modeling
specified_predictors <- c("Avg_TMAX", "Total_PRCP", "Hrs_Above_80_lag", "Hrs_Above_75_lag",
                          "Hrs_Below_50_lag", "Hrs_Between_50_86_lag", "Avg_HourlyDryBulbTemp_lag",
                          "Avg_HourlyWetBulbTemp_lag", "Hrs_Below_50_current", "DSCI_fema",
                          "Avg_HourlyRelativeHumidity_current", "Avg_HourlyWetBulbTemperature_current", 
                          "Avg_HourlyWindSpeed_current", "None.x", "None.y",  "DSCI_ma", "average_daylight_hours", 
                          "Average_MEL", "MEL_Prop")

# Initiate lists to store models and summaries
gam_models <- list()
summary_models <- list()

# Loop through each specified predictor
for (predictor in specified_predictors) {
  # Check if predictor exists in the dataset
  if (!predictor %in% names(mir_2part_all)) {
    cat("Skipping", predictor, "because it is not in the dataset.\n")
    next
  }

  # Model with all records
  formula <- as.formula(paste0(response_var, " ~ s(", predictor, ")"))
  model <- gam(formula, data = mir_2part_all, family = gaussian())
  gam_models[[predictor]] <- model
  summary_models[[predictor]] <- summary(model)

  cat("Model for", predictor, "\n")

  # Print or plot the summaries here...
  print(summary(model))
  plot(model)
}
```

### The all records model shows the following predictors as significant:
#### MIR ~ s(Hrs_Below_50_lag) 0.0105
#### MIR ~ s(Hrs_Below_50_current) 0.00333
#### MIR ~ s(DSCI_fema) 0.0214
#### MIR ~ s(Avg_HourlyRelativeHumidity_current) 0.00936
#### MIR ~ s(Avg_HourlyWindSpeed_current) 0.00366
#### MIR ~ s(average_daylight_hours) <2e-16
#### MIR ~ s(Average_MEL) 0.0258
#### MIR ~ s(MEL_Prop) <2e-16
## Give the concern with splitting the data further, we will continue with the all records model.

### Now that we have a better idea of the relationships and potential significance.
#### We will have another look at the correlation matrix for the predictors.

```{r}
library(reshape2) # or library(data.table)
library(plotly)

# Assuming continuous_vars is a dataframe with just the continuous variables you want to correlate
# Create a correlation matrix
cor_matrix <- cor(mir_2part %>% 
                    select(Hrs_Above_80_lag, Hrs_Above_75_lag, Hrs_Below_50_lag, 
                           Avg_HourlyDryBulbTemp_lag, Avg_HourlyWetBulbTemp_lag, 
                           Hrs_Below_50_current, Avg_HourlyWetBulbTemperature_current, 
                           Avg_HourlyWindSpeed_current, None.x, average_daylight_hours, 
                           Average_MEL, MEL_Prop, DSCI_ma),
                  use = "complete.obs")  # 'use' parameter handles missing values

# Melt the correlation matrix into a long format
cor_data <- melt(cor_matrix, na.rm = TRUE)

# Ensure cor_data has the expected column names after melting
if (!all(c("Var1", "Var2") %in% names(cor_data))) {
  stop("Var1 and Var2 not found in melted data. Check the melting process.")
}

# Create the interactive heatmap using plotly
heatmap_plot <- plot_ly(
  data = cor_data,
  x = ~Var1,
  y = ~Var2,
  z = ~value,
  type = 'heatmap',
  colorscale = 'Portland'
) %>%
  layout(
    title = 'Correlation Matrix',
    xaxis = list(title = 'Variable 1', tickangle = 45),
    yaxis = list(title = 'Variable 2', autorange = "reversed")
  )

# If you're running this in an interactive R session, this will display the plot
heatmap_plot
```
#### While correlation between predictors is a concern for the GAM model, it is not as much of a concern for the random forest. 
#### I will create a new variable that represents the difference between the average dry bulb temperature and the average wet bulb temperature.
```{r}
# Create a new dataframe by copying mir_2part
mir_2part_updated <- mir_2part

# Add the new variable to the new dataframe
mir_2part_updated$Dry_Wet_Diff <- mir_2part_updated$Avg_HourlyDryBulbTemp_lag - mir_2part_updated$Avg_HourlyWetBulbTemp_lag

# Now, mir_2part remains unchanged, and mir_2part_updated includes the new variable.
```

## 1 part GAM
```{r}
# Load necessary libraries
library(tidymodels)
library(dplyr)
library(mgcv) # For GAM
library(readr) # For read_csv

# Read the dataset
mir_1part_all <- read_csv(here("data", "processed-data", "all_mir_clean_joined.csv"))

# Define predictors for the continuous outcome
predictors_part2 <- c("Hrs_Below_50_lag", 
                      "Hrs_Below_50_current", "Avg_HourlyWetBulbTemperature_current", 
                      "Avg_HourlyWindSpeed_current", "MEL_Prop", "average_daylight_hours", "Average_MEL")

# Prepare the dataset for continuous outcome
mir_1part_all_continuous <- mir_1part_all %>%
  select(MIR, all_of(predictors_part2)) %>%
  na.omit()

# Split the data
set.seed(123)
data_split <- initial_split(mir_1part_all_continuous, prop = 0.75)
train_data_continuous <- training(data_split)
test_data_gam1 <- testing(data_split)

# Calculate the mean of the response variable for the entire dataset
mean_response <- mean(mir_1part_all_continuous$MIR)

# Create predictions using the mean response for the entire dataset
null_predictions <- rep(mean_response, nrow(mir_1part_all_continuous))

# Calculate RMSE for the null model
null_rmse <- sqrt(mean((null_predictions - mir_1part_all_continuous$MIR)^2))
cat("Null Model RMSE (Entire Dataset):", null_rmse, "\n")

# Training for Continuous Outcome
formula_gam1 <- as.formula(paste("MIR ~", paste("s(", predictors_part2, ")", collapse = " + ")))
model_gam1 <- gam(formula_gam1, data = train_data_continuous, family = gaussian())

# Evaluate on Training Set (Continuous)
predictions_train_gam1 <- predict(model_gam1, newdata = train_data_continuous)
rmse_train_gam1 <- sqrt(mean((predictions_train_gam1 - train_data_continuous$MIR)^2))
cat("Training Set RMSE (Continuous Outcome):", rmse_train_gam1, "\n")

# Evaluate on Test Set (Continuous Outcome)
predictions_test_gam1 <- predict(model_gam1, newdata = test_data_gam1)
rmse_test_gam1 <- sqrt(mean((predictions_test_gam1 - test_data_gam1$MIR)^2))
cat("Test Set RMSE (Continuous Outcome):", rmse_test_gam1, "\n")

# Visualize model effects for insights
plot(model_gam1)

### Boot strapping
# Define the bootstrapping statistic function
boot_statistic <- function(data, indices) {
  # Extract the bootstrap sample
  boot_sample <- data[indices, ]

  # Fit the GAM model on the bootstrap sample
  gam_boot <- gam(formula_gam1, data = boot_sample, family = gaussian())

  # Predict on the out-of-bag (OOB) data
  oob_data <- data[-indices, ]
  predictions_oob <- predict(gam_boot, newdata = oob_data)

  # Calculate RMSE on the OOB predictions
  sqrt(mean((predictions_oob - oob_data$MIR)^2))
}

# Perform bootstrapping
set.seed(123)  # Ensure reproducibility
bootstrap_results <- boot(
  data = train_data_continuous,
  statistic = boot_statistic,
  R = 1000  # For example, perform 1000 bootstrap replicates
)

# Extract the bootstrapped RMSE and calculate the mean and standard deviation
bootstrapped_rmse <- bootstrap_results$t
mean_bootstrapped_rmse <- mean(bootstrapped_rmse)
sd_bootstrapped_rmse <- sd(bootstrapped_rmse)

# Display bootstrapped RMSE results
cat("Bootstrapped RMSE: Mean=", mean_bootstrapped_rmse, ", SD=", sd_bootstrapped_rmse, "\n")

# # Calculate confidence intervals for the bootstrapped RMSE
# boot_ci <- boot.ci(bootstrap_results, type = "basic")
# cat("95% CI for RMSE: Lower =", boot_ci$basic[4], ", Upper =", boot_ci$basic[5], "\n")

# Calculate percentile confidence intervals for the bootstrapped RMSE
boot_ci_percentile <- boot.ci(bootstrap_results, type = "perc")

# Display the percentile confidence intervals for the bootstrapped RMSE
cat("95% CI for RMSE (Percentile): Lower =", boot_ci_percentile$percent[4], ", Upper =", boot_ci_percentile$percent[5], "\n")
```

### Residual plots for continuous outcome:
```{r}
# Step 1: Calculate residuals
train_residuals <- residuals(model_gam1, type = "response")

# Step 2: Create residual plot
plot(predictions_train_gam1, train_residuals,
     xlab = "Predicted Values",
     ylab = "Residuals",
     main = "Residual Plot (Training Set)")

# Step 3: Create predicted vs. actual plot for training set
plot(predictions_train_gam1, train_data_continuous$MIR,
     xlab = "Predicted Values",
     ylab = "Actual Values",
     main = "Predicted vs. Actual Plot (Training Set)")

plot(predictions_test_gam1, test_data_gam1$MIR,
     xlab = "Predicted Values",
     ylab = "Actual Values",
     main = "Predicted vs. Actual Plot (Test Set)")

```

## 1 Part Random Forest
```{r}
# Load the dataset
mir_1part_all <- read_csv(here("data", "processed-data", "all_mir_clean_joined.csv"))

# Define predictors for the continuous outcome
predictors <- c("Hrs_Below_50_lag", "Hrs_Below_50_current", 
                "Avg_HourlyWetBulbTemperature_current", "Avg_HourlyWindSpeed_current", 
                "MEL_Prop", "average_daylight_hours", "Average_MEL")

# Prepare the dataset for continuous outcome
train_data_continuous <- mir_1part_all %>%
  select(MIR, all_of(predictors)) %>%
  na.omit()

# Split the data
set.seed(123)
data_split <- initial_split(train_data_continuous, prop = 0.75)
train_data_continuous <- training(data_split)
test_data_continuous <- testing(data_split)

# Define the tuning grid
tuning_grid <- grid_regular(
  mtry(range = c(1, length(predictors))),
  min_n(range = c(5, 25)), # Adjusted based on preliminary insights
  levels = 10 # Increased resolution
)

# tuning_grid <- grid_regular(
#   mtry(range = c(1, length(predictors))),
#   min_n(range = c(1, 21)),
#   levels = 7
# )

# Define cross-validation folds
cv_folds_continuous <- vfold_cv(train_data_continuous, v = 5)

# Define the Random Forest model specification
rf_model_spec_continuous <- rand_forest(trees = 500, mtry = tune(), min_n = tune()) %>%
  set_engine("ranger", importance = "impurity", seed = 123) %>%
  set_mode("regression")

# Create the recipe
rf_recipe_continuous <- recipe(MIR ~ ., data = train_data_continuous)

# Workflow setup
rf_workflow_continuous <- workflow() %>%
  add_recipe(rf_recipe_continuous) %>%
  add_model(rf_model_spec_continuous)

# Tune the model
tuning_results_continuous <- tune_grid(
  rf_workflow_continuous,
  resamples = cv_folds_continuous,
  grid = tuning_grid,
  metrics = metric_set(yardstick::rmse)
)

# Extract the best hyperparameters
best_hps_continuous <- select_best(tuning_results_continuous, metric = "rmse")

# Fit the final model with the best hyperparameters on the full training dataset
final_fit_continuous <- fit(
  rf_workflow_continuous %>% finalize_workflow(best_hps_continuous),
  data = train_data_continuous
)

### Calculate RMSE for the training and testing sets

# Predict on the training and testing set
train_preds <- predict(final_fit_continuous, new_data = train_data_continuous)$.pred
test_preds <- predict(final_fit_continuous, new_data = test_data_continuous)$.pred

# You need to create tibbles that combine the actual values with the predictions
train_results <- tibble(
  truth = train_data_continuous$MIR,
  estimate = train_preds
)

test_results <- tibble(
  truth = test_data_continuous$MIR,
  estimate = test_preds
)

# Now, you can correctly calculate RMSE for both the training and testing sets
train_rmse_result_rf <- rmse(train_results, truth = truth, estimate = estimate)
test_rmse_result_rf <- rmse(test_results, truth = truth, estimate = estimate)

# Display the RMSE results
cat("Training Set RMSE:", train_rmse_result_rf$.estimate, "\n")
cat("Testing Set RMSE:", test_rmse_result_rf$.estimate, "\n")

# Continue with bootstrapping to evaluate model's robustness
set.seed(123) # for reproducibility
bootstrap_results <- boot(
  data = train_data_continuous,
  statistic = boot_fit_and_evaluate,
  R = 100 # number of bootstrap resamples
)

# Calculate the bootstrapped RMSE
mean_rmse_boot_rf <- mean(bootstrap_results$t)
sd_rmse_boot_rf <- sd(bootstrap_results$t)
ci_rmse_boot_rf <- boot.ci(bootstrap_results, type = "perc")  # Percentile interval

# Extracting the correct CI values for percentile CI
lower_bound <- ci_rmse_boot_rf$percent[4]
upper_bound <- ci_rmse_boot_rf$percent[5]

# Output the bootstrapping results
cat("Bootstrapped RMSE: Mean=", mean_rmse_boot_rf, ", SD=", sd_rmse_boot_rf, "\n")
cat("95% CI for RMSE (Percentile): ", lower_bound, "-", upper_bound, "\n")

autoplot(tuning_results_continuous)
```
### The random forest model has evidence of overfitting as the RMSE on the test set is higher than the training set.

## LASSO is not appropriate for non-linear relationships.

### The LASSO model suggests overfitting as well. The RMSE on the test set is higher than the training set.
### We will try modifying the tuning. 
```{r}
library(tidymodels)
library(readr) # For read_csv
library(yardstick) # For metrics

# Load the dataset
mir_1part_all <- read_csv(here("data", "processed-data", "all_mir_clean_joined.csv"))

# Define predictors for the continuous outcome
predictors <- c("Hrs_Below_50_lag", "Hrs_Below_50_current", 
                "Avg_HourlyWetBulbTemperature_current", "Avg_HourlyWindSpeed_current", 
                "MEL_Prop", "average_daylight_hours", "Average_MEL")

# Prepare the dataset for continuous outcome
train_data_continuous <- mir_1part_all %>%
  select(MIR, all_of(predictors)) %>%
  na.omit()

# Split the data
set.seed(123)
data_split <- initial_split(train_data_continuous, prop = 0.75)
train_data_continuous <- training(data_split)
test_data_continuous <- testing(data_split)

# Define a more refined tuning grid for LASSO with a wider range of penalty values
tuning_grid_lasso <- expand.grid(
  penalty = 10^seq(-10, 1, length.out = 50)  # Increase the range and resolution of the penalty parameter
)

# Increase the number of cross-validation folds for a more robust evaluation
cv_folds_continuous <- vfold_cv(train_data_continuous, v = 10)

# Define the LASSO model specification
lasso_model_spec <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

# Create the recipe
#lasso_recipe <- recipe(MIR ~ ., data = train_data_continuous)
lasso_recipe <- recipe(MIR ~ ., data = train_data_continuous) %>%
  update_role(all_of(predictors), new_role = "predictor") %>%
  step_poly(all_of(predictors), degree = 2)

# Workflow setup
lasso_workflow <- workflow() %>%
  add_recipe(lasso_recipe) %>%
  add_model(lasso_model_spec)

# Tune the model using the refined tuning grid and increased folds
tuning_results_lasso <- tune_grid(
  lasso_workflow,
  resamples = cv_folds_continuous,
  grid = tuning_grid_lasso,
  metrics = metric_set(yardstick::rmse)
)

# Extract the best hyperparameters
best_hps_lasso <- select_best(tuning_results_lasso, metric = "rmse")

# Fit the final model with the best hyperparameters on the full training dataset
final_fit_lasso <- fit(
  lasso_workflow %>% finalize_workflow(best_hps_lasso),
  data = train_data_continuous
)

# Predict on the training and testing set using the final LASSO model
train_preds_lasso <- predict(final_fit_lasso, new_data = train_data_continuous)$.pred
test_preds_lasso <- predict(final_fit_lasso, new_data = test_data_continuous)$.pred

# Calculate RMSE for the training and testing sets using the predictions
train_rmse_result_lasso <- yardstick::rmse(data.frame(truth = train_data_continuous$MIR, estimate = train_preds_lasso), truth = truth, estimate = estimate)
test_rmse_result_lasso <- yardstick::rmse(data.frame(truth = test_data_continuous$MIR, estimate = test_preds_lasso), truth = truth, estimate = estimate)

# Display the RMSE results
cat("Training Set RMSE (LASSO):", train_rmse_result_lasso$.estimate, "\n")
cat("Testing Set RMSE (LASSO):", test_rmse_result_lasso$.estimate, "\n")

# Visualize the tuning results
autoplot(tuning_results_lasso)

```
### There still seems to be overfitting. The RMSE on the test set is higher than the training set.
### We will try a different model

# Generalized Additive Models (GAMs) with sparsity
### GAMs are flexible models that can capture non-linear relationships between predictors and the outcome.
### Acts similar to LASSO adding a penalty similar to LASSO to enforce sparsity and variable selection
```{r}
library(tidymodels)
library(mgcv)
library(readr) # For read_csv

# Load the dataset
mir_1part_all <- read_csv(here("data", "processed-data", "all_mir_clean_joined.csv"))

# Prepare the dataset for continuous outcome
train_data_continuous <- mir_1part_all %>%
  select(MIR, Hrs_Below_50_lag, Hrs_Below_50_current, Avg_HourlyWetBulbTemperature_current, 
         Avg_HourlyWindSpeed_current, MEL_Prop, average_daylight_hours, Average_MEL) %>%
  na.omit()

# Split the data
set.seed(123)
data_split <- initial_split(train_data_continuous, prop = 0.75)
train_data_continuous <- training(data_split)
test_data_continuous <- testing(data_split)

# Define cross-validation folds
cv_folds_continuous <- vfold_cv(train_data_continuous, v = 5)

# Placeholder for storing CV results
cv_results <- vector("list", length(cv_folds_continuous$splits))

# Loop through CV folds
for (i in seq_along(cv_folds_continuous$splits)) {
  split <- cv_folds_continuous$splits[[i]]

  # Extract training and validation data for this fold
  train_split <- training(split)
  valid_split <- testing(split)

  # Fit a GAM with sparsity
  gam_model <- gam(MIR ~ s(Hrs_Below_50_lag) + s(Hrs_Below_50_current) + 
                     s(Avg_HourlyWetBulbTemperature_current) + s(Avg_HourlyWindSpeed_current) + 
                     s(MEL_Prop) + s(average_daylight_hours) + s(Average_MEL),
                   data = train_split, method = "REML", select = TRUE)

  # Make predictions on the validation set
  preds <- predict(gam_model, newdata = valid_split, type = "response")

  # Create a tibble that includes both the actual values and predictions
  results_tibble <- tibble(
    truth = valid_split$MIR,
    estimate = preds
  )

  # Calculate RMSE for this fold using the results tibble
  cv_results[[i]] <- rmse(results_tibble, truth = truth, estimate = estimate)
}

# First, verify the structure of the first element to ensure we understand it correctly
print(cv_results[[1]])

# If cv_results[[1]] is indeed a tibble with an .estimate column, proceed with the following:
avg_rmse_GAMS <- mean(sapply(cv_results, function(x) x$.estimate))

# Display the average RMSE
cat("Average CV RMSE:", avg_rmse_GAMS, "\n")

# Fit the final GAM model on the full training data
final_gam_model <- gam(MIR ~ s(Hrs_Below_50_lag) + s(Hrs_Below_50_current) + 
                         s(Avg_HourlyWetBulbTemperature_current) + s(Avg_HourlyWindSpeed_current) + 
                         s(MEL_Prop) + s(average_daylight_hours) + s(Average_MEL),
                       data = train_data_continuous, method = "REML", select = TRUE)

# Create predictions for the training set using the final GAM model
train_preds <- predict(final_gam_model, newdata = train_data_continuous, type = "response")

# Create a tibble that includes both the actual values and predictions for the training set
train_results <- tibble(
  truth = train_data_continuous$MIR,  # Actual values
  estimate = train_preds  # Predicted values
)

# Calculate RMSE for the training set
train_rmse_result_GAMS <- rmse(train_results, truth = truth, estimate = estimate)

# Make predictions for the test set using the final GAM model
test_preds <- predict(final_gam_model, newdata = test_data_continuous, type = "response")

# Create a tibble that includes both the actual values and predictions for the test set
test_results <- tibble(
  truth = test_data_continuous$MIR,  # Actual values
  estimate = test_preds  # Predicted values
)

# Calculate RMSE for the test set
test_rmse_result_GAMS <- rmse(test_results, truth = truth, estimate = estimate)

# Display the RMSE for the training set
cat("Training Set RMSE:", train_rmse_result_GAMS$.estimate, "\n")
# Display the RMSE for the test set
cat("Test Set RMSE:", test_rmse_result_GAMS$.estimate, "\n")

# Calculate AIC for the final GAM model
aic_value <- AIC(final_gam_model)

# Display the AIC value
cat("AIC for the final GAM model:", aic_value, "\n")

```
# Gradient Boosting Machine (GBM)
```{r}

library(tidymodels)
library(dplyr)
library(yardstick)
library(ranger) # For Random Forest
library(boot) # For bootstrapping

# Load the dataset
mir_1part_all <- read_csv(here("data", "processed-data", "all_mir_clean_joined.csv"))

# Define predictors for the continuous outcome
predictors <- c("Hrs_Below_50_lag", "Hrs_Below_50_current", 
                "Avg_HourlyWetBulbTemperature_current", "Avg_HourlyWindSpeed_current", 
                "MEL_Prop", "average_daylight_hours", "Average_MEL")

# Prepare the dataset for continuous outcome
train_data_continuous <- mir_1part_all %>%
  select(MIR, all_of(predictors)) %>%
  na.omit()

# Split the data
set.seed(123)
data_split <- initial_split(train_data_continuous, prop = 0.75)
train_data_continuous <- training(data_split)
test_data_continuous <- testing(data_split)

# Define the tuning grid
tuning_grid <- grid_regular(
  trees(range = c(50, 500)), # Adjusted for GBM
  tree_depth(range = c(1, 5)), # Shallow trees
  learn_rate(range = c(0.01, 0.1)), # Learning rate for gradient descent
  min_n(range = c(5, 25)), # Adjusted based on preliminary insights
  levels = 10 # Increased resolution
)

# Define cross-validation folds
cv_folds_continuous <- vfold_cv(train_data_continuous, v = 5)

# Define the GBM model specification
gbm_model_spec_continuous <- boost_tree(trees = tune(), tree_depth = tune(), learn_rate = tune(), min_n = tune()) %>%
  set_engine("xgboost", objective = "reg:squarederror", nthread = parallel::detectCores()) %>%
  set_mode("regression")

# Create the recipe
gbm_recipe_continuous <- recipe(MIR ~ ., data = train_data_continuous)

# Workflow setup
gbm_workflow_continuous <- workflow() %>%
  add_recipe(gbm_recipe_continuous) %>%
  add_model(gbm_model_spec_continuous)

# Tune the model
tuning_results_continuous <- tune_grid(
  gbm_workflow_continuous,
  resamples = cv_folds_continuous,
  grid = tuning_grid,
  metrics = metric_set(yardstick::rmse)
)

# Extract the best hyperparameters
best_hps_continuous <- select_best(tuning_results_continuous, metric = "rmse")

# Fit the final model with the best hyperparameters on the full training dataset
final_fit_continuous <- fit(
  gbm_workflow_continuous %>% finalize_workflow(best_hps_continuous),
  data = train_data_continuous
)

### Calculate RMSE for the training and testing sets

# Predict on the training and testing set
train_preds <- predict(final_fit_continuous, new_data = train_data_continuous)$.pred
test_preds <- predict(final_fit_continuous, new_data = test_data_continuous)$.pred

# You need to create tibbles that combine the actual values with the predictions
train_results <- tibble(
  truth = train_data_continuous$MIR,
  estimate = train_preds
)

test_results <- tibble(
  truth = test_data_continuous$MIR,
  estimate = test_preds
)

# Now, you can correctly calculate RMSE for both the training and testing sets
train_rmse_result_gbm <- rmse(train_results, truth = truth, estimate = estimate)
test_rmse_result_gbm <- rmse(test_results, truth = truth, estimate = estimate)

# Display the RMSE results
cat("Training Set RMSE:", train_rmse_result_gbm$.estimate, "\n")
cat("Testing Set RMSE:", test_rmse_result_gbm$.estimate, "\n")

# Define a function to fit the model and calculate RMSE
boot_fit_and_evaluate <- function(data, indices) {
  # Extract the bootstrap sample
  bootstrap_data <- data[indices, ]

  # Fit the model on the bootstrap sample
  model <- gbm(MIR ~ ., data = bootstrap_data, distribution = "gaussian",
               n.trees = 1000, interaction.depth = 4, shrinkage = 0.01)

  # Predict on the original dataset
  preds <- predict(model, newdata = data, n.trees = 1000)

  # Compute RMSE
  rmse <- sqrt(mean((preds - data$MIR)^2))

  # Return RMSE
  return(rmse)
}

# Continue with bootstrapping to evaluate model's robustness
set.seed(123) # for reproducibility
bootstrap_results <- boot(
  data = train_data_continuous,
  statistic = boot_fit_and_evaluate,
  R = 100 # number of bootstrap resamples
)

# Calculate the bootstrapped RMSE
mean_rmse <- mean(bootstrap_results$t)
sd_rmse <- sd(bootstrap_results$t)
ci_rmse <- boot.ci(bootstrap_results, type = "perc")  # Percentile interval

# Extracting the correct CI values for percentile CI
lower_bound <- ci_rmse$percent[4]
upper_bound <- ci_rmse$percent[5]

# Output the bootstrapping results
cat("Bootstrapped RMSE: Mean=", mean_rmse, ", SD=", sd_rmse, "\n")
cat("95% CI for RMSE (Percentile): ", lower_bound, "-", upper_bound, "\n")

autoplot(tuning_results_continuous)

```

# SUmmarize the RSME values
```{r}
rmse_table <- data.frame(
  Model = c("Null Model", "Generalized Additive Model", "Random Forest", "LASSO",
            "Generalized Additive Model with Sparsity", "Gradient Boosting Machine"),
  Train_RMSE = c(null_rmse, rmse_train_gam1, train_rmse_result_rf$.estimate, 
                 train_rmse_result_lasso$.estimate, train_rmse_result_GAMS$.estimate, 
                 train_rmse_result_gbm$.estimate),
  Test_RMSE = c(null_rmse, rmse_test_gam1, test_rmse_result_rf$.estimate, 
                test_rmse_result_lasso$.estimate, test_rmse_result_GAMS$.estimate, 
                test_rmse_result_gbm$.estimate)
)
# Sort the table by Test_RMSE column in descending order
rmse_table <- rmse_table[order(-rmse_table$Test_RMSE), ]

# Print the sorted table
print(rmse_table)

#save as an RDS file
saveRDS(rmse_table, here("results", "tables", "rmse_table.rds"))

```

### 2024-04-21 R/eda-code/EDA_unified.qmd
Part 5
Analysis is complete. Readme files and removal of outdated datasets and codes will be removed throughout the week. Additionally, the manuscript will be finalized this weel

@@
 title: "Class Project"
-author: Andrew Ruiz
+author: Andrew J. Ruiz
 date: "2024-03-13"
@@
 library(RColorBrewer)
+library(ggcorrplot)
+library(readxl)
+library(corrplot)
+library(plotly)
+library(reshape2)
+library(rlang)
 ```
@@
 # This file will contain the EDA code for the class project.
-### As the project progresses and the models finalized, the EDA QMD will be updated accordingly.
 
-## MIR heatmap
-### This heat map illustrates the timing and intensity of mosquito infection rates (MIR) over the weeks of the year.
-#### At least one previous study that a non-zero MIR at or before week 31 is common in years when human cases of EEE are reported. 
+# Calculate Minimum Infection Rate (MIR) by trap type and by week
+### The Minimum Infection Rate (MIR) is a common metric used to estimate the prevalence of mosquito-borne viruses in a population.
+It is calculated as the number of positive mosquito pools per 1,000 mosquitoes tested.
+The MIR can provide insights into the risk of virus transmission to humans and animals.
+#### Mosquito traps are designed to attract different life stages and species of mosquitoes.
+For this project, the focus will be on CDC Light Traps baited with CO2.
+These traps are used to trap a wide range of species, including the main zoonotic vector of EEE, Culiseta melanura (MEL).
 ```{r}
-# Load the saved function
-create_heatmap <- readRDS(here("data", "processed-data", "rds","create_heatmap_function.rds"))
+# Read the virus isolation dataset
+pcr_data <- read_excel(here("data", "raw-data", "virus_iso.xlsx"))
 
-# Assuming MIR_by_week is loaded and prepared in this QMD as well
-# Now use the loaded function to create the heatmap
-MIR_by_week_heatmap <- create_heatmap(MIR_by_week)
+# Add week number based on the collection data
+# This code ensures that the ISO 8601 standard is applied to the week number calculation
+# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
+# Weeks start on Monday and end on Sunday.
+# The last week of the year, week 52 or 53, is the one that contains December 28th.
+# we will use week_num to join tables
+# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08
+pcr_data <- pcr_data %>%
+  mutate(
+    # Directly format the collection date into "YYYY-WW"
+    week_num = paste0(year(`Collection Date`), "-", 
+                      sprintf("%02d", isoweek(`Collection Date`)))
+  )
 
-# Display the heatmap
-MIR_by_week_heatmap
+filtered_pcr_data <- pcr_data %>%
+  #filter(`Trap Type` %in% trap_types_to_include) %>%
+  mutate(
+    Submitted_for_Testing = ifelse(`Submitted for Testing` == "Y", 1, 0),  # Convert to numeric
+    Is_Positive = ifelse(Result == "Positive", 1, 0),  # Convert Positive results to numeric
+    Pool_Size_Numeric = as.numeric(`Pool Size`)  # Ensure Pool Size is numeric for summing - Corrected
+  )
+
+#calculate MIR for all sites, all traps by week
+MIR_by_week <- filtered_pcr_data %>%
+  group_by(week_num) %>%
+  summarise(
+    Pools_Tested = sum(Submitted_for_Testing, na.rm = TRUE),
+    Positive_Pools = sum(Is_Positive, na.rm = TRUE),
+    Total_Tested = sum(ifelse(Submitted_for_Testing == 1, Pool_Size_Numeric, 0), na.rm = TRUE),  # Sum Pool Size for submitted tests - Corrected
+    .groups = 'drop'
+  ) %>%
+  filter(Pools_Tested > 0) %>%  # Filter out weeks where no pools were tested
+  mutate(
+    MIR = ifelse(Total_Tested > 0, (Positive_Pools / Total_Tested) * 1000, 0),  # Calculate MIR
+    Week = as.numeric(substring(week_num, 6)), 
+    Year = as.numeric(substring(week_num, 1, 4))  # Extract week number and year
+  )
+
+# Save as CSV
+write_csv(MIR_by_week, here("data", "processed-data", "MIR_by_week.csv"))
+
+# Save as RDS
+saveRDS(MIR_by_week, here("data", "processed-data", "rds", "MIR_by_week.rds"))
+View(MIR_by_week)
+
+#CALCULATE MIR BY TRAP TYPE
+# Define the trap types to include
+trap_types_to_include <- c("500cc CO2 - CDC Miniature Light Trap", "Dry Ice CO2 - CDC Miniature Light Trap with CO2")
+
+# Filter pcr_data for specific CO2 baited light traps
+filtered_pcr_data_co2 <- pcr_data %>%
+  filter(`Trap Type` %in% trap_types_to_include) %>%
+  mutate(
+    Submitted_for_Testing = ifelse(`Submitted for Testing` == "Y", 1, 0),  # Convert to numeric
+    Is_Positive = ifelse(Result == "Positive", 1, 0),  # Convert Positive results to numeric
+    Pool_Size_Numeric = as.numeric(`Pool Size`)  # Ensure Pool Size is numeric for summing - Corrected
+  )
+
+MIR_by_week_co2 <- filtered_pcr_data_co2 %>%
+  group_by(week_num) %>%
+  summarise(
+    Pools_Tested = sum(Submitted_for_Testing, na.rm = TRUE),
+    Positive_Pools = sum(Is_Positive, na.rm = TRUE),
+    Total_Tested = sum(ifelse(Submitted_for_Testing == 1, Pool_Size_Numeric, 0), na.rm = TRUE),  # Sum Pool Size for submitted tests - Corrected
+    .groups = 'drop'
+  ) %>%
+  filter(Pools_Tested > 0) %>%  # Filter out weeks where no pools were tested
+  mutate(
+    MIR = ifelse(Total_Tested > 0, (Positive_Pools / Total_Tested) * 1000, 0),  # Calculate MIR
+    Week = as.numeric(substring(week_num, 6)), 
+    Year = as.numeric(substring(week_num, 1, 4))  # Extract week number and year
+  )
+
+# Save the calculated MIR for CO2 baited traps by week
+write_csv(MIR_by_week_co2, here("data", "processed-data", "MIR_by_week_co2.csv"))
+saveRDS(MIR_by_week_co2, here("data", "processed-data", "rds", "MIR_by_week_co2.rds"))
+
+# Calculate MIR for MEL only
+# Filter pcr_data for MEL only
+pcr_mel_co2 <- pcr_data %>%
+  filter(`Trap Type` %in% trap_types_to_include) %>% # Filter for CO2 baited traps
+  filter(`Species code` == "MEL") %>% # Filter for pool tests
+  mutate(
+    Submitted_for_Testing = ifelse(`Submitted for Testing` == "Y", 1, 0), # Convert to numeric
+    Is_Positive = ifelse(Result == "Positive", 1, 0), # Ensure Is_Positive is numeric
+    Pool_Size_Numeric = as.numeric(`Pool Size`)  # Ensure Pool Size is numeric for summing - Corrected
+  )
+
+mir_mel_week_co2 <- pcr_mel_co2 %>%
+  group_by(week_num) %>%
+  summarise(
+    Pools_Tested = sum(Submitted_for_Testing, na.rm = TRUE),
+    Positive_Pools = sum(Is_Positive, na.rm = TRUE),
+    Total_Tested = sum(ifelse(Submitted_for_Testing == 1, Pool_Size_Numeric, 0), na.rm = TRUE),  # Sum Pool Size for submitted tests - Corrected
+    .groups = 'drop'
+  ) %>%
+  filter(Pools_Tested > 0) %>%  # Filter out weeks where no pools were tested
+  mutate(
+    MIR = ifelse(Total_Tested > 0, (Positive_Pools / Total_Tested) * 1000, 0),  # Calculate MIR
+    Week = as.numeric(substring(week_num, 6)), 
+    Year = as.numeric(substring(week_num, 1, 4))  # Extract week number and year
+  )
+
+# Save the calculated MIR for MEL in CO2 baited traps by week
+write_csv(mir_mel_week_co2, here("data", "processed-data", "mir_mel_week_co2.csv"))
+saveRDS(mir_mel_week_co2, here("data", "processed-data", "rds", "mir_mel_week_co2.rds"))
+
+# Calculate MIR for all other species
+# filter to remove MEL
+pcr_nomel_co2 <- pcr_data %>%
+  filter(`Trap Type` %in% trap_types_to_include) %>% # Filter for CO2 baited traps
+  filter(`Species code` != "MEL") %>% # Filter for pool tests
+  mutate(
+    Submitted_for_Testing = ifelse(`Submitted for Testing` == "Y", 1, 0), # Convert to numeric
+    Is_Positive = ifelse(Result == "Positive", 1, 0), # Ensure Is_Positive is numeric
+    Pool_Size_Numeric = as.numeric(`Pool Size`)  # Ensure Pool Size is numeric for summing - Corrected
+  )
+
+mir_nomel_week_co2 <- pcr_nomel_co2 %>%
+  group_by(week_num) %>%
+  summarise(
+    Pools_Tested = sum(Submitted_for_Testing, na.rm = TRUE),
+    Positive_Pools = sum(Is_Positive, na.rm = TRUE),
+    Total_Tested = sum(ifelse(Submitted_for_Testing == 1, Pool_Size_Numeric, 0), na.rm = TRUE),  # Sum Pool Size for submitted tests - Corrected
+    .groups = 'drop'
+  ) %>%
+  filter(Pools_Tested > 0) %>%  # Filter out weeks where no pools were tested
+  mutate(
+    MIR = ifelse(Total_Tested > 0, (Positive_Pools / Total_Tested) * 1000, 0),  # Calculate MIR
+    Week = as.numeric(substring(week_num, 6)), 
+    Year = as.numeric(substring(week_num, 1, 4))  # Extract week number and year
+  )
+
+# Save the calculated MIR for all other species in CO2 baited traps by week
+write_csv(mir_nomel_week_co2, here("data", "processed-data", "mir_nomel_week_co2.csv"))
+saveRDS(mir_nomel_week_co2, here("data", "processed-data", "rds", "mir_nomel_week_co2.rds"))
+
+# Finally, we will calculate MIR by year.
+# this will be used for a second model
+mir_year <- filtered_pcr_data_co2 %>%
+  mutate(
+    Year = as.numeric(substring(week_num, 1, 4))  # Extract the year
+  ) %>%
+  group_by(Year) %>%
+  summarise(
+    Pools_Tested = sum(Submitted_for_Testing, na.rm = TRUE),
+    Positive_Pools = sum(Is_Positive, na.rm = TRUE),
+    Total_Tested = sum(ifelse(Submitted_for_Testing == 1, Pool_Size_Numeric, 0), na.rm = TRUE),  # Correctly sum Pool Size for submitted tests
+    .groups = 'drop'
+  ) %>%
+  filter(Pools_Tested > 0) %>%  # Filter out years where Pools_Tested is 0
+  mutate(
+    MIR = ifelse(Total_Tested > 0, (Positive_Pools / Total_Tested) * 1000, 0)  # Calculate MIR
+  )
+filtered_pcr_data_co2
+mir_year
 ```
 
-## This is a scatter plot of the average number of mosquitoes that were captured in light traps and the mosquito infection rate (MIR).
+# Join the MIR data with ither datasets based on week number
+#### First we will process some data that will be used in the join
 ```{r}
-# Load the function from the RDS file
-MIRMEL_scatter_plot <- readRDS(here("data", "processed-data", "RDS", "create_scatter_plot_function.rds"))
+# Load the datasets:
+MIR_by_week_co2 <- read_csv(here("data", "processed-data", "MIR_by_week_co2.csv"))
+wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
+mosquito_wide <- read_csv(here("data", "processed-data", "mosquito_wide_all.csv"))
+drought_fema1_4 <- read_csv(here("data", "processed-data", "drought_fema1_4.csv"))
+drought_ma_counties <- read_csv(here("data", "processed-data", "drought_ma_counties.csv"))
+average_daylight_by_week <- read_csv(here("data", "processed-data", "average_daylight_by_week.csv"))
+average_mel_weekco2 <- read_csv(here("data", "processed-data", "average_mel_weekco2.csv"))
+wx_hourly_clean <- read_csv(here("data", "processed-data", "hourly_wx_clean.csv"))
+mel_prop_weekly <- read_csv(here("data", "processed-data", "mel_proportion_by_week.csv"))
 
-# Assuming combined_data_mir_all is available in your current session
-# Generate the plot
-scatter_plot <- create_scatter_plot(combined_data_mir_all)
+# Before joining, there are some modifications to make:
+# Aggregate weather data by week:
+# Calculate the average maximum temperature (Avg_TMAX) and total precipitation (Total_PRCP) for each week.
+# This simplification allows for weekly comparisons with the MIR data.
+wx_aggregated <- wx_selected %>%
+  group_by(week_num) %>%
+  summarise(Avg_TMAX = mean(TMAX, na.rm = TRUE),
+            Total_PRCP = sum(PRCP, na.rm = TRUE),
+            .groups = 'drop')
+
+# Adjust for a one-week lag in weather data:
+# Create a new column (Lead_week_num) that represents each week's preceding week. 
+# This adjustment accounts for the delayed effect weather might have on MIR.
+wx_aggregated_lead <- wx_aggregated %>%
+  mutate(
+    # Split week_num to get year and week as numeric
+    year = as.integer(substr(week_num, 1, 4)),
+    week = as.integer(substr(week_num, 6, 8)),
+    # Calculate the next week, accounting for year transition
+    Lead_week_num = pmap_chr(list(year, week), function(year, week) {
+      last_week_of_year = ifelse(lubridate::isoweek(lubridate::ymd(paste(year, "12", "28"))) == 53, 53, 52)
+      if (week == last_week_of_year) {
+        # Transition to the first week of the next year
+        lead_year = year + 1
+        paste0(lead_year, "-", sprintf("%02d", 1))
+      } else {
+        # Just add one to the week for the same year
+        paste0(year, "-", sprintf("%02d", week + 1))
+      }
+    })
+  ) %>%
+  select(-year, -week) # Removing intermediate columns
+
+str(wx_aggregated_lead)
+
+# Calculate a new colum in hourly weather data for number of hours the dry bulb temperature is above 75 and 80 degrees
+# Calculating the conditions and averaging other columns
+wx_hourly <- wx_hourly_clean %>%
+  group_by(week_num) %>%
+  summarise(
+    Hrs_Above_80_lag = sum(HourlyDryBulbTemperature >= 80, na.rm = TRUE),
+    Hrs_Above_75_lag = sum(HourlyDryBulbTemperature >= 75, na.rm = TRUE),
+    Hrs_Below_50_lag = sum(HourlyDryBulbTemperature < 50, na.rm = TRUE),
+    Hrs_Between_50_86_lag = sum(HourlyDryBulbTemperature >= 50 & HourlyDryBulbTemperature <= 86, na.rm = TRUE), # Added line for temperature between 50 and 86
+    Avg_HourlyDryBulbTemp_lag = mean(HourlyDryBulbTemperature, na.rm = TRUE),
+    Avg_HourlyWetBulbTemp_lag = mean(HourlyWetBulbTemperature, na.rm = TRUE),
+    .groups = 'drop'
+  )
+wx_hourly_lead <- wx_hourly %>%
+  mutate(
+    Year = as.integer(substr(week_num, 1, 4)),
+    Week = as.integer(substr(week_num, 6, 7)),
+    Lead_week_num = map2_chr(Year, Week, ~ {
+      last_week_of_year = ifelse(isoweek(ymd(paste(.x, "12", "28"))) == 53, 53, 52)
+      if (.y == last_week_of_year) {
+        lead_year = .x + 1
+        paste0(lead_year, "-", sprintf("%02d", 1))
+      } else {
+        paste0(.x, "-", sprintf("%02d", .y + 1))
+      }
+    })
+  )
+
+# Define the path for the updated CSV file
+updated_csv_path <- here("data", "processed-data", "wx_hourly_lead.csv")
+
+# Save the updated dataframe
+write.csv(wx_hourly_lead, updated_csv_path, row.names = FALSE)
+
+# There are some other predictors from the same hourly dataset that are likely to affect the MIR that do not need a lag
+# Calculate the # of hours below 50, average relative humidity, wet bulb temperature, and wind speed for each week
+wx_hourly_current <- wx_hourly_clean %>%
+  group_by(week_num) %>%
+  summarise(
+    Hrs_Below_50_current = sum(HourlyDryBulbTemperature < 50, na.rm = TRUE),
+    Avg_HourlyRelativeHumidity_current = mean(HourlyRelativeHumidity, na.rm = TRUE),
+    Avg_HourlyWetBulbTemperature_current = mean(HourlyWetBulbTemperature, na.rm = TRUE),
+    Avg_HourlyWindSpeed_current = mean(HourlyWindSpeed, na.rm = TRUE),
+    .groups = 'drop'
+  )
+
+#save as csv
+write_csv(wx_hourly_current, here("data", "processed-data", "wx_hourly_current.csv"))
+
+```
+
+#### Now we will join the MIR data with the other datasets based on the week number
+```{r}
+combined_data_mir_all <- MIR_by_week_co2 %>%
+  # Joining wx_aggregated and wx_hourly_summary on a lagged week_num
+  left_join(wx_aggregated_lead, by = c("week_num" = "Lead_week_num")) %>%
+  left_join(wx_hourly_lead, by = c("week_num" = "Lead_week_num")) %>%
+  # Direct joins for other datasets
+  left_join(wx_hourly_current %>% distinct(week_num, .keep_all = TRUE), by = "week_num") %>%
+  left_join(drought_fema1_4 %>% distinct(week_num, .keep_all = TRUE), by = "week_num") %>%
+  left_join(drought_ma_counties %>% distinct(week_num, .keep_all = TRUE), by = "week_num") %>%
+  left_join(average_daylight_by_week %>% distinct(week_num, .keep_all = TRUE), by = "week_num") %>%
+  left_join(average_mel_weekco2 %>% distinct(week_num, .keep_all = TRUE), by = "week_num") %>%
+  left_join(mel_prop_weekly %>% distinct(week_num, .keep_all = TRUE), by = "week_num")
+
+# Verify the structure and first few rows to ensure accuracy
+head(combined_data_mir_all)
+
+# Further cleaning and processing of the combined data
+# Remove unnecessary columns and keep only the first value for each week
+combined_data_mir_all_clean <- combined_data_mir_all %>%
+  ungroup() %>% # Ensure data is not grouped before proceeding
+  dplyr::select(
+    -Year.y, -MapDate.x, -ValidEnd.x, -MapDate.y, -State, 
+    -StatisticFormatID.y, -week_num.y, -Week.y, -RegionName, -Year.x,
+    -StatisticFormatID.x, -FIPS, -Week.x, -ValidStart.x, -ValidStart.y, -County, -ValidEnd.y
+  ) #%>%
+  #summarise(across(everything(), first), .groups = 'drop')
+str(combined_data_mir_all_clean)
+# Save the modified dataframe as a CSV
+write_csv(combined_data_mir_all_clean, here("data", "processed-data", "joined_data_mir_co2.csv"))
+
+```
+### In addition to the MIR data for CO2 traps only, we will also create a dataset for all trap types. 
+```{r}
+all_MIR <- read_csv(here("data", "processed-data", "MIR_by_week.csv"))
+
+#join with other datasets
+all_mir <- all_MIR %>%
+  # Joining wx_aggregated and wx_hourly_summary on a lagged week_num
+  left_join(wx_aggregated_lead, by = c("week_num" = "Lead_week_num")) %>%
+  left_join(wx_hourly_lead, by = c("week_num" = "Lead_week_num")) %>%
+  # Direct joins for other datasets
+  left_join(wx_hourly_current %>% distinct(week_num, .keep_all = TRUE), by = "week_num") %>%
+  left_join(drought_fema1_4 %>% distinct(week_num, .keep_all = TRUE), by = "week_num") %>%
+  left_join(drought_ma_counties %>% distinct(week_num, .keep_all = TRUE), by = "week_num") %>%
+  left_join(average_daylight_by_week %>% distinct(week_num, .keep_all = TRUE), by = "week_num") %>%
+  left_join(average_mel_weekco2 %>% distinct(week_num, .keep_all = TRUE), by = "week_num") %>%
+  left_join(mel_prop_weekly %>% distinct(week_num, .keep_all = TRUE), by = "week_num")
+
+# Verify the structure and first few rows to ensure accuracy
+head(all_mir)
+
+# Now we will clean the data as above
+all_mir_clean <- all_mir %>%
+  ungroup() %>% # Ensure data is not grouped before proceeding
+  dplyr::select(
+    -Year.y, -MapDate.x, -ValidEnd.x, -MapDate.y, -State, 
+    -StatisticFormatID.y, -week_num.y, -Week.y, -RegionName, -Year.x,
+    -StatisticFormatID.x, -FIPS, -Week.x, -ValidStart.x, -ValidStart.y, -County, -ValidEnd.y
+  ) #%>%
+  #summarise(across(everything(), first), .groups = 'drop')
+str(all_mir_clean)
+# Save the modified dataframe as a CSSV
+write_csv(all_mir_clean, here("data", "processed-data", "all_mir_clean_joined.csv"))
+
+```
+
+## This is a scatter plot of the average number of MEL mosquitoes that were captured in light traps and the mosquito infection rate (MIR).
+```{r}
+joined_mir <- read_csv(here("data", "processed-data", "joined_data_mir_co2.csv"))
+
+# code to create a function for scatter plot
+create_scatter_plot <- function(data, x_var, y_var) {
+  # Ensure the variables exist in the dataframe
+  if (!(x_var %in% names(data)) || !(y_var %in% names(data))) {
+    stop("One or both of the specified variables do not exist in the dataframe.")
+  }
+
+  # Generate the scatter plot using the .data pronoun for NSE
+  p <- ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]])) +
+    geom_point() + # Plot points
+    geom_smooth(method = "loess", color = "blue") + # Add a LOESS smoothed curve
+    theme_minimal() +
+    labs(x = x_var, y = y_var, title = paste("Scatter Plot of", x_var, "vs.", y_var))
+
+  print(p)
+}
+
+# Example usage with your dataframe
+create_scatter_plot(joined_mir, "Hrs_Above_80_lag", "MIR")
+
+# DSCI_ma is an indicators of drought severity. The higher the number the more severe and widespread the drought.
+create_scatter_plot(joined_mir, "DSCI_ma", "MIR")
+
+# Propotion of MEL mosquitoes captured vs MIR
+create_scatter_plot(joined_mir, "MEL_Prop", "MIR")
+
+# This is the average high temperature for the week before the mosquito was collected
+create_scatter_plot(joined_mir, "Avg_TMAX", "MIR")
+
+# This is the total precipitation for the week before the mosquito was collected
+create_scatter_plot(joined_mir, "Total_PRCP", "MIR")
+
+# This variable is the number of hours the temperature is between 50 and 86 degrees.
+Below 50, mosquitoes are less active.
+The effect of temperature above 86 on mosquitoes does not increase developemt speed.
+create_scatter_plot(joined_mir, "Hrs_Between_50_86_lag", "MIR")
+
+# This is the average relative humidity for the week the mosquito was collected
+create_scatter_plot(joined_mir, "Avg_HourlyRelativeHumidity_current", "MIR")
+
+# This is the average wind speed for the week the mosquito was collected
+create_scatter_plot(joined_mir, "Avg_HourlyWindSpeed_current", "MIR")
+
+# This is the average dry bulb temperature for the week before the mosquito was collected
+create_scatter_plot(joined_mir, "Avg_HourlyDryBulbTemp_lag", "MIR")
+
+# this is the total # of mosquitoes tested for the week
+# this should not be a predictor but it is interesting to see if there is a relationship
+# it might be better to calculate the average count per trap
+create_scatter_plot(joined_mir, "Total_Tested", "MIR")
+```
+
+#Distribution of MIR
+### Before deciciding on models, we should check how the outcome data is distributed.
+#### Load MIR data
+```{r}
+# Load the dataset:
+joined_mir <- read_csv(here("data", "processed-data", "joined_data_mir_co2.csv"))
+```
+#### Remove rows with missing cells 
+```{r}
+# # check for missing MIR values
+# na_rows_mir = joined_mir[is.na(joined_mir$MIR), ]
+# head(na_rows_mir)  # To view the first few rows
+# 
+# # Remove rows with missing MIR values
+# joined_mir <- joined_mir[!is.na(joined_mir$MIR), ]
+# 
+# 
+# #save the cleaned data
+# write_csv(joined_mir, here("data", "processed-data", "joined_data_mir_co2_cleaned.csv"))
+```
+
+### Visualize the distribution of MIR
+```{r}
+p <- ggplot(joined_mir, aes(x = MIR)) +
+  geom_density(fill = "tomato", alpha = 0.7) +  # Adjust color and transparency
+  theme_minimal() +  # Cleaner theme
+  labs(title = "Density Plot of MIR",
+       x = "MIR",
+       y = "Density") +
+  theme(plot.title = element_text(hjust = 0.5),  # Center the title
+        text = element_text(size = 12))  # Adjust text size
 
 # Display the plot
-print(scatter_plot)
+print(p)
+
+#save the plot
+ggsave(here("results", "figures", "mir_density_plot.png"), plot = p, width = 8, height = 6, units = "in")
 
 ```
+#### Based on the histogram and density plot, the outcome of interest (MIR) appears to be right-skewed and not normally distributed.
+This may impact the choice of modeling techniques and the interpretation of results.
+
+### Check for excess zeros and over dispersion
+```{r}
+
+# Check for excess zeros
+zero_proportion <- sum(joined_mir$MIR == 0) / nrow(joined_mir)
+paste("Proportion of zeros:", round(zero_proportion, 4))
+
+# Calculate the mean and variance of MIR
+mir_mean <- mean(joined_mir$MIR, na.rm = TRUE)
+mir_variance <- var(joined_mir$MIR, na.rm = TRUE)
+
+# Now print the mean and variance
+cat("Mean of MIR:", mir_mean, "\n")
+cat("Variance of MIR:", mir_variance, "\n")
+
+# Calculate the variance-to-mean ratio (VMR)
+vmr <- mir_variance / mir_mean
+cat("Variance-to-Mean Ratio (VMR) of MIR:", vmr, "\n")
+
+# Check for overdispersion
+if(vmr > 1) {
+  cat("Evidence of overdispersion: VMR is greater than 1.\n")
+} else {
+  cat("No evidence of overdispersion: VMR is not greater than 1.\n")
+}
+```
+#### Given the excess zeros and the overdispersion in the MIR data, it may be appropriate to consider a Generalized Additive Model (GAM). GAM allow for 2-part models that can models 
+# Predictors
+#### Closer look at the possible predictors for the model
+```{r}
+# Load the dataset:
+# combined_data_mir_all: Contains weekly MIR data and possible predictors for the model.
+joined_mir <- read_csv(here("data", "processed-data", "joined_data_mir_co2.csv"))
+
+# Assuming 'joined_mir' is your dataframe
+continuous_vars <- joined_mir[, c("Avg_TMAX", "Total_PRCP", "Hrs_Above_80_lag", 
+                                  "Hrs_Above_75_lag", "Hrs_Below_50_lag", 
+                                  "Avg_HourlyDryBulbTemp_lag", "Avg_HourlyRelativeHumidity_current", 
+                                  "Avg_HourlyWetBulbTemperature_current", "Avg_HourlyWindSpeed_current", 
+                                  "DSCI_fema", "average_daylight_hours", 
+                                  "Average_MEL", "Total_MEL", "MEL_Prop")]
+
+# Create a correlation matrix
+cor_matrix <- cor(continuous_vars, use = "complete.obs")  # 'use' parameter handles missing values
+
+#cor_matrix[abs(cor_matrix) < 0.5] <- NA
+
+# Ensure cor_matrix is a correct correlation matrix
+if (!is.matrix(cor_matrix)) {
+  cor_matrix <- as.matrix(cor_matrix)
+}
+
+# Melt the correlation matrix into a long format
+cor_data <- melt(cor_matrix, na.rm = TRUE)
+
+# Check that Var1 and Var2 exist
+if (!all(c("Var1", "Var2") %in% names(cor_data))) {
+  stop("Var1 and Var2 not found in melted data. Check the melting process.")
+}
+
+# Create the interactive heatmap
+plot_ly(
+  data = cor_data,
+  x = ~Var1,
+  y = ~Var2,
+  z = ~value,
+  type = 'heatmap',
+  colorscale = 'Portland'
+) %>%
+  layout(
+    title = 'Correlation Matrix',
+    xaxis = list(title = 'Variable 1', tickangle = 45),
+    yaxis = list(title = 'Variable 2', autorange = "reversed")
+  )
+
+```

### 2024-04-21 R/processing-code/MADAproject_pt2_ruiz.qmd
Part 5
Analysis is complete. Readme files and removal of outdated datasets and codes will be removed throughout the week. Additionally, the manuscript will be finalized this weel

@@
 ## This file contains the process for loading, cleaning and exploring the datasets.
+### There is also a little visualizing data
 
@@
 
-# Correctly expand "Both" in Test Type, then adjust to "EEE" and "WNV"
-pcr_data_expanded <- pcr_data %>%
-  mutate(Test_Type_Expanded = case_when(
-    `Test Type` == "Both" ~ "EEE&WNV", # Temporarily mark "Both" for special handling
-    TRUE ~ `Test Type`
-  )) %>%
-  separate_rows(Test_Type_Expanded, sep = "&") %>%
-  mutate(
-    Year = year(`Collection Date`),
-    Week = week(`Collection Date`),
-    Submitted_for_Testing = `Submitted for Testing` == "Y",
-    # Assuming positive results are marked as "Positive" in `Result`
-    Is_Positive = Result == "Positive"
-  )
-
 # Save as RDS
-saveRDS(pcr_data_expanded, here("data", "processed-data", "rds", "pcr_data_expanded.rds"))
+saveRDS(pcr_data_expanded, here("data", "processed-data", "rds", "pcr_data"))
 
 # Save as CSV
-write_csv(pcr_data_expanded, here("data", "processed-data", "pcr_data_expanded.csv"))
+write_csv(pcr_data_expanded, here("data", "processed-data", "pcr_data.csv"))
 
@@
 ### The data includes all the positive PCR results for EEE and WNV for all mosquitoes submitted for testing
-
 ### These files were obtained from: https://www.mass.gov/lists/arbovirus-surveillance-plan-and-historical-data
-
 ### The virus isolation data is stored in separate spreadsheets by year from 2014 to 2020
-
 ### This code will load all the files, combine them, and then save the combined file in the "raw data" folder
-
 ```{r}
@@
 hourly_wx <- read_csv(hourly_wx_path, show_col_types = FALSE)
-
+View(hourly_wx)
 hourly_wx_clean <- hourly_wx %>%
@@
 ```{r}
-# read file with drought data by county for MA, RI, CT for the years 2007-2023
-drought_ma_ri_ct = read_csv(here("data", "raw-data", "comp_stat_county_ma_ri_ct.csv"), show_col_types = FALSE)
-
-# read file with drought data by county for FEMA Region 4 for the years 2007-2023
-drought_fema4 = read_csv(here("data", "raw-data", "comp_stat_fema4.csv"), show_col_types = FALSE)
-
+# read file with cumulative drought data by county for all MA counties for the years 2006-2023
+drought_ma_counties = read_csv(here("data", "raw-data", "comp_stat_ma_counties.csv"), show_col_types = FALSE)
+
+# read file with drought data by county for FEMA Region 1-4 for the years 2007-2023
+drought_fema1_4 = read_csv(here("data", "raw-data", "comp_stat_fema1_4.csv"), show_col_types = FALSE)
+#View(drought_fema1_4)
 # Add week number based on MapDate for both files
@@
 # the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08
-drought_ma_ri_ct <- drought_ma_ri_ct %>%
+
+# Additionally, this code calculates the Drought Severity and Coverage Index (DSCI) for each dataset
+# The DSCI is calculated as the sum of D0, D1, D2, D3, and D4 categories
+# This index provides a measure of the overall drought severity and can range from 0-500,
+# where 0 is no drought and 500 is the most severe drought over the entire area
+
+drought_ma_counties <- drought_ma_counties %>%
   mutate(
-    # First, ensure MapDate is in a proper date format
+    # Convert MapDate from a numeric format to a Date object
     MapDate = as.Date(as.character(MapDate), format = "%Y%m%d"),
-    # Then, format the collection date into "YYYY-WW"
-    week_num = paste0(year(MapDate), "-", 
-                      sprintf("%02d", isoweek(MapDate)))
-  )
-
-drought_fema4 <- drought_fema4 %>%
+    # Create a week_num column formatted as "YYYY-WW"
+    week_num = paste0(year(MapDate), "-", sprintf("%02d", isoweek(MapDate))),
+    # Calculate the DSCI, assuming D0 to D4 are numeric
+    DSCI_ma = rowSums(select(., D0:D4), na.rm = TRUE)
+  )
+#View(drought_ma_counties)
+
+drought_fema1_4 <- drought_fema1_4 %>%
  mutate(
-    # First, ensure MapDate is in a proper date format
+    # Convert MapDate from a numeric format to a Date object
     MapDate = as.Date(as.character(MapDate), format = "%Y%m%d"),
-    # Then, format the collection date into "YYYY-WW"
-    week_num = paste0(year(MapDate), "-", 
-                      sprintf("%02d", isoweek(MapDate)))
+    # Create a week_num column formatted as "YYYY-WW"
+    week_num = paste0(year(MapDate), "-", sprintf("%02d", isoweek(MapDate))),
+    # Calculate the DSCI, assuming D0 to D4 are numeric
+        DSCI_fema = rowSums(select(., c("D0", "D1", "D2", "D3", "D4")))
   )
@@
 # Define the path for the RDS and CSV files
-rds_path_drought_ma_ri_ct<- here("data", "processed-data", "rds", "drought_ma_ri_ct.rds")
-csv_path_drought_ma_ri_ct <- here("data", "processed-data", "drought_ma_ri_ct.csv")
+rds_path_drought_ma_counties<- here("data", "processed-data", "rds", "drought_ma_counties.rds")
+csv_path_drought_ma_counties <- here("data", "processed-data", "drought_ma_counties.csv")
+# Save the dataframe as an RDS and CSV files
+saveRDS(drought_ma_counties, rds_path_drought_ma_counties)
+write.csv(drought_ma_counties, csv_path_drought_ma_counties, row.names = FALSE)
+
+# Define the path for the RDS and CSV files
+rds_path_drought_fema1_4<- here("data", "processed-data", "rds", "drought_fema1_4.rds")
+csv_path_drought_fema1_4 <- here("data", "processed-data", "drought_fema1_4.csv")
 # Save the wx_selected dataframe as an RDS and CSV files
-saveRDS(drought_ma_ri_ct, rds_path_drought_ma_ri_ct)
-write.csv(drought_ma_ri_ct, csv_path_drought_ma_ri_ct, row.names = FALSE)
-
-# Define the path for the RDS and CSV files
-rds_path_drought_fema4<- here("data", "processed-data", "rds", "drought_fema4.rds")
-csv_path_drought_fema4 <- here("data", "processed-data", "drought_fema4.csv")
-# Save the wx_selected dataframe as an RDS and CSV files
-saveRDS(drought_fema4, rds_path_drought_fema4)
-write.csv(drought_fema4, csv_path_drought_fema4, row.names = FALSE)
+saveRDS(drought_fema1_4, rds_path_drought_fema1_4)
+write.csv(drought_fema1_4, csv_path_drought_fema1_4, row.names = FALSE)
 ```
@@
 
+## Bird data
+#### This dataset comes from the annual North American Breeding Bird Survey that is conducted in a joint effort by: USGS Eastern Ecological Science Center, Environment and Climate Change Canada, and the Mexican National Commission for the Knowledge and Use of Biodiversity.
+#### There are 2 (5 and 11) routes in Bristol County, MA. The data is from 2007-2022. However, route 11 is mission data for 2013 and route 5 ends in 2019.
+#### For route 11, 2013, we will use the data from 2012 and 2014 to estimate the missing data.
+for route 5, 2020-2022, we will use the data from 2019 to estimate the missing data.
+```{r}
+# Load the dataset
+file_path_bird <- here("data", "raw-data", "Massach.csv")
+bird_data_raw <- read.csv(file_path_bird)
+
+# Filter and clean up the data
+bird_bbs <- bird_data_raw %>%
+  filter(Route %in% c(5, 11), Year >= 2006 & Year <= 2022) %>%
+  select(-AOU, -Count10, -Count20, -Count30, -Count40, -Count50)
+
+# Summarize the data
+bird_bbs_summarized <- bird_bbs %>%
+  group_by(Route, Year) %>%
+  summarize(
+    StopTotal = sum(StopTotal, na.rm = TRUE),
+    SpeciesTotal = sum(SpeciesTotal, na.rm = TRUE),
+    .groups = 'drop'
+  )
+
+# Check for missing data within Routes 5 and 11
+missing_data <- map_df(c(5, 11), ~tibble(
+  Route = .x,
+  missing_years = setdiff(2006:2022, bird_bbs_summarized$Year[bird_bbs_summarized$Route == .x])
+))
+
+# Identify all combinations of Route and Year within the range
+all_combinations <- expand.grid(Route = unique(bird_bbs_summarized$Route),
+                                Year = 2006:2022)
+
+# Join with the summarized data to identify missing rows
+bird_bbs_full <- left_join(all_combinations, bird_bbs_summarized, by = c("Route", "Year"))
+
+# Impute missing data by taking the average of the previous and next year for each route
+bird_bbs_imputed <- bird_bbs_full %>%
+  group_by(Route) %>%
+  arrange(Route, Year) %>%
+  mutate(
+    StopTotal = ifelse(is.na(StopTotal), (lag(StopTotal) + lead(StopTotal)) / 2, StopTotal),
+    SpeciesTotal = ifelse(is.na(SpeciesTotal), (lag(SpeciesTotal) + lead(SpeciesTotal)) / 2, SpeciesTotal)
+  ) %>%
+  ungroup()
+
+# Handling the last records for Route 5 (2020-2022)
+# Since there's no future data to average, you might replicate the most recent data (e.g., from 2019)
+# Or consider another imputation method suitable for your analysis and data characteristics
+
+# Find the most recent year's data for Route 5
+recent_data_route_5 <- bird_bbs_imputed %>% 
+  filter(Route == 5, Year == max(Year[Route == 5 & !is.na(StopTotal)]))
+
+# Create a template for the missing years
+missing_years_route_5 <- tibble(Route = 5, Year = 2020:2022)
+
+# Replicate the most recent data for the missing years
+imputed_missing_years_route_5 <- bind_rows(lapply(missing_years_route_5$Year, function(y) {
+  mutate(recent_data_route_5, Year = y)
+})) %>%
+  bind_rows(bird_bbs_imputed) %>%
+  arrange(Route, Year)
+
+# Check if any adjustments or specific imputations are needed
+print(imputed_missing_years_route_5)
+
+bird_clean <- imputed_missing_years_route_5 %>%
+  drop_na()
+
+# Save the cleaned bird data as an csv
+write_csv(bird_clean, here("data", "processed-data", "bird_clean.csv"))
+```
+
 # Processing and cleaning the Mosquito trap data
-
 ### Examine the variables and change all column names to lower case
-
 ```{r}
@@
 ## we will examine the dataset before making changes.
-
 ### first we will Identify missing data
-
 ```{r}
@@
 # Step 4: Add week_num column
-wide_data <- wide_data %>%
+mosquito_wide <- wide_data %>%
   mutate(
@@
 # Output the transformed data to inspect its structure
-head(wide_data)
+head(mosquito_wide)
 
@@
 # Now we will filter the the mosquito data to match the time period of the virus data. 
-# Since mosquito populations can be affected by the previous year's populations, we will filter for 2011-
+# Since mosquito populations can be affected by the previous year's populations, we will filter for 2007-2023
 
@@
 write_csv(filtered_mosquito_data, file_path_filtered_csv)
-
+View(mosquito_wide_filtered.csv)
 # Define the path for the RDS file
@@
 # Define path to the data file
-trap_wide_path <- here("data", "processed-data", "mosquito_wide.csv")
+trap_wide_path <- here("data", "processed-data", "mosquito_wide_all.csv")
 
@@
 
-# Calculate the proportion of MEL out of the total for all species
+# Assuming all species columns are next to each other, calculate the proportion of MEL
 proportion_mel_by_year <- trap_proportion %>%
+  # Calculate row-wise totals for all species
+  rowwise() %>%
+  mutate(Total_All = sum(c_across(ABS:OCH), na.rm = TRUE)) %>%
+  ungroup() %>%
+  # Then group by year and summarize
   group_by(Year) %>%
   summarise(Total_MEL = sum(MEL, na.rm = TRUE),
-            Total_All = sum(MEL + PER + PIP + RES + SAL + TRI + SAP, na.rm = TRUE),
-            MEL_Prop = Total_MEL / Total_All)
+            Total_All = sum(Total_All, na.rm = TRUE), # Sum of all species by year
+            MEL_Prop = Total_MEL / Total_All) # Proportion of MEL
+
+# Print the results
+print(proportion_mel_by_year)
+
+# Calculate the proportion of MEL by week
+# Assuming all species columns are next to each other, calculate the proportion of MEL by week_num
+proportion_mel_by_week <- trap_proportion %>%
+  # Calculate row-wise totals for all species
+  rowwise() %>%
+  mutate(Total_All = sum(c_across(ABS:OCH), na.rm = TRUE)) %>%
+  ungroup() %>%
+  # Then group by week_num and summarize
+  group_by(week_num) %>%
+  summarise(Total_MEL = sum(MEL, na.rm = TRUE),
+            Total_All = sum(Total_All, na.rm = TRUE), # Sum of all species by week_num
+            MEL_Prop = Total_MEL / Total_All) # Proportion of MEL
+
+# Print the results
+print(proportion_mel_by_week)
+
+# save the data as CSV
+write_csv(proportion_mel_by_week, here("data", "processed-data", "mel_proportion_by_week.csv"))
 
@@
 
-### This will load sunset and sunrise times.
+### This will load sunset and sunrise times from the suncalc package. However, it can be slow to complete.
 #### the amount of sunlight affects mosquitoes activities and provides cues for them to enter diapause.
@@
 The data will be saved as a CSV file and an RDS file for future use.
+
 ```{r}
@@
 
-#Average daylight by week_num
+#here
+# Add 'week_num' calculation here
 sun_times_df <- sun_times_df %>%
   mutate(
-    # Combine Date with Sunrise and Sunset times to create full POSIXct datetime objects
+    # Convert Date column to Date object if it's not already
+    Date = as.Date(Date),
     Sunrise_dt = as.POSIXct(paste(Date, Sunrise), format="%Y-%m-%d %H:%M:%S", tz="UTC"),
     Sunset_dt = as.POSIXct(paste(Date, Sunset), format="%Y-%m-%d %H:%M:%S", tz="UTC"),
-    # Now calculate the difference in hours between Sunset and Sunrise
-    daylight_hours = as.numeric(difftime(Sunset_dt, Sunrise_dt, units = "hours"))
-  )
-
-# Aggregate data by week and calculate average daylight hours
+    # Calculate daylight hours
+    daylight_hours = as.numeric(difftime(Sunset_dt, Sunrise_dt, units = "hours")),
+    # Calculate week number as "YYYY-WW"
+    week_num = paste(year(Date), sprintf("%02d", isoweek(Date)), sep = "-")
+  )
+
+# Now, proceed to aggregate by 'week_num'
 average_daylight_by_week <- sun_times_df %>%
   group_by(week_num) %>%
-  summarise(average_daylight_hours = mean(daylight_hours))
-
-# Print first few rows of the resulting data frame
+  summarise(average_daylight_hours = mean(daylight_hours, na.rm = TRUE))  # Ensure NA values are handled
+
+# Optionally, save or print your aggregated data
+write.csv(average_daylight_by_week, here("data", "processed-data", "average_daylight_by_week.csv"), row.names = FALSE)
+# Print the first few rows to verify
 head(average_daylight_by_week)

### 2024-04-21 products/manuscript/Manuscript.qmd
Part 5
Analysis is complete. Readme files and removal of outdated datasets and codes will be removed throughout the week. Additionally, the manuscript will be finalized this weel

@@
 One indicator of the level of virus activity present in the mosquito
-population can be calculated from the information contained in dataset.
-The minimum infection rate (MIR) can be calculated at various temporal
-and spatial scales from trap site over one day to all sites over an
-entire season. MIR is calculated as: $$
+population can be calculated from the information contained in the
+mosquito dataset. The minimum infection rate (MIR) can be calculated at
+various temporal and spatial scales from trap site over one day to all
+sites over an entire season. MIR is calculated as: $$
 MIR = \left( \frac{\text{Number of positive mosquito pools}}{\text{Total number of mosquitoes tested}} \right) \times 1000
@@
 
-*weather data is currently being reprocessed because I have decided to
-switch from daily summaries to hourly summaries. new analysis will be
-added in version 5* Weather data played a significant role in this phase
-of the analysis. Hourly weather summaries were downloaded from NOAA's
-Climate Data Portal. The dataset covers the years 2007-2023 at the
-Taunton Airport in Bristol County.
+Weather plays a significant role in the lifecyle of the mosquito and
+influences the rate at which a pathogen can develop in a mosquito.
+Hourly and weather summaries were downloaded from NOAA's Climate Data
+Portal. The datasets cover the years 2007-2023 at the Taunton Airport in
+Bristol County.
 
@@
 the time between when a mosquito ingests a virus and when it can
-transmit it [@guStatisticalEstimationDegree2006].
+transmit it [@guStatisticalEstimationDegree2006]. To simplify this
+variable, the number of hours the temperature was ≥50°F and ≤86°F. These
+thresholds were chosen based on the known temperature range for mosquito
+development and virus transmission.
 
 In addition to the weather summaries, drought data will be included in
-the models. Drought data was obtained from the US Drought Monitor and
-the Federal Emergency Management Agency (FEMA). The data includes weekly
-drought severity ratings for the entire United States. The data is
-available at the county level and will be used to calculate the number
-of weeks in drought conditions and the severity of the drought. For this
-project, drought data for Bristol County, MA, will be used and
-incorporated with regional drought indicators.
+the models. West Nile Virus (WNV) research suggests that drought has
+been a driver in WNV outbreaks. While the lack of water can lead to
+smaller mosquito populations, drought can lead to higher prevelance of
+mosquito infection [@paullDroughtImmunityDetermine2017]. Drought data
+was obtained from the US Drought Monitor and the Federal Emergency
+Management Agency (FEMA). The data includes weekly drought severity
+ratings for the entire United States. The data is available at county
+and regional levels and will be used to calculate the number of weeks in
+drought conditions and the severity of the drought. For this project,
+drought data for Bristol County, MA, will be used and incorporated with
+FEMA East Coast regions (I-IV) drought indicators.
 
@@
 
-*This section is under development and will be completed after analysis
-is complete. The following is a general outline of the methods.*
-
 Symptomatic EEE, and to a lesser extent WNV, human cases are rare. On
@@
 [@centersfordiseasecontrolandpreventionPreventionEasternEquine2022]. The
-low incidence of human cases poses significant challenges for
+low incidence of human cases poses significant challenges for using
 traditional statistical models, making it difficult to accurately
@@
 
-As stated earlier, MIR will be the main outcome of interest for this
-project.
+MIR will be the main outcome of interest for this project. MIR values
+will be calculated for each week during the surveillance season over the
+entire county. To reduce bias, only mosquitoes captured using carbon
+dioxide-baited CDC light traps will be used to to calculate MIR. These
+traps are the standard EEE surveillance and are used to monitor the
+actively feeding mosquito population. Additionally, these trap types
+have been used in this area for over 40 years, many at the same sites
+currently in use.
+
+After MIR was calculated and joined with weather and climate data, the
+distribution of the MIR was examined more closely. About 72% of the
+records were zero values, which is not uncommon for this type of data.
+The non-zero values were skewed to the left and there was strong
+evidence of over-dispersion.
+
+![](images/mir_density_plot.png){fig-align="center"}
+
+The average MIR was 1.26, with an observed variance of 7.56, resulting
+in a Variance-to-Mean Ratio (VMR) of 6.00. This pronounced
+overdispersion suggests that mosquito infection rates were not uniformly
+distributed over time or space. Instead, they may have been influenced
+by a range of factors, such as environmental conditions or variations in
+amplifying host abundance.
+
+The data was visualized to identify trends and relationships between the
+variables. There were no clear patterns in the data between MIR and the
+potential predictors. However, several varibles showed a non-linear
+relationship with MIR. For example, the relationship between MIR and
+average temperature was not linear. Instead, MIR increased with
+temperature up to a certain point, then decreased. This suggests that
+the relationship between temperature and MIR may be more complex than a
+simple linear model can capture.
+
+In response to the observed non-linear relationships in the data, a
+two-stage Generalized Additive Model (GAM) analysis was used. GAMs
+provide a versatile framework for capturing complex, non-linear
+relationships between response variables and multiple predictors, which
+are often encountered in ecological data sets.
+
+In the first stage of the analysis, the GAM was utilized to address the
+high incidence of zeros in the dataset—an indicator of overdispersion.
+This zero-inflated model allowed us to separately model the presence
+versus absence of the outcome variable, thereby appropriately accounting
+for the excessive zeros.
+
+Subsequently, the second stage of the GAM focused on modeling the
+continuous part of the MIR data, relating it to the suite of potential
+predictors. This step incorporated smooth functions of the predictors,
+permitting the data to inform the shape of the relationship without
+imposing a predetermined form, as would be the case in a traditional
+linear model.
+
+While the two-stage GAM appeared to be a good fit, there were too few
+observations to provide a robust estimate of the relationship between
+the predictors and MIR. The original dataset contained over 40,000
+observations, but after reformatting the data to fit a rectangular shape
+needed for analysis, there were less than 400 observations in the final
+set. After the zeros were removed for part 2 of the GAM, the sample
+number of observations left was less than 100. Because of this, it was
+decided to use a single GAM with a dataset that included zero and
+non-zero observations. Additionally, all trap types were included in the
+analysis, not just the carbon dioxide-baited CDC light traps. While some
+precision was lost when the one part model was adopted, the gain in
+observations was deemed more valuable.
 
@@
 amplification in the enzootic cycle. Understanding the population
-dynamics might provide more insight into what is driving the 2019 spike.
+dynamics might provide more insight into what is driving the spikes,
+like in 2019.
 
@@
 and total precipitation by week). The outcome was whether there was at
-least 1 positive pool in a week. The predictor in the first model was
-the average maximum daily temperature the week before the pool week. The
-logistic model suggest that there was a relationship between the two.
-(The R code for the model can be found in the EDA-Code file:
-logistic\_.pos_temp_precip.R) The average temperature in the week before
-is positively associated with the likelihood of a positive result, with
-an odds ratio of approximately 1.069 for each unit increase in
-temperature (Estimate = 0.06695, SE = 0.02033, p \< 0.001), indicating
-that higher temperatures significantly increase the probability of a
-positive outcome. The model's improvement is highlighted by a decrease
-in residual deviance to 405.79 from the null deviance of 417.24, upon
-incorporating Lagged_Avg_TMAX as a predictor, demonstrating a better fit
-to the observed data. However, with a residual deviance of 405.79 on 299
-degrees of freedom, it is evident that not all variability in the
-outcome is explained, suggesting the presence of other influential
-factors. The model's AIC of 409.79 indicates a balance between model
-complexity and predictive power, supporting its relative utility in
-understanding the data. However, this relationship should be examined
-more closely as temperature is likely auto correlated with progressing
-weeks during summer months. While precipitation has been demonstrated to
-influence the population dynamics it did not show to have any effect on
-a positive result.
-
-Another model assessed whether the temperature precipitation in April,
-May, and June of each year effected the date of the first positive pool
-test. The date of the first positive pool was identified and assigned a
-day number from January 1st of that year. The predictors in this linear
-model were the average maximum temperature for April, May, and June and
-the sum of precipitation for the same time period. None of these
-predictors were significant.
+least 1 positive test for pool in a given week. The predictor in the
+first model was the average maximum daily temperature the week before
+the mosquito collection week. The logistic model suggests that there was
+a relationship between the two. (The R code for the model can be found
+in the analysis-code file: analysis.qmd) The average temperature in the
+week before is positively associated with the likelihood of a positive
+result, with an odds ratio of approximately 1.069 for each unit increase
+in temperature (Estimate = 0.06695, SE = 0.02033, p \< 0.001),
+indicating that higher temperatures significantly increase the
+probability of a positive outcome. The model's improvement is
+highlighted by a decrease in residual deviance to 405.79 from the null
+deviance of 417.24, upon incorporating Lagged_Avg_TMAX as a predictor,
+demonstrating a better fit to the observed data. However, with a
+residual deviance of 405.79 on 299 degrees of freedom, it is evident
+that not all variability in the outcome is explained, suggesting the
+presence of other influential factors. The model's AIC of 411.59
+indicates a balance between model complexity and predictive power,
+supporting its relative utility in understanding the data. However, this
+relationship should be examined more closely as temperature is likely
+auto correlated with progressing weeks during summer months. While
+precipitation has been demonstrated to influence the population dynamics
+it did not show to have any effect on a positive result. This was
+initailly surprising, but upon after considering the specific habitat of
+the main EEE enzootic vector, *Culiseta melanura*, it makes sense. MEL
+larvae develop in well insulated crypts that are less sensitive
+evaporative processes than surface water. The effect of precipitation on
+MEL larval habitat abundance may be less pronounced than other vector
+species that develop in open, standing water, like ditches and
+containers. Another model assessed whether the temperature precipitation
+in April, May, and June of each year effected the date of the first
+positive pool test (R file: PT3_edacode). The date of the first positive
+pool was identified and assigned a day number from January 1st of that
+year. The predictors in this linear model were the average maximum
+temperature for April, May, and June and the sum of precipitation for
+the same time period. None of these predictors were significant.
 
@@
 dependent, continuous variable in a linear regression and the average
-maximum temperature for the previous week is used as the predictor, the model suggests there
-is a significant inverse relationship. As average temperature increases,
-the MIR decreases (model_mir_temp.R). This relationship is
-counterintutive, especially given the result of the logistic regression
-and requires more examination.
-
-More models assessing the effects of weather and other factors on
-positive pools by mosquito species and MIR values will be developed in
-Part 4 of the assignment. Future models will examine the the effects of
-weather conditions on mosquito population dynamics.
+maximum temperature for the previous week is used as the predictor, the
+model suggests there is a significant inverse relationship. As average
+temperature increases, the MIR decreases (model_mir_temp.R). This
+relationship is counterintutive, especially given the result of the
+logistic regression and requires more examination.
 
 ## Statistical analysis
+Before any models were assessed, the RSME for the null model was calculated (2.40). This was used as the basis for assessing model improvement.
 
-*UPDATE for Pt 4* *a better source for weather data was found and it
-being processed.* *the analysis will be updated in the next version of
-the document*
+Analysis began with a Generalized Additive Model (GAM) with predictors that were selected based on the exploratory single variable GAM.
+The variables were chosen if their p=values were <0.05.
+Eight variables met that condition including: the number of hours below 50°F the week before the mosquito collection week (Hrs_Below_50_lag), the number of hours below 50°F the week of collection (Hrs_Below_50_current), the drought severity index for East Coast FEMA regions (DSCI_fema), the average hourly relative humidity the week of collection (Avg_HourlyRelativeHumidity_current), the average hourly windspeed the week of collection (Avg_HourlyWindSpeed_current), the average number of daylight hours per day (average_daylight_hours), the average number of MEL captured by trap (Average_MEL), and the proportion of MEL captured out of the total number of mosquitoes trapped (MEL_Prop).
 
-![](images/correlation_plot.png){fig-align="center"}
+All models included training (.75) and a testing (.25) dataset that were generated using the random seed 123.
+All analysis can be found in the analysis.qmd file.
+The first model was a GAM with all eight predictors, followed by a Random Forest, LASSO, Generalized Additive Model with Sparsity (GAMS), and a Gradient Boosting Machine (GBM) model.
 
-# Results
+As expected, the RSME is lower for the training dataset than the testing dataset for all models. The RMSE values are displayed in the table below.
 
-*updated in later phases of the project*
+```{r}
+# Define the file path using here()
+file_path_rds <- here("results", "tables", "rmse_table.rds")
 
-## Full analysis
-
-*These will be generated in Pt 5* 
+# Read the RDS file
+rmse_table <- readRDS(file_path_rds)
+rmse_table
+```
 # Discussion
-*this section will be
-completed when analysis is done*
+The null model RMSE was 2.40.
+All models showed improvement over the null model on both the training and testing datasets.
+The GAM shows significant improvement over the null model for the training set, but only slightly improves over the null for the testing set.
+This suggests that beyond capturing the mean response, this GAM is not well generalized.
+The substantial improvement in the training set RSME for GBM and Random Forest coupled with less improvement over the null for the testing set, suggest a tendency for overfitting.
+These models may be capturing the training data's patterns that do not gerneralize well to unseen data.
+The the gap between the training and testing RMSE for the LASSO and GAMS models suggest there is less overfitting and are slightly better at generalizing than the others.
+However, given the non-linear relationships, the LASSO model is likely not appropriate for this analysis.
 
@@
 
-*this section will be updated after analysis is complete* The accuracy
+The accuracy
 and reliability of the mosquito collection data is excellent. The same
@@
 sorting, identifying and counting species over the entire period.
+However, the data is not without limitations.
+Mosquito surveillance is inherently biased.
+Traps are set in locations that are most likely to capture infected mosquitoes.
+While this is a positive for early warning of virus activity, it is not necessarily representative of the surrounding area.
+Probably the biggest limitation is the small number of observations.
+The mosquito season is short in Massachusetts and funding for trapping and testing mosquitoes is limited.
+Mosquito control efforts, like pesticide applications, are missing from this analysis.
+These efforts are designed to lower the levels of virus transmitting mosquitoes and would likely provide more insight into the relationship between the predictors and virus activity.
 
@@
 option in later phases of this project.
+Ideally, the outcome would be human or animal cases.
+However, these cases are rare and modeling them given the limited amount of surveillance data may not be possible.
+
+Initially, the lack of significance for most temperature and precipitation variable was surprising.
+However, upon closer examination, it makes sense.
+The main enzootic vector of EEE and the one targeted for surveillance, *Culiseta melanura*, develops in well insulated crypts that are less sensitive to ambient air temperature changes.
+The effect of temperature changes on MEL development speed may be less pronounced than other vector species that develop in open, standing water, like ditches and containers.
+Precipitation increases the amount of standing water and increases the chances of larval survival during the mosquito season and also during the winter months for those species that overwinter in the larval stage.
+For EEE, the main enzootic vector, MEL, overwinters as larvae.
 
@@
 
-## Conclusions *this section will be completed when analysis is done*
+Likely the biggest limitation 
+## Conclusions 
+While all models demonstrated improvement over null model, none of them were robust enough to be used for reliable prediction.
+In the next phases of this project, the data will be expanded to include more years and cover a larger portion of Massachusetts.
+In additon to increasing the the number of observations, other sources of data will be incorporated.
+Including virus activity from surrounding regions and areas from which birds migrate to Massachusetts.
 

### 2024-04-21 R/analysis-code/archived files/2partGAMmodel.R

@@ -1,0 +1,114 @@
+library()
+###Two Part Model###
+mir_2part <- read_csv(here("data", "processed-data", "joined_data_mir_co2.csv"))
+
+library(dplyr)
+str(mir_2part)
+# Creating a binary variable for zero vs. positive MIR
+mir_2part <- mir_2part %>%
+  mutate(MIR_Positive = as.integer(MIR > 0))
+
+# Fitting the logistic regression model
+logistic_model <- glm(MIR_Positive ~ Avg_HourlyRelativeHumidity_current + Avg_HourlyWindSpeed_current + Hrs_Below_50_current,
+                      family = binomial(link = "logit"),
+                      data = mir_2part)
+
+summary(logistic_model)
+
+
+# Filtering the dataset for positive MIR values
+positive_mir_data <- mir_2part %>% 
+  filter(MIR > 0)
+
+# Fitting the linear regression model to the positive MIR values
+linear_model <- lm(MIR ~ Avg_HourlyRelativeHumidity + Avg_TMAX + DSCI_ma + Avg_TMAX,
+                   data = positive_mir_data)
+
+summary(linear_model)
+
+# Plot the diagnostic plots for the linear model
+par(mfrow = c(2, 2))
+plot(linear_model)
+### These suggest that the issues of non-linearity, heteroscedasticity, and influential outliers need to be addressed using different models
+
+# For this we will use a two-part model, which consists of a GAM logistic regression model to predict the probability of observing a positive MIR value,
+
+library(mgcv)
+# Fit a GAM with a smoothing function for each predictor
+gam_model <- gam(MIR ~ s(Avg_TMAX),
+                 data = mir_2part, 
+                 family = gaussian(),
+                 method = "REML")
+
+# Check the summary for detailed output
+summary(gam_model)
+
+# Plot the smooths to interpret the effect of each predictor
+plot(gam_model)
+
+
+
+##2 stage GAM
+
+zero_gam_model <- gam(I(MIR > 0) ~ s(DSCI_ma) + s(Average_MEL) + s(Avg_HourlyWindSpeed_current),
+                      family = binomial(link = "logit"), 
+                      data = mir_2part)
+
+summary(zero_gam_model)
+plot(zero_gam_model)
+
+names(mir_2part)
+positive_gam_model <- gam(MIR ~ s(Hrs_Between_50_86_lag), 
+                          data = mir_2part[mir_2part$MIR > 0, ], 
+                          family = gaussian())
+# Check the summary for detailed output
[…]

### 2024-04-21 R/analysis-code/archived files/Annual_models.R

@@ -1,0 +1,2 @@
+####Annual Model####
+

### 2024-04-21 R/analysis-code/archived files/statistical-analysis.R

@@ -1,0 +1,39 @@
+library(here)
+library(tidymodels)
+library(ggplot2)
+library(GGally)
+library(ggcorrplot)
+library(glmnet)
+library(ranger)
+library(dplyr)
+
+# Load the datasets:
+# Joined_data_MIR_CO2: Contains information on mosquito infection rates (MIR) and weather data.
+joined_data_mir_co2 <- read_csv(here("data", "processed-data", "joined_data_mir_co2.csv"))
+
+# Assuming combined_data_mir_all is loaded and prepared
+continuous_vars_model <- joined_data_mir_co2[, c("MIR", "Average_MEL", "Total_PRCP", "Hours_Above_80.y", 
+                                                   "Hours_Above_80.x", "Hours_Below_50.x", "Hours_Below_50.y", 
+                                                   "Avg_HourlyRelativeHumidity.y", "dsci", "dsdi")]
+
+# Imputing NA values for Average_MEL with the mean (or choose another method as appropriate)
+continuous_vars_model$Average_MEL[is.na(continuous_vars_model$Average_MEL)] <- mean(continuous_vars_model$Average_MEL, na.rm = TRUE)
+
+# Removing dsdi from the analysis due to zero variance
+continuous_vars_model <- continuous_vars_model[, !names(continuous_vars_model) %in% c("dsdi")]
+
+# Recompute the correlation matrix without dsdi and with imputed values for Average_MEL
+cor_matrix <- cor(continuous_vars_model, use = "complete.obs")
+
+# Visualize the updated correlation matrix
+cor_mir_plot = ggcorrplot(cor_matrix, method = "circle", hc.order = TRUE, type = "lower",
+           lab = TRUE, lab_size = 3, colors = c("gold", "snow", "tomato"))
+
+# Print the updated correlation matrix
+print(cor_matrix)
+# Save the plot as a PNG file
+ggsave(filename = here("results", "figures", "correlation_plot.png"), plot = cor_mir_plot, width = 10, height = 8, dpi = 300)
+
+# Or save as a PDF for high-quality printing
+ggsave(filename = here("results", "figures", "correlation_plot.pdf"), plot = cor_mir_plot, width = 10, height = 8)
+

### 2024-04-21 R/analysis-code/statistical-analysis.R

@@ -1,39 +1,0 @@
-library(here)
-library(tidymodels)
-library(ggplot2)
-library(GGally)
-library(ggcorrplot)
-library(glmnet)
-library(ranger)
-library(dplyr)
-
-# Load the datasets:
-# Joined_data_MIR_CO2: Contains information on mosquito infection rates (MIR) and weather data.
-joined_data_mir_co2 <- read_csv(here("data", "processed-data", "joined_data_mir_co2.csv"))
-
-# Assuming combined_data_mir_all is loaded and prepared
-continuous_vars_model <- joined_data_mir_co2[, c("MIR", "Average_MEL", "Total_PRCP", "Hours_Above_80.y", 
-                                                   "Hours_Above_80.x", "Hours_Below_50.x", "Hours_Below_50.y", 
-                                                   "Avg_HourlyRelativeHumidity.y", "dsci", "dsdi")]
-
-# Imputing NA values for Average_MEL with the mean (or choose another method as appropriate)
-continuous_vars_model$Average_MEL[is.na(continuous_vars_model$Average_MEL)] <- mean(continuous_vars_model$Average_MEL, na.rm = TRUE)
-
-# Removing dsdi from the analysis due to zero variance
-continuous_vars_model <- continuous_vars_model[, !names(continuous_vars_model) %in% c("dsdi")]
-
-# Recompute the correlation matrix without dsdi and with imputed values for Average_MEL
-cor_matrix <- cor(continuous_vars_model, use = "complete.obs")
-
-# Visualize the updated correlation matrix
-cor_mir_plot = ggcorrplot(cor_matrix, method = "circle", hc.order = TRUE, type = "lower",
-           lab = TRUE, lab_size = 3, colors = c("gold", "snow", "tomato"))
-
-# Print the updated correlation matrix
-print(cor_matrix)
-# Save the plot as a PNG file
-ggsave(filename = here("results", "figures", "correlation_plot.png"), plot = cor_mir_plot, width = 10, height = 8, dpi = 300)
-
-# Or save as a PDF for high-quality printing
-ggsave(filename = here("results", "figures", "correlation_plot.pdf"), plot = cor_mir_plot, width = 10, height = 8)
-

### 2024-04-21 R/eda-code/MIR.R

@@ -1,202 +1,0 @@
-library(readxl)
-library(dplyr)
-library(lubridate)
-library(tidyr) # For separate_rows()
-library(here)
-
-# Read the dataset
-pcr_data <- read_excel(here("data", "raw-data", "virus_iso.xlsx"))
-
-# Add week number based on the collection data
-# This code ensures that the ISO 8601 standard is applied to the week number calculation
-# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
-# Weeks start on Monday and end on Sunday.
-# The last week of the year, week 52 or 53, is the one that contains December 28th.
-# we will use week_num to join tables
-# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08
-pcr_data <- pcr_data %>%
-  mutate(
-    # Directly format the collection date into "YYYY-WW"
-    week_num = paste0(year(`Collection Date`), "-", 
-                      sprintf("%02d", isoweek(`Collection Date`)))
-  )
-
-
-# Correctly expand "Both" in Test Type, then adjust to "EEE" and "WNV"
-pcr_data_expanded <- pcr_data %>%
-  mutate(Test_Type_Expanded = case_when(
-    `Test Type` == "Both" ~ "EEE&WNV", # Temporarily mark "Both" for special handling
-    TRUE ~ `Test Type`
-  )) %>%
-  separate_rows(Test_Type_Expanded, sep = "&") %>%
-  mutate(
-    Year = year(`Collection Date`),
-    Week = week(`Collection Date`),
-    Submitted_for_Testing = `Submitted for Testing` == "Y",
-    # Assuming positive results are marked as "Positive" in `Result`
-    Is_Positive = Result == "Positive"
-  )
-
-# Save as RDS
-saveRDS(pcr_data_expanded, here("data", "processed-data", "rds", "pcr_data_expanded.rds"))
-
-# Save as CSV
-write_csv(pcr_data_expanded, here("data", "processed-data", "pcr_data_expanded.csv"))
-
-# Filter records where Submitted for Testing is "Y"
-pcr_data_expanded_filtered <- pcr_data_expanded %>%
-  filter(Submitted_for_Testing)
-
-# Save as RDS
[…]

### 2024-04-21 R/eda-code/Model_mir_temp.R

@@ -1,175 +1,0 @@
-# Load necessary libraries for data manipulation, date handling, and file path management
-library(lubridate)
-library(here)
-library(dplyr)
-library(ggplot2)
-
-
-# Load the datasets:
-# MIR_by_week: Contains information on mosquito pool tests, including the week number and MIR.
-# wx_selected: Weather data with daily maximum temperature (TMAX) and precipitation (PRCP).
-# mosquito_wide: Another dataset related to mosquito data, although it's loaded, it's not used in the subsequent analysis.
-MIR_by_week_co2 <- read_csv(here("data", "processed-data", "MIR_by_week_co2.csv"))
-wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
-mosquito_wide <- read_csv(here("data", "processed-data", "mosquito_wide.csv"))
-drought_fema4 <- read_csv(here("data", "processed-data", "drought_fema4.csv"))
-drought_ma_ri_ct <- read_csv(here("data", "processed-data", "drought_ma_ri_ct.csv"))
-*average_moon_phase_by_week <- read_csv(here("data", "processed-data", "average_moon_phase_by_week.csv"))
-average_daylight_by_week <- read_csv(here("data", "processed-data", "average_daylight_by_week.csv"))
-average_mel_weekco2 <- read_csv(here("data", "processed-data", "average_mel_week_co2.csv"))
-wx_hourly_clean <- read_csv(here("data", "processed-data", "hourly_wx_clean.csv"))
-
-# Aggregate weather data by week:
-# Calculate the average maximum temperature (Avg_TMAX) and total precipitation (Total_PRCP) for each week.
-# This simplification allows for weekly comparisons with the MIR data.
-wx_aggregated <- wx_selected %>%
-  group_by(week_num) %>%
-  summarise(Avg_TMAX = mean(TMAX, na.rm = TRUE),
-            Total_PRCP = sum(PRCP, na.rm = TRUE),
-            .groups = 'drop')
-
-# Adjust for a one-week lag in weather data:
-# Create a new column (Lagged_week_num) that represents each week's preceding week. 
-# This adjustment accounts for the delayed effect weather might have on MIR.
-wx_aggregated <- wx_aggregated %>%
-  mutate(
-    # Split week_num to get year and week as numeric
[…]

### 2024-04-21 R/eda-code/archived_files/MIR.R

@@ -1,0 +1,202 @@
+library(readxl)
+library(dplyr)
+library(lubridate)
+library(tidyr) # For separate_rows()
+library(here)
+
+# Read the dataset
+pcr_data <- read_excel(here("data", "raw-data", "virus_iso.xlsx"))
+
+# Add week number based on the collection data
+# This code ensures that the ISO 8601 standard is applied to the week number calculation
+# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
+# Weeks start on Monday and end on Sunday.
+# The last week of the year, week 52 or 53, is the one that contains December 28th.
+# we will use week_num to join tables
+# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08
+pcr_data <- pcr_data %>%
+  mutate(
+    # Directly format the collection date into "YYYY-WW"
+    week_num = paste0(year(`Collection Date`), "-", 
+                      sprintf("%02d", isoweek(`Collection Date`)))
+  )
+
+
+# Correctly expand "Both" in Test Type, then adjust to "EEE" and "WNV"
+pcr_data_expanded <- pcr_data %>%
+  mutate(Test_Type_Expanded = case_when(
+    `Test Type` == "Both" ~ "EEE&WNV", # Temporarily mark "Both" for special handling
+    TRUE ~ `Test Type`
+  )) %>%
+  separate_rows(Test_Type_Expanded, sep = "&") %>%
+  mutate(
+    Year = year(`Collection Date`),
+    Week = week(`Collection Date`),
+    Submitted_for_Testing = `Submitted for Testing` == "Y",
+    # Assuming positive results are marked as "Positive" in `Result`
+    Is_Positive = Result == "Positive"
+  )
+
+# Save as RDS
+saveRDS(pcr_data_expanded, here("data", "processed-data", "rds", "pcr_data_expanded.rds"))
+
+# Save as CSV
+write_csv(pcr_data_expanded, here("data", "processed-data", "pcr_data_expanded.csv"))
+
+# Filter records where Submitted for Testing is "Y"
+pcr_data_expanded_filtered <- pcr_data_expanded %>%
+  filter(Submitted_for_Testing)
+
+# Save as RDS
[…]

### 2024-04-21 R/eda-code/archived_files/Model_mir_temp.R

@@ -1,0 +1,298 @@
+# Load necessary libraries for data manipulation, date handling, and file path management
+library(lubridate)
+library(here)
+library(dplyr)
+library(ggplot2)
+library(pscl)
+library(MASS)
+library(readr)
+library(purrr)
+
+
+# Load the datasets:
+MIR_by_week_co2 <- read_csv(here("data", "processed-data", "MIR_by_week_co2.csv"))
+wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
+mosquito_wide <- read_csv(here("data", "processed-data", "mosquito_wide_all.csv"))
+drought_fema1_4 <- read_csv(here("data", "processed-data", "drought_fema1_4.csv"))
+drought_ma_counties <- read_csv(here("data", "processed-data", "drought_ma_counties.csv"))
+#average_moon_phase_by_week <- read_csv(here("data", "processed-data", "average_moon_phase_by_week.csv"))
+average_daylight_by_week <- read_csv(here("data", "processed-data", "average_daylight_by_week.csv"))
+average_mel_weekco2 <- read_csv(here("data", "processed-data", "average_mel_weekco2.csv"))
+wx_hourly_clean <- read_csv(here("data", "processed-data", "hourly_wx_clean.csv"))
+bird_clean <- read_csv(here("data", "processed-data", "bird_clean.csv"))
+mel_prop_weekly <- read_csv(here("data", "processed-data", "mel_proportion_by_week.csv"))
+
+# Aggregate weather data by week:
+# Calculate the average maximum temperature (Avg_TMAX) and total precipitation (Total_PRCP) for each week.
+# This simplification allows for weekly comparisons with the MIR data.
+wx_aggregated <- wx_selected %>%
+  group_by(week_num) %>%
+  summarise(Avg_TMAX = mean(TMAX, na.rm = TRUE),
+            Total_PRCP = sum(PRCP, na.rm = TRUE),
+            .groups = 'drop')
+
+# Adjust for a one-week lag in weather data:
+# Create a new column (Lagged_week_num) that represents each week's preceding week. 
+# This adjustment accounts for the delayed effect weather might have on MIR.
+wx_aggregated <- wx_aggregated %>%
+  mutate(
[…]

### 2024-04-21 R/eda-code/archived_files/PT3_edacode.R

@@ -1,0 +1,85 @@
+## ---- packages --------
+#load needed packages. make sure they are installed.
+library(here) #for data loading/saving
+library(dplyr)
+library(skimr)
+library(ggplot2)
+library(tidyverse)
+library(lubridate)
+library(readr)
+library(forecast)
+library(tsibble) # For time series data wrangling
+library(feasts) # For decomposing and visualizing time series
+
+pcr <- read_csv(here("data", "processed-data", "pcr_data_expanded_filtered.csv"))
+wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
+
+# Step 2: Calculate the average TMAX for April, May, and June separately
+avg_temp_by_month <- wx_selected %>%
+  filter(month(DATE) %in% c(4, 5, 6)) %>%
+  group_by(Year = year(DATE), Month = month(DATE)) %>%
+  summarise(Avg_TMAX = mean(TMAX, na.rm = TRUE), .groups = 'drop')
+
+# Pivot the average temperatures into separate columns for each month
+avg_temp_by_month_wide <- avg_temp_by_month %>%
+  pivot_wider(names_from = Month, values_from = Avg_TMAX, names_prefix = "Avg_TMAX_M") %>%
+  # Ensure all years are present even if some data might be missing
+  complete(Year, fill = list(Avg_TMAX_M4 = NA, Avg_TMAX_M5 = NA, Avg_TMAX_M6 = NA))
+
+# Step 4: Identify the first positive test date each year in mosquito_data
+first_positive_dates <- pcr %>%
+  filter(Result == "Positive") %>%
+  group_by(Year = year(`Collection Date`)) %>%
+  summarise(First_Positive_Date = min(`Collection Date`), .groups = 'drop')
+
+# Step 5: Merge the datasets on Year
+combined_data_firstpos <- left_join(first_positive_dates, avg_temp_by_month_wide, by = "Year")
+
+# Step 6: Convert First_Positive_Date to a numerical value (e.g., day of the year)
+combined_data_firstpos <- combined_data_firstpos %>%
+  mutate(First_Positive_DOY = yday(First_Positive_Date))
+
+# Explore the relationship via linear regression with Avg_TMAX_M4, Avg_TMAX_M5, Avg_TMAX_M6 as predictors
[…]

### 2024-04-21 R/eda-code/archived_files/logistic_.pos_temp_precip.R

@@ -1,0 +1,71 @@
+library(here)
+library(dplyr)
+library(lubridate)
+library(readr)
+
+
+# Load the PCR data set which includes information on mosquito pools tested for viruses, 
+# results of those tests, and the date of collection.
+pcr <- read_csv(here("data", "processed-data", "pcr_data_expanded_filtered.csv"))
+
+# Load the weather data set which includes daily maximum temperatures (TMAX) and other weather variables.
+wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
+
+# Aggregate the weather data by week to calculate the average maximum temperature for each week.
+# This simplifies the daily temperature data into a more manageable weekly summary,
+# making it easier to correlate with weekly mosquito test results.
+wx_selected <- wx_selected %>%
+  group_by(week_num) %>%
+  summarise(
+    Avg_TMAX = mean(TMAX, na.rm = TRUE),
+    Total_PRCP = sum(PRCP, na.rm = TRUE)
+  ) %>%
+  ungroup()
+
+# View the first few rows to confirm it worked
+head(wx_selected)
+
+
+# Create a lagged version of the average weekly maximum temperature to explore the potential delayed effect of temperature on mosquito test results.
+# The lag function shifts the temperature data by one week, assuming that the impact of temperature on virus transmission might not be immediate.
+weather_weekly <- wx_selected %>%
+  arrange(week_num) %>%
+  mutate(Lagged_Avg_TMAX = lag(Avg_TMAX, 1))
+
+weather_weekly <- wx_selected %>%
+  arrange(week_num) %>%
+  mutate(
+    Lagged_Avg_TMAX = lag(Avg_TMAX, 1),
+    Lagged_Total_PRCP = lag(Total_PRCP, 1)
+  )
+
+weather_weekly <- wx_selected %>%
+  arrange(week_num) %>%
+  mutate(
+    Lagged_Avg_TMAX = lag(Avg_TMAX, 1),
+    Lagged_Total_PRCP = lag(Total_PRCP, 1),
+    Binary_PRCP = if_else(Lagged_Total_PRCP > 0, 1, 0)
+  )
+# Aggregate the PCR data by week to determine if there was at least one positive result in each week.
+# This transformation reduces the detailed test results into a weekly summary,
+# indicating the presence or absence of positive tests each week.
+pcr_weekly <- pcr %>%
+  group_by(week_num) %>%
[…]

### 2024-04-21 R/eda-code/logistic_.pos_temp_precip.R

@@ -1,67 +1,0 @@
-library(here)
-library(dplyr)
-library(lubridate)
-library(readr)
-
-
-# Load the PCR data set which includes information on mosquito pools tested for viruses, 
-# results of those tests, and the date of collection.
-pcr <- read_csv(here("data", "processed-data", "pcr_data_expanded_filtered.csv"))
-
-# Load the weather data set which includes daily maximum temperatures (TMAX) and other weather variables.
-wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
-
-# Aggregate the weather data by week to calculate the average maximum temperature for each week.
-# This simplifies the daily temperature data into a more manageable weekly summary,
-# making it easier to correlate with weekly mosquito test results.
-wx_selected <- wx_selected %>%
-  group_by(week_num) %>%
-  summarise(Avg_TMAX = mean(TMAX, na.rm = TRUE)) %>%
-  ungroup()
-wx_selected <- wx_selected %>%
-  group_by(week_num) %>%
-  summarise(
-    Avg_TMAX = mean(TMAX, na.rm = TRUE),
-    Total_PRCP = sum(PRCP, na.rm = TRUE)
-  ) %>%
-  ungroup()
-
-# Create a lagged version of the average weekly maximum temperature to explore the potential delayed effect of temperature on mosquito test results.
-# The lag function shifts the temperature data by one week, assuming that the impact of temperature on virus transmission might not be immediate.
-weather_weekly <- wx_selected %>%
-  arrange(week_num) %>%
-  mutate(Lagged_Avg_TMAX = lag(Avg_TMAX, 1))
-
-weather_weekly <- wx_selected %>%
-  arrange(week_num) %>%
-  mutate(
-    Lagged_Avg_TMAX = lag(Avg_TMAX, 1),
-    Lagged_Total_PRCP = lag(Total_PRCP, 1)
-  )
-
-weather_weekly <- wx_selected %>%
-  arrange(week_num) %>%
-  mutate(
-    Lagged_Avg_TMAX = lag(Avg_TMAX, 1),
-    Lagged_Total_PRCP = lag(Total_PRCP, 1),
-    Binary_PRCP = if_else(Lagged_Total_PRCP > 0, 1, 0)
-  )
-# Aggregate the PCR data by week to determine if there was at least one positive result in each week.
-# This transformation reduces the detailed test results into a weekly summary,
[…]

### 2024-04-24 project-review_EMollinedo.qmd

---
title: Project Review
author: Erick Mollinedo
date: 04-24-2024
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Factors influencing abundance & virus isolation in mosquitoes

Name of project author(s): Andrew Ruiz

Name of project reviewer: Erick Mollinedo

# Specific project content evaluation
Evaluate the different parts of the project by filling in the sections below.

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

The background is really strong, you included a lot of important information. However, I would suggest to shape the background/introduction section to look more like the Introduction section from a peer reviewed article. This means, have it in a single section and each paragraph represents a specific topic, meanwhile the last one has your justification (Questions/Hypothesis). I would also add references to your questions/hypothesis section. But as I said, this is a suggestion, the format you used looks good to me but I am not completely sure if this is what the instructor and TAs are expecting for the final manuscript document.

### Summary assessment (PICK ONE, DELETE THE OTHERS)

* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

I am not clear what your questions for this project are. In your manuscript, you listed 3 questions but then you mention that they will not be addressed in the project. Then you mention that they will hopefully inform research that wil answer the following questions, but I am not sure what are these other additional questions. Please clarify on the background section of your manuscript what are your aims, described as your questions/hypothesis.

### Summary assessment

* question/hypotheses somewhat explained

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

Good job describing the sources of the data, I appreciate the thoroughness of all variables. However, I would suggest to make it simpler (maybe not go into much detail) and list them as part of the Methods section in the main Manuscript. Maybe you could move most of it into the description of the data sources and what each variable means in the Supplementary material section, so you can go deeper detailing the descriptions.

### Summary assessment

* source and overall structure of data well explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

For what I saw in the main Manuscript document, I sense all the exploratory analyses are correct. I would suggest to move them to the Supplementary material document, since this should not be part of your results in the main Manuscript. Also, there were some errors while running the data processing code `MADAproject_pt2_ruiz.qmd`, but I will go deeper into details in later sections. I also suggest to add an explanation (in the processing code file) before each code chunk that describes what is the purpose of what you are coding, for example: "Load the file, merge columns and then save it as a data frame...". So far, you list what the dataset includes and the source, but I believe this should go into the Supplementary material.

### Summary assessment

* some weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

You included all the ML modeling components in my opinion and evaluated different models. However, it was not clear to me what you were modeling since your main questions/hypothesis are not clear on the Manuscript. I saw the following question in the analysis.qmd code that I thing is the main question you are trying to answer, correct? 'Does the total precipitation in April, May, and June influence the date of the first positive test?' If so, please specify it in your Manuscript. Also, add a short description of what you are aiming to to before each code chunk, and after each code chunk just add a brief explanation of what you observed/what those results mean. This will be helpful to you when integrating everything into the Results section.

### Summary assessment

* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

The results are missing in the main Manuscript. I believe you have those in your `analysis.qmd` file, but they have not been incorporated to the Manuscript file. It also caught my attention that you mention some of the results of conducting modeling under the 'Basic statistical analysis' section in Methods, which I believe these are part of the results. So I would suggest in general, to review the sections of the manuscript file so they are correctly labeled: Background, Methods, Results, Discussion.

### Summary assessment

* results are poorly presented, hard to understand, poor quality

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

Lacking conclusions and poor discussion. This is due to the main results not being included, I believe you have those in your code files but you did not include them into the manuscript.

### Summary assessment

* major parts of discussion missing or wrong

## Further comments

Overall I think you understand the whole process of data exploring and analysis, but you missed to merge all of your work in the manuscript. You have done a very good job stating the initial parts of the analysis, and I also suggest to move some of them to the Supplementary materials (most of the details about data processing and variables descriptions). Focus on working on the Manuscript so it renders as a complete peer reviewed paper with all his sections: Abstract, Introduction, Methods, Results, Discussion, Conclusions. This is what summarizes everything located in your code files.
Also, clearly state your hypothesis/questions, so all of us understand the rationale of your project.

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

I believe everything is right in his place. By having a glimpse, the folders with data have a lot of files, so I recommend carefully to select which data will be saved as processed data, you might not need all of those to be saved (or maybe you do, just a thought). Also, there were a couple .csv files floating around, but it's ok, it happens to all of us. Just check if these are not essential and maybe you could delete them. Also, make sure to update your Readme files so they show the correct structure of your folders.

### Summary assessment

* well structured

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

I suggest to include not just titles of what each chunk of code is doing. I would suggest to add a description of what you are aiming to do in the code (I mentioned before an example of how to do it). Also, you should add a brief sentence or line, explaining the results of your code when doing data exploring or analysis, so it is clear why you made decisions and why you decided to move to a new direction.

### Summary assessment

* decently documented with some gaps

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

Small parts of your codes didn't run, but overall didn't stopped the majority of the code to being executed. Make sure your R objects are correctly referenced, this is one of the first things that I noticed like for example the object `pcr_data_expanded`, which showed to me as an error that the object is not available in line #113 of the `MADAproject_pt2_ruiz.qmd` file. Another error I saw was when you were loading certain objects, for example: `trap_location = read_excel(file_path_trap)` showed me as an error that the path does not exist. To correct this, I suggest you implement loading the data using relative paths, instead of absolute paths using the `here()` function, something like this: `trap_location <- read_excel(here(file_path_trap))`.

### Summary assessment

* small parts not reproducible or required manual intervention

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

By looking at the codes, you implemented various ML modeling methods and assessments. Good job on this!
And related as previous comments, make sure to clearly state your questions/hypothesis so it is clear why you did all the statistical analysis as described in your EDA and statistical analysis codes.

### Summary assessment

* strong level of thorougness

## Further comments

A suggestion: To simplify when loading your packages, check which packages are already included in main ones. For example, if you load `tidyverse` this will already include `ggplot2`, `dplyr`, `lubridate`, same goes for `tidymodels`, which includes packages such as `parsnip`, `yardstick` and others. This way, you avoid repeating or re-loading certain packages.
Also to reiterate, make sure your code runs smoothly when loading, saving files using functions with relative paths, and add a little more description into your code chunks from all your coding files.

### 2024-04-24 products/manuscript/Manuscript.qmd
Updates to manuscript

@@
 models best describe the relationship between these factors?
-
-**This project will lay the foundation for dissertation research that
-will begin in June, 2024.**
 
@@
 ## Questions/Hypotheses to be addressed
-
-*This project will lay the foundation for my dissertation research.*
-**The goal for this phase of the project is to answer which
-environmental, ecological, and temporal factors influence the
-transmission cycles for EEE in southeastern Massachusetts.** From prior
-research we know that the mosquito life cyle is influenced by a number
+**Which environmental,ecological, and temporal factors influence the
+transmission cycles for EEE in southeastern Massachusetts and which
+models best describe the relationship between these factors?** 
+From prior
+research we know that the mosquito lifecyle is influenced by a number
 of factors. Temperature plays a major role in the mosquito life cyle
@@
 will have more time to transmit the virus. However, MEL larvae develop
-in well insulated crypts that are less sensitive to ambient air
+in well-insulated crypts that are less sensitive to ambient air
 temperature changes, so the effect of temperature changes on MEL
@@
 for EEE, especially in terms of quantifying the role these two factors
-play in EEE transmission. Specifically, there's a gap in our
-understanding of how these factors influence the number of mosquitoes
-caught in traps and the rate of virus infection among those mosquitoes.
-This gap is more pronounced when distinguishing the impacts on enzootic
-(mosquitoes that primarily feed on birds) versus bridge vector
-(mosquitoes that can transmit the virus from birds to humans) mosquito
-species. Even less well described is the effect the factors have on
-enzootic vs. bridge vector mosquito species.
+play in EEE transmission. Specifically, there's a gap in the
+understanding of how these factors influence the number of EEE vectors mosquitoes
+collected in traps and the rate of virus infection among those mosquitoes.
 
@@
 control measures.
-
-While the following questions will not be addressed in this project, it
-will hopefully inform research that will answer the following questions
-in dissertation research beginning this summer:
-
-1\. Can an AI-driven model that leverages historic mosquito
-surveillance, ecological, and environmental data accurately quantify the
-risk of human EEEv infection in southeastern Massachusetts?
-
-2\. What is the potential of machine learning algorithms to identify
-early warning signals for EEE outbreaks, enabling timely public health
-interventions in Massachusetts to prevent human infections?
-
-3\. How can machine learning models use real-time data on mosquito
-activity, environmental factors, and weather to accurately predict when
-the risk of EEE transmission to humans is low enough to lift evening
-outdoor activity restrictions? \*
 

### 2024-04-25 R/analysis-code/analysis.qmd
further edits and file cleaning

@@
 # results of those tests, and the date of collection.
-pcr <- read_csv(here("data", "processed-data", "pcr_data_expanded_filtered.csv"))
+pcr <- read_csv(here("data", "processed-data", "filtered_pcr_data.csv"))
 
@@
   group_by(week_num) %>%
-  summarise(Positive_Result = max(as.integer(Is_Positive), na.rm = TRUE))
+  summarise(Positive_Result = ifelse(all(is.na(Is_Positive)), 0, max(as.integer(Is_Positive), na.rm = TRUE)))
 
@@
 ```{r}
-pcr <- read_csv(here("data", "processed-data", "pcr_data_expanded_filtered.csv"))
+pcr <- read_csv(here("data", "processed-data", "pcr_data.csv"))
 wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
@@
 # Load and prepare mosquito data
-first_positive_dates <- pcr_data_expanded_filtered %>%
+first_positive_dates <- pcr %>%
   filter(Result == "Positive") %>%
@@
 
-# # load the dataset with a unique dataframe name
-# mir_2part_all <- read_csv(here("data", "processed-data", "all_mir_clean_joined.csv"))
+# load the dataset with a unique dataframe name
+mir_2part_all <- read_csv(here("data", "processed-data", "all_mir_clean_joined.csv"))
 

### 2024-04-25 R/eda-code/EDA_unified.qmd
further edits and file cleaning

@@
 
+#save as csv
+write_csv(filtered_pcr_data, here("data", "processed-data", "filtered_pcr_data.csv"))
+
 #calculate MIR for all sites, all traps by week
@@
 saveRDS(MIR_by_week, here("data", "processed-data", "rds", "MIR_by_week.rds"))
-View(MIR_by_week)
+str(MIR_by_week)
 
@@
 joined_mir <- read_csv(here("data", "processed-data", "joined_data_mir_co2.csv"))
-
+joined_mir
 # Assuming 'joined_mir' is your dataframe
@@
                                   "Average_MEL", "Total_MEL", "MEL_Prop")]
+
+# Summarize missing values for each variable
+missing_values <- sapply(joined_mir[, c("Avg_TMAX", "Total_PRCP", "Hrs_Above_80_lag", 
+                                        "Hrs_Above_75_lag", "Hrs_Below_50_lag", 
+                                        "Avg_HourlyDryBulbTemp_lag", "Avg_HourlyRelativeHumidity_current", 
+                                        "Avg_HourlyWetBulbTemperature_current", "Avg_HourlyWindSpeed_current", 
+                                        "DSCI_fema", "average_daylight_hours", 
+                                        "Average_MEL", "Total_MEL", "MEL_Prop")], function(x) sum(is.na(x)))
+print(missing_values)
 

### 2024-04-25 R/processing-code/MADAproject_pt2_ruiz.qmd
further edits and file cleaning

@@
 # Construct the path to the Excel file
-#file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")
 file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")
@@
 # Save as RDS
-saveRDS(pcr_data_expanded, here("data", "processed-data", "rds", "pcr_data"))
+saveRDS(pcr_data, here("data", "processed-data", "rds", "pcr_data.rds"))
 
 # Save as CSV
-write_csv(pcr_data_expanded, here("data", "processed-data", "pcr_data.csv"))
+write_csv(pcr_data, here("data", "processed-data", "pcr_data.csv"))
 
@@
 # Construct the path to the file
-file_path_trap = here("data", "processed-data", "BCMCP_trap_sites_2021.xlsx")
+file_path_trap = here("data", "raw-data", "BCMCP_trap_sites_2021.xlsx")
 
@@
 hourly_wx <- read_csv(hourly_wx_path, show_col_types = FALSE)
-View(hourly_wx)
+str(hourly_wx)
 hourly_wx_clean <- hourly_wx %>%
@@
   select(STATION, DATE, HourlyDryBulbTemperature, HourlyRelativeHumidity, HourlyWetBulbTemperature, HourlyWindSpeed, week_num)
-
-str(hourly_wx_clean)
 
@@
 drought_fema1_4 = read_csv(here("data", "raw-data", "comp_stat_fema1_4.csv"), show_col_types = FALSE)
-#View(drought_fema1_4)
+
 # Add week number based on MapDate for both files
@@
   )
-#View(drought_ma_counties)
 
@@
 # Construct the path to the human case data file
-file_path_human <- here("data", "processed-data", "human_cases.xlsx")
+file_path_human <- here("data", "raw-data", "human_cases.xlsx")
 
@@
 #date of onset is missing for some records, however, they are not needed for this project. 
-```
-
-## Bird data
-#### This dataset comes from the annual North American Breeding Bird Survey that is conducted in a joint effort by: USGS Eastern Ecological Science Center, Environment and Climate Change Canada, and the Mexican National Commission for the Knowledge and Use of Biodiversity.
-#### There are 2 (5 and 11) routes in Bristol County, MA. The data is from 2007-2022. However, route 11 is mission data for 2013 and route 5 ends in 2019.
-#### For route 11, 2013, we will use the data from 2012 and 2014 to estimate the missing data.
-for route 5, 2020-2022, we will use the data from 2019 to estimate the missing data.
-```{r}
-# Load the dataset
-file_path_bird <- here("data", "raw-data", "Massach.csv")
-bird_data_raw <- read.csv(file_path_bird)
-
-# Filter and clean up the data
-bird_bbs <- bird_data_raw %>%
-  filter(Route %in% c(5, 11), Year >= 2006 & Year <= 2022) %>%
-  select(-AOU, -Count10, -Count20, -Count30, -Count40, -Count50)
-
-# Summarize the data
-bird_bbs_summarized <- bird_bbs %>%
-  group_by(Route, Year) %>%
-  summarize(
-    StopTotal = sum(StopTotal, na.rm = TRUE),
-    SpeciesTotal = sum(SpeciesTotal, na.rm = TRUE),
-    .groups = 'drop'
-  )
-
-# Check for missing data within Routes 5 and 11
-missing_data <- map_df(c(5, 11), ~tibble(
-  Route = .x,
-  missing_years = setdiff(2006:2022, bird_bbs_summarized$Year[bird_bbs_summarized$Route == .x])
-))
-
-# Identify all combinations of Route and Year within the range
-all_combinations <- expand.grid(Route = unique(bird_bbs_summarized$Route),
-                                Year = 2006:2022)
-
-# Join with the summarized data to identify missing rows
-bird_bbs_full <- left_join(all_combinations, bird_bbs_summarized, by = c("Route", "Year"))
-
-# Impute missing data by taking the average of the previous and next year for each route
-bird_bbs_imputed <- bird_bbs_full %>%
-  group_by(Route) %>%
-  arrange(Route, Year) %>%
-  mutate(
-    StopTotal = ifelse(is.na(StopTotal), (lag(StopTotal) + lead(StopTotal)) / 2, StopTotal),
-    SpeciesTotal = ifelse(is.na(SpeciesTotal), (lag(SpeciesTotal) + lead(SpeciesTotal)) / 2, SpeciesTotal)
-  ) %>%
-  ungroup()
-
-# Handling the last records for Route 5 (2020-2022)
-# Since there's no future data to average, you might replicate the most recent data (e.g., from 2019)
-# Or consider another imputation method suitable for your analysis and data characteristics
-
-# Find the most recent year's data for Route 5
-recent_data_route_5 <- bird_bbs_imputed %>% 
-  filter(Route == 5, Year == max(Year[Route == 5 & !is.na(StopTotal)]))
-
-# Create a template for the missing years
-missing_years_route_5 <- tibble(Route = 5, Year = 2020:2022)
-
-# Replicate the most recent data for the missing years
-imputed_missing_years_route_5 <- bind_rows(lapply(missing_years_route_5$Year, function(y) {
-  mutate(recent_data_route_5, Year = y)
-})) %>%
-  bind_rows(bird_bbs_imputed) %>%
-  arrange(Route, Year)
-
-# Check if any adjustments or specific imputations are needed
-print(imputed_missing_years_route_5)
-
-bird_clean <- imputed_missing_years_route_5 %>%
-  drop_na()
-
-# Save the cleaned bird data as an csv
-write_csv(bird_clean, here("data", "processed-data", "bird_clean.csv"))
 ```
@@
  mutate(
-    `collection date` = as.Date(`collection date`),
     week_num = paste0(year(`collection date`), "-", sprintf("%02d", isoweek(`collection date`)))
@@
 # Output the transformed data to inspect its structure
-head(mosquito_wide)
+str(mosquito_wide)
 
@@
 # Save the wide-format data to the specified path
-write_csv(wide_data, file_path_wide)
+write_csv(mosquito_wide, file_path_wide)
 
@@
 # Save the wide-format data as an RDS file
-saveRDS(wide_data, file_path_wide_rds)
-
-# Now we will filter the the mosquito data to match the time period of the virus data. 
-# Since mosquito populations can be affected by the previous year's populations, we will filter for 2007-2023
-
-# Filter wide_data for the specified date range
-filtered_mosquito_data <- wide_data %>%
-  filter(`collection date` >= as.Date("2007-01-01") & `collection date` <= as.Date("2023-12-31"))
-
-# Define the path for the CSV file
-file_path_filtered_csv <- here("data", "processed-data", "mosquito_wide_filtered.csv")
-
-# Save the filtered wide-format data to the specified CSV path
-write_csv(filtered_mosquito_data, file_path_filtered_csv)
-View(mosquito_wide_filtered.csv)
-# Define the path for the RDS file
-file_path_filtered_rds <- here("data", "processed-data", "rds", "mosquito_wide_filtered.rds")
-
-# Save the filtered wide-format data as an RDS file
-saveRDS(filtered_mosquito_data, file_path_filtered_rds)
+saveRDS(mosquito_wide, file_path_wide_rds)
 
@@
 
-# Read the data
-trap_proportion <- read_csv(trap_wide_path) %>%
+# Read the data from the file
+trap_wide_data <- read_csv(trap_wide_path)
+
+trap_wide_data <- trap_wide_data %>%
   mutate(Year = year(as.Date(`collection date`, format = "%Y-%m-%d")))
-
+trap_wide_data
 # Assuming all species columns are next to each other, calculate the proportion of MEL
-proportion_mel_by_year <- trap_proportion %>%
-  # Calculate row-wise totals for all species
+proportion_mel_by_year <- trap_wide_data %>%
   rowwise() %>%
@@
   ungroup() %>%
-  # Then group by year and summarize
   group_by(Year) %>%
-  summarise(Total_MEL = sum(MEL, na.rm = TRUE),
+  summarise(
+    Total_MEL = sum(MEL, na.rm = TRUE),
     Total_All = sum(Total_All, na.rm = TRUE), # Sum of all species by year
-            MEL_Prop = Total_MEL / Total_All) # Proportion of MEL
+    MEL_Prop = Total_MEL / Total_All # Proportion of MEL
+  )
 
@@
 # Assuming all species columns are next to each other, calculate the proportion of MEL by week_num
-proportion_mel_by_week <- trap_proportion %>%
+proportion_mel_by_week <- trap_wide_data %>%
   # Calculate row-wise totals for all species
@@
 # Load the datasets
-yearly_mel_ave <- readRDS(here("data", "processed-data", "rds", "yearly_mel_ave.rds"))
+yearly_mel_ave <- readRDS(here("data", "processed-data", "rds", "yearly_mel_average.rds"))
 summed_prcp_by_year_filtered <- readRDS(here("data", "processed-data", "rds", "summed_prcp_by_year_filtered_2013_2020.rds"))
@@
 
-plot_mel_precip = ggplot(data = combined_data) +
+plot_mel_precip <- ggplot(data = combined_data) +
   geom_col(aes(x = factor(Year), y = Average_MEL), fill = "blue") +
-  geom_line(aes(x = factor(Year), y = Total_PRCP, group = 1), color = "red", linetype = "dashed") +
+  geom_line(aes(x = factor(Year), y = Total_PRCP, group = 1), color = "red", linetype = "dashed", linewidth = 1) +
   geom_point(aes(x = factor(Year), y = Total_PRCP), color = "red", size = 3) +
@@
         axis.title.x = element_text(face = "bold"))
+
 plot_mel_precip
@@
 ```
-
-### Calcualte the average moon phase by week_num.
-#### some resaecrh suggests that moon phase can affect mosquito activity.
-##### The moon phase is calculated using the getMoonIllumination function from the lunar package.
-The function returns the percentage of the moon's illuminated surface area and the moon's phase angle.
-The phase angle is used to determine the moon phase, which can be one of the following: "New Moon", "First Quarter", "Full Moon", or "Last Quarter".
-The phase angle is converted to a value between 0 and 1, where 0 corresponds to a New Moon, 0.5 corresponds to a Full Moon, and 1 corresponds to a New Moon.
-The average moon phase for each week is calculated by taking the mean of the moon phase values for each day in the week.
-
-```{r}
-# Generate a sequence of dates from 2007-01-01 to 2023-12-31
-dates <- seq(as.Date("2007-01-01"), as.Date("2023-12-31"), by = "day")
-
-# Function to get moon phase for a given date
-get_moon_phase <- function(date) {
-  illumination <- getMoonIllumination(date)
-  phase <- illumination$phase
-  return(phase)
-}
-
-# Calculate moon phase for each date
-moon_phases <- sapply(dates, get_moon_phase)
-
-# Create a data frame with dates and their corresponding moon phases
-moon_phases_df <- data.frame(Date = dates, MoonPhase = moon_phases)
-
-# View the first few rows of the data frame
-str(moon_phases_df)
-
-# Add week_num column to moon_phases_df
-moon_phases_df <- moon_phases_df %>%
-  mutate(
-    week_num = paste0(year(Date), "-", sprintf("%02d", isoweek(Date)))
-  )
-
-# Calculate average moon phase by week_num
-average_moon_phase_by_week <- moon_phases_df %>%
-  group_by(week_num) %>%
-  summarise(AverageMoonPhase = mean(MoonPhase, na.rm = TRUE))
-
-# View the results
-head(average_moon_phase_by_week)
-
-```

### 2024-04-25 products/manuscript/Manuscript.qmd
further edits and file cleaning

@@
 
-*This project will inform the development of my dissertation prospectus
-by identifying significant relationships between variables and guide the
-methods development.* Eastern equine encephalitis (EEE) is caused by an
-Alphavirus transmitted to humans by the bite of an infected mosquito.
-Human infections are rare, but serious. About 30-50% of symptomatic
-cases lead to death and those who survive are left with life-long
-disability. The transmission cycle for EEE includes two distinct cycles:
-an amplifying enzootic cycle in which the virus is transmitted between
-the *Culiseta melanura* mosquito and birds and an epizootic cycle in
-which bridge vector mosquitoes transmit the EEE virus (EEEv) from birds
-to humans and other mammals [@goZoonoticEncephalitidesCaused2014].
-Because C. melanura feeds almost exclusively on birds, it is not
-considered a direct human threat. However, EEEv isolations in *C.
-melanura* are an early warning sign that EEEv is circulating in the
-ecosystem. Isolations in bridge vector mosquito species indicate
-heightened transmission risk to humans.
+Eastern equine encephalitis (EEE) is a rare but serious disease caused by an Alphavirus transmitted to humans by infected mosquitoes.
+With a fatality rate of 30-50% and potential lifelong disability for survivors, understanding its transmission cycle is crucial.
+EEE transmission involves two cycles: an enzootic cycle between Culiseta melanura mosquitoes and birds, and an epizootic cycle where bridge vector mosquitoes transmit the virus to humans and other mammals.
+While C.
+melanura poses minimal direct threat to humans, isolations of EEE virus in these mosquitoes serve as early indicators of virus circulation.
+Conversely, isolations in bridge vector species signal heightened transmission risk.
+This research aims to identify significant relationships between variables to inform dissertation prospectus development and guide methodological approaches.
 :::
@@
 
-Moreover, the influence of other variables remains even less explored.
-Birds, especially passerine species, serve as amplifying hosts for EEEV
-and WNV, potentially affecting the level of virus present in an area.
-Despite their importance in the virus transmission cycle, the effect of
-fluctuations in bird populations on the prevalence of these viruses
-throughout a season has received minimal attention. Finally, given the
-migratory nature of many passerine species, it's worth exploring whether
-EEEV activity in the warmer regions of the United States, during the
-months preceding mosquito season in Massachusetts, could serve as an
-early indicator of virus activity in the state. Understanding these
-dynamics is crucial for predicting outbreaks and implementing effective
-control measures.
-
 ## Data collection and processing
@@
 FEMA East Coast regions (I-IV) drought indicators.
-
-### Human and animal case data
-
-While the goal of this project is to set a foundation for models that
-predict human cases of EEE later this year. Human case data will not be
-an outcome in the models for this project. However, the case data will
-incorporated into the models. Some have hypothesized that EEE activity
-in the southern United States could serve as an early warning signal for
-EEE activity in Massachusetts. The rationale is that birds that migrate
-from the south to Massachusetts could bring the virus with them. In
-active EEE years, more infected birds may arrive in Massachusetts and
-spread the virus to local mosquitoes at a higher rate. Florida is a
-known hotspot for EEE activity and is the only state that reports more
-human cases than Massachusetts. The Florida Department of Health reports
-EEE activity in animals and humans annually. In addition to the wild and
-domestic animal case reports, many mosquito control programs in Florida
-maintain flocks of sentinel chickens that are tested for EEE and WNV.
-The sentinel chicken data is used to monitor virus activity in the
-mosquito population.
-
-### Bird observation data
-
-The North American Breeding Bird Survey (BBS) is conducted annually in
-late spring or early summer. The United State Geological Survey (USGS)
-maintains the BBS database. The BBS is a long-term, volunteer-driven
-bird monitoring program that collects data on bird populations across
-North America. The data is used to track the status and trends of bird
-populations over time. The BBS data includes information on the number
-of birds observed, the location of the observation, and the date of the
-observation. The data is collected by trained volunteers who conduct
-roadside surveys along predetermined routes. The data is used to
-estimate the population size and distribution of bird species across the
-continent. The BBS data will be used to estimate the population size and
-distribution of bird species in Massachusetts.
-
-Since birds are the amplifying hosts for EEE, their population size and
-distribution can have a significant impact on the prevalence of the
-virus in the environment. Additionally, the migratory patterns of birds
-can influence the spread of the virus across regions. By incorporating
-bird observation data into the models, we can better understand the
-relationship between bird populations and EEE activity in Massachusetts.
-
-### Temporal scale
-
-For most models in this project, the outcome, MIR, will be calculated by
-week. While the main temporal scale for this project is the week, the
-data will be aggregated to other temporal scales. The week is the most
-common temporal scale for mosquito surveillance data. The week number
-will be calculated according to ISO 8601 standards, and stored in a
-YYYY-WW format. For example, week 8 in 2019 will be displayed as
-2019-08. The week number will provide a common field by which all
-datasets with a temporal attribute can be joined. It will also allow for
-grouping records and making calculations by week number. This format
-helps to standardize temporal scales and make comparisons between years
-easier.
-
-While mosquito surveillance and weather data can be aggregated to the
-week, other datasets, such as the Annual BBS cannot appropriately be
-aggregated by week. Additionally, mosquito surveillance in Massachusetts
-is conducted from June to October and EEE activity in Florida occurs
-year-round. For these predictors, the temporal scale will be the year.
 
@@
 ## Exploratory/Descriptive analysis
-
-It is important to know where the mosquitoes infected with EEE are
-captured. Below is a table showing the virus isolations for EEE in
-trapped mosquitoes by county between 2014 and 2020.
-
-![](images/eee_isolation_county-04.png){fig-align="center" width="192"}
-
 One factor to consider is if early virus isolations are associated with
@@
 ![](images/eee_isolation_cases-04.png){fig-align="center" width="251"}
-
-Unfortunately, the time horizon is limited by data availability.
-Additionally, EEE human infections are rare, so traditional statistical
-methods may be inadequate to suggest any relationships. What is clear is
-that 2019 was an exceptional year for EEE.
 
@@
 ## Basic statistical analysis
+Basic linear models assessed environmental factors' impact on the EEE virus transmission cycle.
+Initial models investigated the link between virus isolation in mosquitoes and weather conditions (average temperature by week and total precipitation).
+The outcome was whether at least one positive test occurred for a pool in a given week.
+The predictor in the first model was the average maximum daily temperature from the previous week.
+The logistic model revealed a significant relationship.
+Higher temperatures were positively associated with the likelihood of a positive result (odds ratio ≈ 1.069 for each unit increase in temperature; Estimate = 0.06695, SE = 0.02033, p < 0.001).
+Incorporating Lagged_Avg_TMAX as a predictor improved the model, reducing residual deviance to 405.79 from the null deviance of 417.24, indicating a better fit.
+However, residual deviance on 299 degrees of freedom suggests unexplained variability, indicating other influential factors.
+The model's AIC of 411.59 suggests a balance between complexity and predictive power, supporting its utility.
+Yet, this relationship warrants closer examination due to likely autocorrelation of temperature with progressing weeks during summer months
 
-The initial tests have been basic regression models.
-
-The first models explored the relationship between virus isolation in
-mosquitoes and weather conditions (average temperature by week number
-and total precipitation by week). The outcome was whether there was at
-least 1 positive test for pool in a given week. The predictor in the
-first model was the average maximum daily temperature the week before
-the mosquito collection week. The logistic model suggests that there was
-a relationship between the two. (The R code for the model can be found
-in the analysis-code file: analysis.qmd) The average temperature in the
-week before is positively associated with the likelihood of a positive
-result, with an odds ratio of approximately 1.069 for each unit increase
-in temperature (Estimate = 0.06695, SE = 0.02033, p \< 0.001),
-indicating that higher temperatures significantly increase the
-probability of a positive outcome. The model's improvement is
-highlighted by a decrease in residual deviance to 405.79 from the null
-deviance of 417.24, upon incorporating Lagged_Avg_TMAX as a predictor,
-demonstrating a better fit to the observed data. However, with a
-residual deviance of 405.79 on 299 degrees of freedom, it is evident
-that not all variability in the outcome is explained, suggesting the
-presence of other influential factors. The model's AIC of 411.59
-indicates a balance between model complexity and predictive power,
-supporting its relative utility in understanding the data. However, this
-relationship should be examined more closely as temperature is likely
-auto correlated with progressing weeks during summer months. While
+While
 precipitation has been demonstrated to influence the population dynamics

### 2024-04-25 R/eda-code/archived_files/Model_mir_temp.R

@@ -1,298 +1,0 @@
-# Load necessary libraries for data manipulation, date handling, and file path management
-library(lubridate)
-library(here)
-library(dplyr)
-library(ggplot2)
-library(pscl)
-library(MASS)
-library(readr)
-library(purrr)
-
-
-# Load the datasets:
-MIR_by_week_co2 <- read_csv(here("data", "processed-data", "MIR_by_week_co2.csv"))
-wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
-mosquito_wide <- read_csv(here("data", "processed-data", "mosquito_wide_all.csv"))
-drought_fema1_4 <- read_csv(here("data", "processed-data", "drought_fema1_4.csv"))
-drought_ma_counties <- read_csv(here("data", "processed-data", "drought_ma_counties.csv"))
-#average_moon_phase_by_week <- read_csv(here("data", "processed-data", "average_moon_phase_by_week.csv"))
-average_daylight_by_week <- read_csv(here("data", "processed-data", "average_daylight_by_week.csv"))
-average_mel_weekco2 <- read_csv(here("data", "processed-data", "average_mel_weekco2.csv"))
-wx_hourly_clean <- read_csv(here("data", "processed-data", "hourly_wx_clean.csv"))
-bird_clean <- read_csv(here("data", "processed-data", "bird_clean.csv"))
-mel_prop_weekly <- read_csv(here("data", "processed-data", "mel_proportion_by_week.csv"))
-
-# Aggregate weather data by week:
-# Calculate the average maximum temperature (Avg_TMAX) and total precipitation (Total_PRCP) for each week.
-# This simplification allows for weekly comparisons with the MIR data.
-wx_aggregated <- wx_selected %>%
-  group_by(week_num) %>%
-  summarise(Avg_TMAX = mean(TMAX, na.rm = TRUE),
-            Total_PRCP = sum(PRCP, na.rm = TRUE),
-            .groups = 'drop')
-
-# Adjust for a one-week lag in weather data:
-# Create a new column (Lagged_week_num) that represents each week's preceding week. 
-# This adjustment accounts for the delayed effect weather might have on MIR.
-wx_aggregated <- wx_aggregated %>%
-  mutate(
[…]

### 2024-04-25 R/eda-code/archived_files/PT3_edacode.R

@@ -1,85 +1,0 @@
-## ---- packages --------
-#load needed packages. make sure they are installed.
-library(here) #for data loading/saving
-library(dplyr)
-library(skimr)
-library(ggplot2)
-library(tidyverse)
-library(lubridate)
-library(readr)
-library(forecast)
-library(tsibble) # For time series data wrangling
-library(feasts) # For decomposing and visualizing time series
-
-pcr <- read_csv(here("data", "processed-data", "pcr_data_expanded_filtered.csv"))
-wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
-
-# Step 2: Calculate the average TMAX for April, May, and June separately
-avg_temp_by_month <- wx_selected %>%
-  filter(month(DATE) %in% c(4, 5, 6)) %>%
-  group_by(Year = year(DATE), Month = month(DATE)) %>%
-  summarise(Avg_TMAX = mean(TMAX, na.rm = TRUE), .groups = 'drop')
-
-# Pivot the average temperatures into separate columns for each month
-avg_temp_by_month_wide <- avg_temp_by_month %>%
-  pivot_wider(names_from = Month, values_from = Avg_TMAX, names_prefix = "Avg_TMAX_M") %>%
-  # Ensure all years are present even if some data might be missing
-  complete(Year, fill = list(Avg_TMAX_M4 = NA, Avg_TMAX_M5 = NA, Avg_TMAX_M6 = NA))
-
-# Step 4: Identify the first positive test date each year in mosquito_data
-first_positive_dates <- pcr %>%
-  filter(Result == "Positive") %>%
-  group_by(Year = year(`Collection Date`)) %>%
-  summarise(First_Positive_Date = min(`Collection Date`), .groups = 'drop')
-
-# Step 5: Merge the datasets on Year
-combined_data_firstpos <- left_join(first_positive_dates, avg_temp_by_month_wide, by = "Year")
-
-# Step 6: Convert First_Positive_Date to a numerical value (e.g., day of the year)
-combined_data_firstpos <- combined_data_firstpos %>%
-  mutate(First_Positive_DOY = yday(First_Positive_Date))
-
-# Explore the relationship via linear regression with Avg_TMAX_M4, Avg_TMAX_M5, Avg_TMAX_M6 as predictors
[…]

### 2024-04-25 R/eda-code/archived_files/logistic_.pos_temp_precip.R

@@ -1,71 +1,0 @@
-library(here)
-library(dplyr)
-library(lubridate)
-library(readr)
-
-
-# Load the PCR data set which includes information on mosquito pools tested for viruses, 
-# results of those tests, and the date of collection.
-pcr <- read_csv(here("data", "processed-data", "pcr_data_expanded_filtered.csv"))
-
-# Load the weather data set which includes daily maximum temperatures (TMAX) and other weather variables.
-wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
-
-# Aggregate the weather data by week to calculate the average maximum temperature for each week.
-# This simplifies the daily temperature data into a more manageable weekly summary,
-# making it easier to correlate with weekly mosquito test results.
-wx_selected <- wx_selected %>%
-  group_by(week_num) %>%
-  summarise(
-    Avg_TMAX = mean(TMAX, na.rm = TRUE),
-    Total_PRCP = sum(PRCP, na.rm = TRUE)
-  ) %>%
-  ungroup()
-
-# View the first few rows to confirm it worked
-head(wx_selected)
-
-
-# Create a lagged version of the average weekly maximum temperature to explore the potential delayed effect of temperature on mosquito test results.
-# The lag function shifts the temperature data by one week, assuming that the impact of temperature on virus transmission might not be immediate.
-weather_weekly <- wx_selected %>%
-  arrange(week_num) %>%
-  mutate(Lagged_Avg_TMAX = lag(Avg_TMAX, 1))
-
-weather_weekly <- wx_selected %>%
-  arrange(week_num) %>%
-  mutate(
-    Lagged_Avg_TMAX = lag(Avg_TMAX, 1),
-    Lagged_Total_PRCP = lag(Total_PRCP, 1)
-  )
-
-weather_weekly <- wx_selected %>%
-  arrange(week_num) %>%
-  mutate(
-    Lagged_Avg_TMAX = lag(Avg_TMAX, 1),
-    Lagged_Total_PRCP = lag(Total_PRCP, 1),
-    Binary_PRCP = if_else(Lagged_Total_PRCP > 0, 1, 0)
-  )
-# Aggregate the PCR data by week to determine if there was at least one positive result in each week.
-# This transformation reduces the detailed test results into a weekly summary,
-# indicating the presence or absence of positive tests each week.
-pcr_weekly <- pcr %>%
-  group_by(week_num) %>%
[…]

### 2024-04-26 CMiller-project-review-Ruiz.qmd
C Miller Part 5 Review
C Miller Part 5 Review

---
title: Project Review Factors Influencing Abundance & Virus Isolation in Mosquitoes
author: Clarke Miller
date: 26 April 2024
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: **Factors Influencing Abundance & Virus Isolation in Mosquitoes**
Subtitle:  ***Which environmental, ecological, and temporal factors impact mosquito populations***

Name of project author(s):  Andrew Ruiz

Name of project reviewer:  Clarke Miller

## Background, Context and Motivation

*How well is the context of the project described?*

Very well.  This looks great.

*Is a comprehensive background, including summary of previous/related work given?*

Yes.  Very much so.

*Is the project well placed into the context of existing work (including proper referencing of existing work).* 
Yes, very much so.

*Is it clear why the project was undertaken and what new information it hopes to provide?*

Yes.  Very much so.

***Personal Message***  
Andrew, this looks great.  From our conversations over Discord, you were very critical of your own work.  This project looks awesome.  You have a couple of things to polish off, but, wow.  You obviosuly have the topic covered, you have the coding skills, you got this.

### Feedback and Comments

No real feedback on this part.  You did a great job in this category.

### Summary assessment (PICK ONE, DELETE THE OTHERS)

* strong contextualization and motivation

## Question description

*How well and clear are the question(s)/hypotheses the project aims to address described?*

Very clear.  The questions are very well outlined and discussed.

*Is it clear how the questions relate to the data?*

Yes.

### Feedback and Comments

No real feedback on this part.  You did a great job in this category.

### Summary assessment

* question/hypotheses fully clear

## Data description

*How well is the data overall described?*

Very well.  I would like to see a table summary of the data.

*Is the source provided?*

Yes.

*Is a codebook or other meta-information available that makes it clear what the data is?*

I did not see one.

### Feedback and Comments

No real feedback on this part.  You did a great job in this category.  Just maybe add a summary table and if there is a codebook add it to the raw data and label it.

### Summary assessment

* source and overall structure of data well explained

## Data wrangling and exploratory analysis

*How well is the data cleaned/processed and explored?*

Yes.  You did a great job with this.

*Are all steps reasonable and well explained?*

Yes.  The steps are reasonable and explained.

*Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?*

At this point in time, no.  There are placeholders for this material in the document.

### Feedback and Comments

No real feedback on this part.  You did a great job in this category.  Despite being a work in progress, this section is well done.

### Summary assessment

* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

*Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?*

At this point, this section of the document is incomplete.

### Feedback and Comments

At this point, this section of the document is incomplete.  That being said, the description of the intended analysis is appropriate.

### Summary assessment

* strong and reasonable analysis/description.

## Presentation

*How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?*

There are no results at this time.

### Feedback and Comments

At this point, this section of the document is incomplete.

### Summary assessment

* results are poorly presented, hard to understand, poor quality

At this point, this section of the document is incomplete.

## Discussion/Conclusions

*Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?*

At this point, this section of the document is incomplete.

### Feedback and Comments

At this point, this section of the document is incomplete.

### Summary assessment
* major parts of discussion missing or wrong

At this point, this section of the document is incomplete.

## Further comments

A large portion of this work is missing.  It is my understanding that the author is working through these sections again as new data (rainfall/weather data) was made avaiable.

The basic bones of the project look good.

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure

*Is the project well structured? Are files in well labeled folders?*

Yes.

*Do files have reasonable names?*

Yes.

*Are all "junk" files not needed for analysis/reproduction removed?*

Not at this time.

*By just looking at files and folders, can you get an idea of how things fit together?*

 Yes.

### Feedback and Comments

At this point, there are a lot of "sandbox" files still around.

### Summary assessment

* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

## Documentation

*How well is the project documented?*

Fairly well.  The readme file is being used... it could use a better description.

*Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?*

Yes, as far as the work has been done.

### Feedback and Comments

At this point, this section of the document is incomplete.

### Summary assessment

* decently documented with some gaps

## Reproducibility

*Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?*

Yes.  As far as it goes.  At this point, this section of the document is incomplete.

### Feedback and Comments

At this point, this section of the document is incomplete.

### Summary assessment

* fully reproducible without issues

## Thoroughness

*How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?*

Once complete, the described work will be very complete.  At this point this section is incomplete.

### Feedback and Comments

At this point, this section of the document is incomplete.

### Summary assessment

* decent level of thoroughness

## Further comments

_Add any other comments regarding the overall project here. Write anything you think can help your classmate improve their project._

Just finish this project off the way you've described and it will be great.  Hang in there!

### 2024-04-26 R/processing-code/MADAproject_pt2_ruiz.qmd
testing changes

@@
 # Construct the path to the veterinary case data file
-file_path_vet <- here("data", "processed-data", "animal_cases.xlsx")
+file_path_vet <- here("data", "raw-data", "animal_cases.xlsx")
 
@@
 # Calculate the average number of MEL by year
-average_mel_by_year <- trap_proportion %>%
+average_mel_by_year <- trap_wide_data %>%
   group_by(Year) %>%

### 2024-04-26 R/analysis-code/analysis.qmd
updating manuscript

@@
 library(gbm)
+library(knitr)
+library(gt)
 ```
@@
 
-```
+# Use gt to create a styled table
+gt_table <- gt(rmse_table) %>%
+  tab_header(
+    title = "RMSE Comparison of Models",
+    subtitle = "Sorted by Test RMSE in descending order"
+  ) %>%
+  cols_label(
+    Model = "Model",
+    Train_RMSE = "Training RMSE",
+    Test_RMSE = "Testing RMSE"
+  )
+
+# Define the file path using 'here'
+png_path <- here("results", "tables", "rmse_table.png")
+
+# Save the gt table as a PNG
+gtsave(gt_table, filename = png_path)
+```

### 2024-04-26 products/manuscript/Manuscript.qmd
updating manuscript

@@
 
-Eastern equine encephalitis (EEE) is a rare but serious disease caused by an Alphavirus transmitted to humans by infected mosquitoes.
-With a fatality rate of 30-50% and potential lifelong disability for survivors, understanding its transmission cycle is crucial.
-EEE transmission involves two cycles: an enzootic cycle between Culiseta melanura mosquitoes and birds, and an epizootic cycle where bridge vector mosquitoes transmit the virus to humans and other mammals.
-While C.
-melanura poses minimal direct threat to humans, isolations of EEE virus in these mosquitoes serve as early indicators of virus circulation.
-Conversely, isolations in bridge vector species signal heightened transmission risk.
-This research aims to identify significant relationships between variables to inform dissertation prospectus development and guide methodological approaches.
+Eastern equine encephalitis (EEE) is a rare but serious disease caused
+by an Alphavirus transmitted to humans by infected mosquitoes. With a
+fatality rate of 30-50% and potential lifelong disability for survivors,
+understanding its transmission cycle is crucial. EEE transmission
+involves two cycles: an enzootic cycle between Culiseta melanura
+mosquitoes and birds, and an epizootic cycle where bridge vector
+mosquitoes transmit the virus to humans and other mammals. While C.
+melanura poses minimal direct threat to humans, isolations of EEE virus
+in these mosquitoes serve as early indicators of virus circulation.
+Conversely, isolations in bridge vector species signal heightened
+transmission risk. This research aims to identify significant
+relationships between variables to inform dissertation prospectus
+development and guide methodological approaches.
 :::
@@
 ## Questions/Hypotheses to be addressed
+
 **Which environmental,ecological, and temporal factors influence the
 transmission cycles for EEE in southeastern Massachusetts and which
-models best describe the relationship between these factors?** 
-From prior
-research we know that the mosquito lifecyle is influenced by a number
-of factors. Temperature plays a major role in the mosquito life cyle
-virus transmission dynamics. As temperature increases, the rate of
+models best describe the relationship between these factors?** From
+prior research we know that the mosquito lifecyle is influenced by a
+number of factors. Temperature plays a major role in the mosquito life
+cyle virus transmission dynamics. As temperature increases, the rate of
 larval development also increases and the time between when a mosquito
@@
 play in EEE transmission. Specifically, there's a gap in the
-understanding of how these factors influence the number of EEE vectors mosquitoes
-collected in traps and the rate of virus infection among those mosquitoes.
+understanding of how these factors influence the number of EEE vectors
+mosquitoes collected in traps and the rate of virus infection among
+those mosquitoes.
 
@@
 ## Exploratory/Descriptive analysis
+
 One factor to consider is if early virus isolations are associated with
@@
 ## Basic statistical analysis
-Basic linear models assessed environmental factors' impact on the EEE virus transmission cycle.
-Initial models investigated the link between virus isolation in mosquitoes and weather conditions (average temperature by week and total precipitation).
-The outcome was whether at least one positive test occurred for a pool in a given week.
-The predictor in the first model was the average maximum daily temperature from the previous week.
-The logistic model revealed a significant relationship.
-Higher temperatures were positively associated with the likelihood of a positive result (odds ratio ≈ 1.069 for each unit increase in temperature; Estimate = 0.06695, SE = 0.02033, p < 0.001).
-Incorporating Lagged_Avg_TMAX as a predictor improved the model, reducing residual deviance to 405.79 from the null deviance of 417.24, indicating a better fit.
-However, residual deviance on 299 degrees of freedom suggests unexplained variability, indicating other influential factors.
-The model's AIC of 411.59 suggests a balance between complexity and predictive power, supporting its utility.
-Yet, this relationship warrants closer examination due to likely autocorrelation of temperature with progressing weeks during summer months
 
-While
-precipitation has been demonstrated to influence the population dynamics
-it did not show to have any effect on a positive result. This was
-initailly surprising, but upon after considering the specific habitat of
-the main EEE enzootic vector, *Culiseta melanura*, it makes sense. MEL
-larvae develop in well insulated crypts that are less sensitive
-evaporative processes than surface water. The effect of precipitation on
-MEL larval habitat abundance may be less pronounced than other vector
-species that develop in open, standing water, like ditches and
-containers. Another model assessed whether the temperature precipitation
-in April, May, and June of each year effected the date of the first
-positive pool test (R file: PT3_edacode). The date of the first positive
-pool was identified and assigned a day number from January 1st of that
-year. The predictors in this linear model were the average maximum
-temperature for April, May, and June and the sum of precipitation for
-the same time period. None of these predictors were significant.
+Basic linear models assessed environmental factors' impact on the EEE
+virus transmission cycle. Initial models investigated the link between
+virus isolation in mosquitoes and weather conditions (average
+temperature by week and total precipitation). The outcome was whether at
+least one positive test occurred for a pool in a given week. The
+predictor in the first model was the average maximum daily temperature
+from the previous week. The logistic model revealed a significant
+relationship. Higher temperatures were positively associated with the
+likelihood of a positive result (odds ratio ≈ 1.069 for each unit
+increase in temperature; Estimate = 0.06695, SE = 0.02033, p \< 0.001).
+Incorporating Lagged_Avg_TMAX as a predictor improved the model,
+reducing residual deviance to 405.79 from the null deviance of 417.24,
+indicating a better fit. However, residual deviance on 299 degrees of
+freedom suggests unexplained variability, indicating other influential
+factors. The model's AIC of 411.59 suggests a balance between complexity
+and predictive power, supporting its utility. Yet, this relationship
+warrants closer examination due to likely autocorrelation of temperature
+with progressing weeks during summer months
+
+While precipitation has been demonstrated to influence the population
+dynamics it did not show to have any effect on a positive result. This
+was initailly surprising, but upon after considering the specific
+habitat of the main EEE enzootic vector, *Culiseta melanura*, it makes
+sense. MEL larvae develop in well insulated crypts that are less
+sensitive evaporative processes than surface water. The effect of
+precipitation on MEL larval habitat abundance may be less pronounced
+than other vector species that develop in open, standing water, like
+ditches and containers. Another model assessed whether the temperature
+precipitation in April, May, and June of each year effected the date of
+the first positive pool test (R file: PT3_edacode). The date of the
+first positive pool was identified and assigned a day number from
+January 1st of that year. The predictors in this linear model were the
+average maximum temperature for April, May, and June and the sum of
+precipitation for the same time period. None of these predictors were
+significant.
 
@@
 ## Statistical analysis
-Before any models were assessed, the RSME for the null model was calculated (2.40). This was used as the basis for assessing model improvement.
 
-Analysis began with a Generalized Additive Model (GAM) with predictors that were selected based on the exploratory single variable GAM.
-The variables were chosen if their p=values were <0.05.
-Eight variables met that condition including: the number of hours below 50°F the week before the mosquito collection week (Hrs_Below_50_lag), the number of hours below 50°F the week of collection (Hrs_Below_50_current), the drought severity index for East Coast FEMA regions (DSCI_fema), the average hourly relative humidity the week of collection (Avg_HourlyRelativeHumidity_current), the average hourly windspeed the week of collection (Avg_HourlyWindSpeed_current), the average number of daylight hours per day (average_daylight_hours), the average number of MEL captured by trap (Average_MEL), and the proportion of MEL captured out of the total number of mosquitoes trapped (MEL_Prop).
+Before any models were assessed, the RSME for the null model was
+calculated (2.40). This was used as the basis for assessing model
+improvement.
 
-All models included training (.75) and a testing (.25) dataset that were generated using the random seed 123.
-All analysis can be found in the analysis.qmd file.
-The first model was a GAM with all eight predictors, followed by a Random Forest, LASSO, Generalized Additive Model with Sparsity (GAMS), and a Gradient Boosting Machine (GBM) model.
+Analysis began with a Generalized Additive Model (GAM) with predictors
+that were selected based on the exploratory single variable GAM. The
+variables were chosen if their p=values were \<0.05. Eight variables met
+that condition including: the number of hours below 50°F the week before
+the mosquito collection week (Hrs_Below_50_lag), the number of hours
+below 50°F the week of collection (Hrs_Below_50_current), the drought
+severity index for East Coast FEMA regions (DSCI_fema), the average
+hourly relative humidity the week of collection
+(Avg_HourlyRelativeHumidity_current), the average hourly windspeed the
+week of collection (Avg_HourlyWindSpeed_current), the average number of
+daylight hours per day (average_daylight_hours), the average number of
+MEL captured by trap (Average_MEL), and the proportion of MEL captured
+out of the total number of mosquitoes trapped (MEL_Prop).
 
-As expected, the RSME is lower for the training dataset than the testing dataset for all models. The RMSE values are displayed in the table below.
+All models included training (.75) and a testing (.25) dataset that were
+generated using the random seed 123. All analysis can be found in the
+analysis.qmd file. The first model was a GAM with all eight predictors,
+followed by a Random Forest, LASSO, Generalized Additive Model with
+Sparsity (GAMS), and a Gradient Boosting Machine (GBM) model.
 
-```{r}
-# Define the file path using here()
-file_path_rds <- here("results", "tables", "rmse_table.rds")
+As expected, the RSME is lower for the training dataset than the testing
+dataset for all models. The RMSE values are displayed in the table
+below.
 
-# Read the RDS file
-rmse_table <- readRDS(file_path_rds)
-rmse_table
-```
+![](images/rmse_table-01.png)
+
 # Discussion
-The null model RMSE was 2.40.
-All models showed improvement over the null model on both the training and testing datasets.
-The GAM shows significant improvement over the null model for the training set, but only slightly improves over the null for the testing set.
-This suggests that beyond capturing the mean response, this GAM is not well generalized.
-The substantial improvement in the training set RSME for GBM and Random Forest coupled with less improvement over the null for the testing set, suggest a tendency for overfitting.
-These models may be capturing the training data's patterns that do not gerneralize well to unseen data.
-The the gap between the training and testing RMSE for the LASSO and GAMS models suggest there is less overfitting and are slightly better at generalizing than the others.
-However, given the non-linear relationships, the LASSO model is likely not appropriate for this analysis.
+
+The null model RMSE was 2.40. All models showed improvement over the
+null model on both the training and testing datasets. The GAM shows
+significant improvement over the null model for the training set, but
+only slightly improves over the null for the testing set. This suggests
+that beyond capturing the mean response, this GAM is not well
+generalized. The substantial improvement in the training set RSME for
+GBM and Random Forest coupled with less improvement over the null for
+the testing set, suggest a tendency for overfitting. These models may be
+capturing the training data's patterns that do not gerneralize well to
+unseen data. The the gap between the training and testing RMSE for the
+LASSO and GAMS models suggest there is less overfitting and are slightly
+better at generalizing than the others. However, given the non-linear
+relationships, the LASSO model is likely not appropriate for this
+analysis.
 
@@
 
-The accuracy
-and reliability of the mosquito collection data is excellent. The same
-procedures and sentinel sites have were in place for the entire
-analytical horizon. Further, there were only two people responsible for
-sorting, identifying and counting species over the entire period.
-However, the data is not without limitations.
-Mosquito surveillance is inherently biased.
-Traps are set in locations that are most likely to capture infected mosquitoes.
-While this is a positive for early warning of virus activity, it is not necessarily representative of the surrounding area.
-Probably the biggest limitation is the small number of observations.
-The mosquito season is short in Massachusetts and funding for trapping and testing mosquitoes is limited.
-Mosquito control efforts, like pesticide applications, are missing from this analysis.
-These efforts are designed to lower the levels of virus transmitting mosquitoes and would likely provide more insight into the relationship between the predictors and virus activity.
+The accuracy and reliability of the mosquito collection data is
+excellent. The same procedures and sentinel sites have were in place for
+the entire analytical horizon. Further, there were only two people
+responsible for sorting, identifying and counting species over the
+entire period. However, the data is not without limitations. Mosquito
+surveillance is inherently biased. Traps are set in locations that are
+most likely to capture infected mosquitoes. While this is a positive for
+early warning of virus activity, it is not necessarily representative of
+the surrounding area. Probably the biggest limitation is the small
+number of observations. The mosquito season is short in Massachusetts
+and funding for trapping and testing mosquitoes is limited. Mosquito
+control efforts, like pesticide applications, are missing from this
+analysis. These efforts are designed to lower the levels of virus
+transmitting mosquitoes and would likely provide more insight into the
+relationship between the predictors and virus activity.
 
@@
 a correction factor for variable pool numbers and I will explore this
-option in later phases of this project.
-Ideally, the outcome would be human or animal cases.
-However, these cases are rare and modeling them given the limited amount of surveillance data may not be possible.
+option in later phases of this project. Ideally, the outcome would be
+human or animal cases. However, these cases are rare and modeling them
+given the limited amount of surveillance data may not be possible.
 
-Initially, the lack of significance for most temperature and precipitation variable was surprising.
-However, upon closer examination, it makes sense.
-The main enzootic vector of EEE and the one targeted for surveillance, *Culiseta melanura*, develops in well insulated crypts that are less sensitive to ambient air temperature changes.
-The effect of temperature changes on MEL development speed may be less pronounced than other vector species that develop in open, standing water, like ditches and containers.
-Precipitation increases the amount of standing water and increases the chances of larval survival during the mosquito season and also during the winter months for those species that overwinter in the larval stage.
-For EEE, the main enzootic vector, MEL, overwinters as larvae.
+Initially, the lack of significance for most temperature and
+precipitation variable was surprising. However, upon closer examination,
+it makes sense. The main enzootic vector of EEE and the one targeted for
+surveillance, *Culiseta melanura*, develops in well insulated crypts
+that are less sensitive to ambient air temperature changes. The effect
+of temperature changes on MEL development speed may be less pronounced
+than other vector species that develop in open, standing water, like
+ditches and containers. Precipitation increases the amount of standing
+water and increases the chances of larval survival during the mosquito
+season and also during the winter months for those species that
+overwinter in the larval stage. For EEE, the main enzootic vector, MEL,
+overwinters as larvae.
 
@@
 
-Likely the biggest limitation 
-## Conclusions 
-While all models demonstrated improvement over null model, none of them were robust enough to be used for reliable prediction.
-In the next phases of this project, the data will be expanded to include more years and cover a larger portion of Massachusetts.
-In additon to increasing the the number of observations, other sources of data will be incorporated.
-Including virus activity from surrounding regions and areas from which birds migrate to Massachusetts.
+Likely the biggest limitation \## Conclusions While all models
+demonstrated improvement over null model, none of them were robust
+enough to be used for reliable prediction. In the next phases of this
+project, the data will be expanded to include more years and cover a
+larger portion of Massachusetts. In additon to increasing the the number
+of observations, other sources of data will be incorporated. Including
+virus activity from surrounding regions and areas from which birds
+migrate to Massachusetts.
 

### 2024-04-29 R/analysis-code/analysis.qmd
Debug and clean
debugged code and updated README files

@@
 # Load the necessary libraries for the analysis
-
 ```{r}
 library(here)
-library(dplyr)
-library(lubridate)
-library(readr)
-library(stargazer)
-library(mgcv)
-library(tidyr)
-library(caret)
-library(randomForest)
-library(tidymodels)
-library(ranger)
-library(dials)
-library(vip)
-library(ggplot2)
-library(boot)
-library(yardstick)
-library(glmnet) # For LASSO
-library(dials)
-library(mlr3)
-library(data.table)
-library(Matrix)
-library(gbm)
-library(knitr)
-library(gt)
+library(dplyr)         # for data manipulation tasks like filtering, mutating, summarizing, and arranging
+library(lubridate)     # for working with dates and times, offering functions to parse, manipulate, and format date-time objects
+library(readr)         # for reading and writing rectangular data in text format, providing faster and more consistent reading of delimited files
+library(stargazer)     # for creating LaTeX and ASCII tables for descriptive statistics and regression models
+library(mgcv)          # for generalized additive models (GAMs) and other generalized ridge regression models
+library(tidyr)         # for tidying and reshaping data for analysis and visualization
+library(caret)         # for machine learning tasks like model training, tuning, and evaluation
+library(randomForest)  # for building random forest models, a popular machine learning algorithm
+library(tidymodels)    # for modeling and machine learning tasks using tidy principles
+library(ranger)        # for fast, robust, and scalable random forest algorithms
+library(dials)         # for tuning hyperparameters in models
+library(vip)           # for variable importance assessment in predictive models
+library(ggplot2)       # for creating elegant and customizable visualizations with the grammar of graphics
+library(boot)          # for bootstrap resampling and estimating standard errors and confidence intervals
+library(yardstick)     # for evaluating predictive models using performance metrics
+library(glmnet)        # for fitting generalized linear models with LASSO or ridge regularization
+library(mlr3)          # for building, evaluating, and comparing machine learning models
+library(data.table)    # for efficient data manipulation and aggregation operations
+library(Matrix)        # for sparse and dense matrix operations
+library(gbm)           # for gradient boosting machines, a powerful machine learning algorithm
+library(knitr)         # for dynamic report generation, integrating R code and output into documents
+library(gt)            # for creating beautiful and customizable tables in R Markdown and other formats
+library(xgboost)       # for extreme gradient boosting, a scalable and accurate machine learning algorithm)
+
 ```
@@
 
-```{r}
+#### This script starts by loading essential R libraries necessary for data manipulation, reading, and date handling, ensuring a smooth setup for data analysis.
+It then loads PCR test results and weather data, providing foundational information on mosquito test results and environmental conditions to study patterns of disease spread and mosquito activity.
+Finally, it calculates average maximum temperatures for key months before the mosquito breeding season, linking climatic conditions to mosquito behavior and potential disease transmission risks.
+```{r}
+# Step 1: load the data
 pcr <- read_csv(here("data", "processed-data", "pcr_data.csv"))
@@
 
-# tuning_grid <- grid_regular(
-#   mtry(range = c(1, length(predictors))),
-#   min_n(range = c(1, 21)),
-#   levels = 7
-# )
-
 # Define cross-validation folds
@@
 
+boot_fit_and_evaluate <- function(data, indices) {
+  # Subset the bootstrap sample
+  boot_sample <- data[indices, ]
+
+  # Predict on the bootstrap sample using the final fitted model
+  predictions <- predict(final_fit_continuous, new_data = boot_sample)
+
+  # Ensure predictions are in the correct format, typically need a vector
+  # If predictions are not a vector, adjust accordingly
+  if (is.data.frame(predictions)) {
+    predictions <- predictions$.pred  # Adjust this based on your predict function output structure
+  }
+
+  # Calculate the RMSE of predictions
+  # Ensure both truth and predictions are either integers or doubles
+  results <- tibble(
+    truth = as.numeric(boot_sample$MIR),  # Convert to numeric to ensure matching types
+    estimate = as.numeric(predictions)  # Convert to numeric to match type
+  )
+
+  # Compute RMSE and return
+  result <- rmse(results, truth = truth, estimate = estimate)
+  return(result$.estimate)
+}
 # Continue with bootstrapping to evaluate model's robustness
@@
 )
-
 # Calculate the bootstrapped RMSE
@@
 
-autoplot(tuning_results_continuous)
+plot <- autoplot(tuning_results_continuous)
+
+# Save the plot to the specified directory
+ggsave(filename = here("results", "figures", "RF_autoplot.png"), plot = plot, width = 10, height = 8)
 ```
@@
 
-autoplot(tuning_results_continuous)
+plotGBM <- autoplot(tuning_results_continuous)
+
+#save plot as a png
+ggsave(filename = here("results", "figures", "GBM_autoplot.png"), plot = plotGBM, width = 10, height = 8)
 

### 2024-04-29 R/eda-code/EDA_unified.qmd
Debug and clean
debugged code and updated README files

@@
 ```{r}
-library(here) #for data loading/saving
-library(dplyr)
-library(skimr)
-library(ggplot2)
-library(tidyverse)
-library(lubridate)
-library(readr)
-library(forecast)
-library(tsibble) # For time series data wrangling
-library(feasts) # For decomposing and visualizing time series
-library(RColorBrewer)
-library(ggcorrplot)
-library(readxl)
-library(corrplot)
-library(plotly)
+library(here)        # for data loading/saving
+library(dplyr)       # for data manipulation
+library(skimr)       # for data summarization
+library(ggplot2)     # for data visualization
+library(lubridate)   # for date and time manipulation
+library(forecast)    # for time series forecasting
+library(tsibble)     # for time series data wrangling
+library(feasts)      # for decomposing and visualizing time series
+library(RColorBrewer) # for color palettes
+library(ggcorrplot)  # for correlation plot
+library(readxl)      # for reading Excel files
+library(corrplot)    # for correlation plot
+library(plotly)      # for interactive plots
+library(tidyr)       # for data reshaping
 library(reshape2)
-library(rlang)
 ```
@@
 average_daylight_by_week <- read_csv(here("data", "processed-data", "average_daylight_by_week.csv"))
-average_mel_weekco2 <- read_csv(here("data", "processed-data", "average_mel_weekco2.csv"))
+average_mel_weekco2 <- read_csv(here("data", "processed-data", "average_mel_week_co2.csv"))
 wx_hourly_clean <- read_csv(here("data", "processed-data", "hourly_wx_clean.csv"))

### 2024-04-29 R/processing-code/MADAproject_pt2_ruiz.qmd
Debug and clean
debugged code and updated README files

@@
 #### The dataset was obtained though and open records request with the Commonwealth of MA.
+##### we will read the data now and process it later in this script
 ```{r trap-data-import}
@@
 saveRDS(trap_location_selected, trap_location_selected_path_rds)
+
+#save as csv
+trap_location_selected_path_csv <- here("data", "processed-data", "trap_sites_selected.csv")
+write_csv(trap_location_selected, trap_location_selected_path_csv)
+
 ```
@@
 #Define the path to the folder with Excel files
-folder_path <- here("data", "processed-data", "virus_data")
+folder_path <- here("data", "raw-data", "virus_data")
 
@@
 ```{r}
+
 # Construct the path to the file
-file_path_wx = here("data", "processed-data", "weather_airport.csv")
+file_path_wx <- here("data", "raw-data", "weather_airport.csv")
 
-# import the data to R and store it in a dataframe named "wx"
-wx = read.csv(file_path_wx)
+# Import the data to R and store it in a dataframe named "wx"
+wx <- read.csv(file_path_wx)
 
+# Check the structure of the DATE column to determine if there are any unexpected values
+str(wx$DATE)
+
+# Check for any missing values in the DATE column
+sum(is.na(wx$DATE))
+
+# If there are no missing values, proceed with the mutation
 wx <- wx %>%
   mutate(
-    DATE = mdy(DATE),  # Parse the DATE column from "month/day/year"
+    DATE = ymd(DATE),  # Parse the DATE column from "year-month-day"
     week_num = paste0(year(DATE), "-", sprintf("%02d", isoweek(DATE)))  # Create week_num
   )
-#load the first few rows of data to verify the data loaded as expected
+
+# Load the first few rows of data to verify the data loaded as expected
 head(wx)
@@
 Missing data in that variable generally means 0 mosquitoes were captured in a trap.
-The other variable are important and so we will remove those records with the other missing data.
+We will remove the rows with missing data in the other columns in th code chunks below.
 
@@
 The change divided the genus *Aedes* into a new genus: *Ochlerotatus*.
-To align with accepted standards, I will change them back to *Aedes*.
+To align with current accepted standards, I will change them back to *Aedes*.
 
@@
 
-### Execute proposed changes.
+### Execute proposed from above
 
@@
 ```{r}
-library(dplyr)
-library(tidyr)
-library(purrr)
-library(lubridate)
 
@@
 ```{r annual precip}
-# Convert DATE from character to Date type using as.Date and specifying the format
-wx$DATE <- as.Date(wx$DATE, format="%m/%d/%y")
+# Construct the path to the file
+file_path_wx <- here("data", "processed-data", "wx_selected.csv")
 
-# The as.Date function treats years in the format "yy" as follows:
-# Years 00-68 are treated as 2000-2068, and years 69-99 are treated as 1969-1999.
-# Adjust this according to your data's actual year range if needed.
+# Import the data to R and store it in a dataframe named "wx"
+wx <- read.csv(file_path_wx)
 
-# Extract the year from each DATE
-wx$Year <- format(wx$DATE, "%Y") %>% as.numeric()
+# Extract the year from each DATE and store it in a new column named "Year"
+wx$Year <- as.integer(format(as.Date(wx$DATE), "%Y"))
 
@@
   filter(Year >= 2007 & Year <= 2020)
+
+# View the filtered dataframe
+head(filtered_wx)
 

### 2024-04-29 data_dictionary/data_dict_search_tool.qmd
Debug and clean
debugged code and updated README files

---
title: "Data Dictionary Search Tool"
output: html_document
runtime: shiny
editor: visual
---

# Data dictionary search tool

## Use this tool to search the data dictionary by file name, column name, or both.

### First code block was used to compile the data dictionary from multiple CSV files. The data dictionary is then saved as a CSV file named "Data_Dictionary.csv".

```{r echo=FALSE}
library(shiny)
library(here)
library(dplyr)
library(readr)
```

## Run this code this code to open the search tool.

### you can also view the entire data dictionary here: (.../data_dictionary/Data_Dictionary.csv)

```{r echo=FALSE}
# Load your data dictionary
data_dictionary <- read_csv(here("data_dictionary", "Data_Dictionary.csv"))

# Define UI
ui <- fluidPage(
    titlePanel("Data Dictionary Search Tool"),
    sidebarLayout(
        sidebarPanel(
            helpText("Use the fields below to search the data dictionary by file name, column name, or both."),
            textInput("file_name", "Enter File Name:", ""),
            textInput("column_name", "Enter Column Name:", ""),
            helpText("Press search to update the results."),
            actionButton("search", "Search"),
            helpText("Tip: The search is case-insensitive and partial matches will also show up in the results.")
        ),
        mainPanel(
            dataTableOutput("results")
        )
    )
)

# Define server logic
server <- function(input, output) {
    # Reactive value for filtered data
    filtered_data <- reactive({
        req(input$search)  # Make sure the search button has been pressed
        data_dictionary %>%
            filter(
                (input$file_name == "" | grepl(input$file_name, FileName, ignore.case = TRUE)) &
                (input$column_name == "" | grepl(input$column_name, ColumnName, ignore.case = TRUE))
            )
    })

    # Output the results
    output$results <- renderDataTable({
        filtered_data()
    })
}

# Run the application
shinyApp(ui = ui, server = server)

```

### 2024-04-29 peer_reviews/CMiller-project-review-Ruiz.qmd
Debug and clean
debugged code and updated README files

--- 2024-04-26 CMiller-project-review-Ruiz.qmd
@@
 You have a couple of things to polish off, but, wow.
-You obviosuly have the topic covered, you have the coding skills, you got this.
+You obviously have the topic covered, you have the coding skills, you got this.
 
@@
 
-*Were the analysis methods appropriate for the data?
-Was the analysis done properly?
-Were different components of the analysis (e.g.
-performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?*
+*Were the analysis methods appropriate for the data? Was the analysis done properly?*
 
-At this point, this section of the document is incomplete.
+Yes, I believe that they were.  I am not an expert on this topic, but the analysis looked reasonable and properly done.
+
+*Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?*
+
+Again, I am not an expert on this topic.  The analysis looked appropriate for the data type.  Yes, the data and analysis were explained well.
 
@@
 
-At this point, this section of the document is incomplete.  That being said, the description of the intended analysis is appropriate.
+No real feedback on this part.  You did a great job in this category.  The description of the intended analysis is appropriate.
 
@@
 
-There are no results at this time.
+The results were presented in a table format.
+The results compared several types of model for the desired data.
+The table included RMSE data for training and test data.
+No figures of the analysis were made.
 
@@
 
-At this point, this section of the document is incomplete.
+No real feedback on this part.  You did a great job in this category.
 
@@
 
-* results are poorly presented, hard to understand, poor quality
+* results are presented ok, with room for improvement
 
-At this point, this section of the document is incomplete.
+At this point, this section of the document seems incomplete.
+I can see a need for tables describing the output discussed in the basic data analysis section.
+Plus, maybe some descriptive figures of the analysis if possible?
 
@@
 
-At this point, this section of the document is incomplete.
+Yes.  The conclusions are discussed, the strengths and limitations of the data and analyses are discussed.
 
@@
 
-At this point, this section of the document is incomplete.
+No real feedback on this section.  The section looked generally complete.
 
 ### Summary assessment
-* major parts of discussion missing or wrong
 
-At this point, this section of the document is incomplete.
+* strong, complete and clear discussion
 
@@
 
-A large portion of this work is missing.  It is my understanding that the author is working through these sections again as new data (rainfall/weather data) was made avaiable.
-
-The basic bones of the project look good.
+The basic bones of the project look good.  It is obvious that a good deal of work has gone into the paper.
 
@@
 
-Not at this time.
+Not at this time.  There are still a lot of extra files in the repo.
 
@@
 
-At this point, this section of the document is incomplete.
+I think that no project can be documented well enough.
+This one needs a bit more work in that area.
+There are still a a lot of readme files talking about how the project will look when it's finished.
+Of course, that might just be github acting up again.
 
@@
 
-Yes.  As far as it goes.  At this point, this section of the document is incomplete.
+Yes.  Everything looks complete and it runs.
 
@@
 
-At this point, this section of the document is incomplete.
+I like how you piled all the code into a single R folder.
+It took a minute for me to find it because it wasn't the format that my other review used.
+I think that is a cleaner way to organize the coding sections.
 
@@
 
-Once complete, the described work will be very complete.  At this point this section is incomplete.
+The overall study looks very thorough and complete.  Again, I am not an expert in this area, but everything looked reasonable and well thought out.  All the salient areas seemed to be addressed.
 
@@
 
-At this point, this section of the document is incomplete.
+No real feedback on this section.  The section looked generally complete.
 
@@
 
-Just finish this project off the way you've described and it will be great.  Hang in there!
+Just finish this project off the way you've described and it will be great!

### 2024-04-29 R/analysis-code/archived files/2partGAMmodel.R

@@ -1,114 +1,0 @@
-library()
-###Two Part Model###
-mir_2part <- read_csv(here("data", "processed-data", "joined_data_mir_co2.csv"))
-
-library(dplyr)
-str(mir_2part)
-# Creating a binary variable for zero vs. positive MIR
-mir_2part <- mir_2part %>%
-  mutate(MIR_Positive = as.integer(MIR > 0))
-
-# Fitting the logistic regression model
-logistic_model <- glm(MIR_Positive ~ Avg_HourlyRelativeHumidity_current + Avg_HourlyWindSpeed_current + Hrs_Below_50_current,
-                      family = binomial(link = "logit"),
-                      data = mir_2part)
-
-summary(logistic_model)
-
-
-# Filtering the dataset for positive MIR values
-positive_mir_data <- mir_2part %>% 
-  filter(MIR > 0)
-
-# Fitting the linear regression model to the positive MIR values
-linear_model <- lm(MIR ~ Avg_HourlyRelativeHumidity + Avg_TMAX + DSCI_ma + Avg_TMAX,
-                   data = positive_mir_data)
-
-summary(linear_model)
-
-# Plot the diagnostic plots for the linear model
-par(mfrow = c(2, 2))
-plot(linear_model)
-### These suggest that the issues of non-linearity, heteroscedasticity, and influential outliers need to be addressed using different models
-
-# For this we will use a two-part model, which consists of a GAM logistic regression model to predict the probability of observing a positive MIR value,
-
-library(mgcv)
-# Fit a GAM with a smoothing function for each predictor
-gam_model <- gam(MIR ~ s(Avg_TMAX),
-                 data = mir_2part, 
-                 family = gaussian(),
-                 method = "REML")
-
-# Check the summary for detailed output
-summary(gam_model)
-
-# Plot the smooths to interpret the effect of each predictor
-plot(gam_model)
-
-
-
-##2 stage GAM
-
-zero_gam_model <- gam(I(MIR > 0) ~ s(DSCI_ma) + s(Average_MEL) + s(Avg_HourlyWindSpeed_current),
-                      family = binomial(link = "logit"), 
-                      data = mir_2part)
-
-summary(zero_gam_model)
-plot(zero_gam_model)
-
-names(mir_2part)
-positive_gam_model <- gam(MIR ~ s(Hrs_Between_50_86_lag), 
-                          data = mir_2part[mir_2part$MIR > 0, ], 
-                          family = gaussian())
-# Check the summary for detailed output
[…]

### 2024-04-29 R/analysis-code/archived files/Annual_models.R

@@ -1,2 +1,0 @@
-####Annual Model####
-

### 2024-04-29 data_dictionary/data_dictionary.R

@@ -1,0 +1,28 @@
+library(here)
+library(readr)
+library(dplyr)
+library(purrr)
+
+# Define the function to get column details
+get_column_details <- function(file) {
+  data <- read_csv(file, guess_max = 1000, show_col_types = FALSE)
+  tibble(
+    FileName = basename(file),
+    ColumnName = names(data),
+    DataType = sapply(data, function(x) class(x)[1]), # Ensures consistent data type as character
+    SampleValue = sapply(data, function(column) toString(column[1])) # First non-NA value
+  )
+}
+
+# Set the path to the directory containing the CSV files
+path_to_files <- here("data", "processed-data")
+
+# List all CSV files in the directory but exclude the "rds" folder
+file_list <- list.files(path = path_to_files, pattern = "\\.csv$", full.names = TRUE)
+file_list <- file_list[!grepl("/rds/", file_list)] # Excludes files in "rds" subfolder
+
+# Apply the function to each file and combine the results
+data_dictionary <- map_df(file_list, get_column_details)
+
+# Save the data dictionary to a CSV file
+write_csv(data_dictionary, here("data", "processed-data", "Data_Dictionary.csv"))

### 2024-04-30 R/analysis-code/analysis.qmd
edits to readme and mauscript

@@
 
+# Extract the fit using the appropriate tidymodels function
+fitted_model <- extract_fit_engine(final_fit_continuous)
+
+# Check if the extracted model is the correct type
+if(inherits(fitted_model, "ranger")) {
+    # If it is the correct type, extract feature importance
+    importance_matrix <- fitted_model$variable.importance
+    feature_importances <- tibble(
+        Feature = names(importance_matrix),
+        Importance = importance_matrix
+    ) %>%
+    arrange(desc(Importance))
+
+    # Print the feature importances
+    print(feature_importances)
+
+    # Plot the feature importances
+    importance_plot <- ggplot(feature_importances, aes(x = reorder(Feature, Importance), y = Importance)) +
+        geom_col(fill = "steelblue") +
+        coord_flip() +
+        labs(title = "Feature Importance from Random Forest", x = "Importance", y = "Features") +
+        theme_minimal()
+
+    # Save the plot
+    ggsave("feature_importance_rf.pdf", plot = importance_plot, width = 8, height = 6)
+} else {
+    cat("The extracted object is not a 'ranger' Random Forest model. Please check the model setup and extraction.\n")
+}
+print(importance_plot)
+
 ### Calculate RMSE for the training and testing sets

### 2024-04-30 products/manuscript/Manuscript.qmd
Debug and clean
debugged code and updated README files
edits to readme and mauscript

@@
 
-# Problem statement
+# Research question
 
 Which environmental,ecological, and temporal factors influence the
-transmission cycles for EEE in southeastern Massachusetts and which
+transmission cycles for EEE in southeastern Massachusetts, and which
 models best describe the relationship between these factors?
@@
 
-Bristol County, Massachusetts covers an area of 691 square miles with a
-population density of 962 people per square mile. It borders Rhode
-Island to the west, Plymouth County, MA to the east, and Norfolk County,
-MA to the north. Naushon Island and Martha's Vineyard are across
-Buzzard's Bay to the Southeast.
+Bristol County, Massachusetts covers 691 square miles with a population
+density of 962 people per square mile. It borders Rhode Island to the
+west, Plymouth County, MA to the east, and Norfolk County, MA to the
+north. Naushon Island and Martha's Vineyard are across Buzzard's Bay to
+the Southeast.
 
-Atlantic White Cedar (AWC) swamps span over 4,300 acres within
-Massachusetts, primarily concentrated in Plymouth and Bristol counties.
-These swamps serve as critical breeding habitats for *Culiseta melanura*
-(MEL)
-[@commonwealthofmassachusettsdivisionoffisheries-wildlifeNaturalCommunityFact2007],
-which is recognized as the primary enzootic vector for Eastern Equine
-Encephalitis (EEE). AWC trees have large root systems that form "crypts"
-and hold water throughout the year. These crypts serve as a main habitat
-for MEL larval growth and provide shelter during winter months.
+Atlantic White Cedar (AWC) swamps cover more than 4,300 acres across
+Massachusetts, with the largest concentrations found in Plymouth and
+Bristol counties. These swamps are crucial breeding grounds for Culiseta
+melanura (MEL), which is identified as the primary enzootic vector for
+Eastern Equine Encephalitis (EEE)
+[@commonwealthofmassachusettsdivisionoffisheries-wildlifeNaturalCommunityFact2007].
+Atlantic White Cedar (AWC) trees possess extensive root systems that
+create "crypts" capable of retaining water year-round. These crypts are
+crucial habitats for the larval growth of Culiseta melanura (MEL) and
+also provide shelter during the winter months.
 
@@
 **Which environmental,ecological, and temporal factors influence the
-transmission cycles for EEE in southeastern Massachusetts and which
+transmission cycles for EEE in southeastern Massachusetts, and which
 models best describe the relationship between these factors?** From
@@
 
+With the exception of the mosquito collection data, all other datasets
+are publicly available for use without restrictions.
+
 ### Mosquito surveillance data
@@
 raw dataset includes over 45,000 records and includes temporal and
-spatial variable. The BCMCP, perform mosquito surveillance from June
+spatial variables. The BCMCP, perform mosquito surveillance from June
 through October annually.
 
-During this interval, CDC light traps baited with carbon dioxide are
-placed at various sentinel sites each week. Designed to attract
-mosquitoes in search of blood meals, light traps serve as a reliable
-measure of the actively feeding mosquito population, thereby
-establishing them as the standard method for EEE surveillance. Some of
-the sentinel site have been used for over 50 years.
+During this period, CDC light traps, baited with carbon dioxide, are
+strategically placed at various sentinel sites weekly. These traps are
+designed to attract mosquitoes seeking blood meals and are recognized as
+a reliable method for monitoring the actively feeding mosquito
+population. Consequently, they are established as the standard approach
+for Eastern Equine Encephalitis (EEE) surveillance. Notably, some of
+these sentinel sites have been in continuous use for over 50 years.
 
-Additionally, the county employs gravid traps, which are specifically
+Additionally, the county deploys gravid traps, which are specifically
 aimed at attracting and capturing gravid *Culex* mosquitoes looking for
@@
 
-The traps are installed in the afternoon and retrieved the following
-morning. After the traps are collected, the next are stored in a freezer
-to preserve and kill the mosquitoes. The mosquitoes are sorted by
-species using a dissecting scope. Counts for each species are recorded
-in the database and a portion of the vector species are driven to the
-Massachusetts State Laboratory in Boston. The lab tests each mosquito
-pool (or batch) for WNV and EEEv using PCR. Results are shared with
-BCMCP on the same day of testing. Virus isolation results are recorded
-in the collection records.
+The traps are set up in the afternoon and collected the following
+morning. Once retrieved, the captured mosquitoes are immediately stored
+on dry ice and then in a freezer. This process kills the mosquito and
+preserves them. The mosquitoes are sorted by species under a dissecting
+scope. Counts for each species are recorded in the database and a
+portion of the vector species are driven to the Massachusetts State
+Laboratory in Boston. The lab tests each mosquito pool (or batch) for
+WNV and EEEv using PCR. Results are shared with BCMCP on the same day of
+testing. Virus isolation results are recorded in the collection records.
 
@@
 
-One indicator of the level of virus activity present in the mosquito
-population can be calculated from the information contained in the
-mosquito dataset. The minimum infection rate (MIR) can be calculated at
-various temporal and spatial scales from trap site over one day to all
-sites over an entire season. MIR is calculated as: $$
+In our investigation into the environmental, ecological, and temporal
+factors that impact the transmission cycles of EEE in southeastern
+Massachusetts, the Mosquito Infection Rate (MIR) emerges as a pivotal
+indicator. MIR provides a quantifiable measure of virus prevalence
+within mosquito populations, making it an essential metric for modeling
+the dynamics of disease spread. By analyzing changes in MIR in response
+to varying environmental conditions and mosquito behaviors, we can
+enhance our understanding of the key drivers behind EEE transmission.
+This understanding, in turn, guides the development of predictive models
+that can accurately forecast periods of heightened transmission risk
+based on observed or anticipated changes in the identified factors. The
+minimum infection rate (MIR) can be calculated at various temporal and
+spatial scales from trap site over one day to all sites over an entire
+season. MIR is calculated as: $$
 MIR = \left( \frac{\text{Number of positive mosquito pools}}{\text{Total number of mosquitoes tested}} \right) \times 1000
-$$ MIR provides an estimate of the prevalence of a specific pathogen
-within a mosquito population. While it does not directly measure the
-risk to humans, a higher MIR in mosquitoes that commonly bite humans can
-indicate a higher risk of transmission. A study pubulihsed in 2007
-indicated that in years with human cases in Massachusetts between 1979
-and 2004, MIR was ≥1.0 by week 31 [@hachiyaHumanEasternEquine2007]. MIR
-allows for the comparison of infection rates across different
-geographical areas or time periods. This is crucial for identifying
-areas or times of higher risk and for evaluating the effectiveness of
-control measures.
+$$ While MIR does not directly measure the risk to humans, a higher MIR
+in mosquitoes that commonly bite humans can indicate a higher risk of
+transmission. A study pubulihsed in 2007 indicated that in years with
+human cases in Massachusetts between 1979 and 2004, the MIR was ≥1.0 by
+week 31 (mid-late July) [@hachiyaHumanEasternEquine2007]. MIR allows for
+the comparison of infection rates across different geographical areas or
+time periods. This is crucial for identifying areas or times of higher
+risk and for evaluating the effectiveness of control measures.
 
@@
 variables, such as the number of hours above a certain temperature
-threshold or degree-days. Degree-days is a measure of the amount of heat
-accumulated over a period of time. It is calculated by summing the
+threshold. or degree-days. Degree-days is a measure of the amount of
+heat accumulated over a period of time. It is calculated by summing the
 difference between the average daily temperature and a base temperature.
@@
 
-## Data aquisition
-
-With the exception of the mosquito collection data, all other datasets
-are publicly available for use without restrictions.
-
 ## Data import and cleaning
@@
 
-The Massachusetts Department of Public Health requires that all mosquito
-control districts submit their collection data in a specific format.
-This format divides each trap event into separate rows based on mosquito
+The Massachusetts Department of Public Health mandates a specific format
+for mosquito control districts to submit their collection data. In this
+format, each trap event is divided into separate rows based on mosquito
 species and whether they were submitted for virus testing. However, this
-format is not useful for certain calculations where a zero count for a
-species is not explicitly recorded. For example, even when sorted by
-species to calculate the average count per trap, the average could be
-inflated since the records will not include trap events with a zero
-count of that species. In order to correct this, the data was pivoted to
-create a column for every unique "species code" then the "pool size" for
-each species code is summed for each trap event. A trap event is when
-"town", "date of collection", "trap type" are all the same. For species
-codes not included in the orginal table for a trap event, the field will
-be assigned a zero value. The resulting table will have one row for
-every trap event.
+format poses limitations for certain calculations where a zero count for
+a species is not explicitly recorded. For instance, when sorting data by
+species to calculate the average count per trap, the absence of records
+with a zero count could inflate the calculated average.
+
+To address this issue, the data was transformed by pivoting it to create
+a column for every unique "species code." Then, the "pool size" for each
+species code was summed for every trap event, defined as instances where
+"town," "date of collection," and "trap type" are identical. For species
+codes not originally included in the table for a trap event, the
+corresponding field was assigned a zero value. As a result, the
+resulting table now contains one row for each trap event.
 
@@
 
-Symptomatic EEE, and to a lesser extent WNV, human cases are rare. On
-average, the United States records only about 9 cases of EEE annually.
-Notably, Massachusetts reports more cases than any other state except
-Florida, highlighting regional disparities in disease occurrence.
-However, the incidence rate for neuroinvasive EEE cases in Bristol
-County, Massachusetts, is estimated to be only around 0.06 per 100,000
-people
+Symptomatic EEE human cases are rare. On average, the United States
+records only about 9 cases of EEE annually. Notably, Massachusetts
+reports more cases than any other state except Florida, highlighting
+regional disparities in disease occurrence. However, the incidence rate
+for neuroinvasive EEE cases in Bristol County, Massachusetts, is
+estimated to be only around 0.06 per 100,000 people
 [@centersfordiseasecontrolandpreventionPreventionEasternEquine2022]. The
@@
 will be calculated for each week during the surveillance season over the
-entire county. To reduce bias, only mosquitoes captured using carbon
-dioxide-baited CDC light traps will be used to to calculate MIR. These
-traps are the standard EEE surveillance and are used to monitor the
-actively feeding mosquito population. Additionally, these trap types
-have been used in this area for over 40 years, many at the same sites
-currently in use.
+entire county.
 
@@
 In the first stage of the analysis, the GAM was utilized to address the
-high incidence of zeros in the dataset—an indicator of overdispersion.
-This zero-inflated model allowed us to separately model the presence
-versus absence of the outcome variable, thereby appropriately accounting
-for the excessive zeros.
+high incidence of zeros in the dataset. This zero-inflated model allowed
+us to separately model the presence versus absence of the outcome
+variable, thereby appropriately accounting for the excessive zeros.
 
-Subsequently, the second stage of the GAM focused on modeling the
-continuous part of the MIR data, relating it to the suite of potential
-predictors. This step incorporated smooth functions of the predictors,
-permitting the data to inform the shape of the relationship without
-imposing a predetermined form, as would be the case in a traditional
-linear model.
+The second stage of the GAM focused on modeling the continuous part of
+the MIR data, relating it to the suite of potential predictors. This
+step incorporated smooth functions of the predictors, permitting the
+data to inform the shape of the relationship without imposing a
+predetermined form, as would be the case in a traditional linear model.
 
-While the two-stage GAM appeared to be a good fit, there were too few
-observations to provide a robust estimate of the relationship between
-the predictors and MIR. The original dataset contained over 40,000
-observations, but after reformatting the data to fit a rectangular shape
-needed for analysis, there were less than 400 observations in the final
-set. After the zeros were removed for part 2 of the GAM, the sample
-number of observations left was less than 100. Because of this, it was
-decided to use a single GAM with a dataset that included zero and
-non-zero observations. Additionally, all trap types were included in the
-analysis, not just the carbon dioxide-baited CDC light traps. While some
-precision was lost when the one part model was adopted, the gain in
-observations was deemed more valuable.
-
-### Schematic of workflow
-
-*Schematic will be refined in later steps. The doc will also be
-reformatted to move the schematic to an appendix* *The image should
-appear upon rendering*
-
-```{mermaid}
-flowchart TB
-    id1["Questions"] --> id2["Data Collection"]
-    id2 --> id3["Data Loading and Cleaning"]
-    id3 --> id4["Data Cleaning"]
-    id4 --> id5["Data Transformation"]
-    id5 --> id6["Data Exploration & Visualization"]
-    id6 --> id7["Data Modeling"]
-    id7 --> id8["Evaluation"]
-
-    id8 -.-> id1
-    id8 --> id9[Interpretation]
-    id9 --> id10[Communication]
-
-    classDef mainProcess fill:#fff,stroke:#333,stroke-width:2px;
-
-    class id1,id2,id3,id4,id5,id6,id7,id8,id9,id10 mainProcess
-```
+While the two-stage Generalized Additive Model (GAM) initially seemed to
+be a suitable approach, there were insufficient observations to yield a
+robust estimate of the relationship between the predictors and Minimum
+Infection Rate (MIR). Despite starting with a dataset containing over
+40,000 observations, the transformation of the data into a rectangular
+shape required for analysis resulted in fewer than 400 observations in
+the final dataset. After removing the zeros for part 2 of the GAM, the
+number of observations decreased to less than 100. Consequently, a
+decision was made to utilize a single GAM with a dataset containing both
+zero and non-zero observations. Although adopting a single-part model
+led to some loss of precision, the increase in the number of
+observations was deemed more advantageous.
 
@@
 model suggests there is a significant inverse relationship. As average
-temperature increases, the MIR decreases (model_mir_temp.R). This
-relationship is counterintutive, especially given the result of the
-logistic regression and requires more examination.
+temperature increases the week prior to collection, the MIR decreases
+(model_mir_temp.R). This relationship is counterintutive, especially
+given the result of the logistic regression and requires more
+examination.
 
 ## Statistical analysis
+
+All detailed model analyses are available in the analysis.qmd file,
+providing a transparent account of the methodologies employed and the
+statistical rigor behind the conclusions drawn.
 
@@
 hourly relative humidity the week of collection
-(Avg_HourlyRelativeHumidity_current), the average hourly windspeed the
+(Avg_HourlyRelativeHumidity_current), the average hourly wind speed the
 week of collection (Avg_HourlyWindSpeed_current), the average number of
@@
 
-![](images/rmse_table-01.png)
+![](images/rmse_table-03.png){fig-align="center"}
 
@@
 
-The null model RMSE was 2.40. All models showed improvement over the
-null model on both the training and testing datasets. The GAM shows
-significant improvement over the null model for the training set, but
-only slightly improves over the null for the testing set. This suggests
-that beyond capturing the mean response, this GAM is not well
-generalized. The substantial improvement in the training set RSME for
-GBM and Random Forest coupled with less improvement over the null for
-the testing set, suggest a tendency for overfitting. These models may be
-capturing the training data's patterns that do not gerneralize well to
-unseen data. The the gap between the training and testing RMSE for the
-LASSO and GAMS models suggest there is less overfitting and are slightly
-better at generalizing than the others. However, given the non-linear
-relationships, the LASSO model is likely not appropriate for this
-analysis.
+The Root Mean Square Error (RMSE) comparison outlined in the table
+provides a clear perspective on the efficacy of different predictive
+models in the context of Eastern Equine Encephalitis (EEE) surveillance
+in southeastern Massachusetts. The null model, which served as a
+baseline with an RMSE of 2.40, indicates the level of error absent any
+predictive influence. This baseline was pivotal for evaluating
+enhancements brought by more complex models.
+
+The Generalized Additive Model (GAM), which incorporated eight predictors selected based on their statistical significance from preliminary analyses, demonstrated concerning signs of overfitting.
+While the model achieved a low RMSE of 1.7 on the training data, indicating a good fit to this dataset, its performance on the testing data was substantially worse, with an RMSE of 3.44.
+This significant increase in RMSE when applied to the testing data, compared to the training data, suggests that the model, despite fitting the training data well, fails to generalize to new, unseen data.
+This discrepancy in performance is further highlighted by the testing RMSE being higher than the RMSE of the null model at 2.40, underscoring the limitations of the GAM in this scenario.
+
+Subsequent models such as the Random Forest, LASSO, Generalized Additive
+Model with Sparsity (GAMS), and Gradient Boosting Machine (GBM) were
+tested to potentially mitigate overfitting and enhance prediction
+accuracy. Notably, the Gradient Boosting Machine model achieved the
+lowest testing RMSE of 2.14, suggesting a model with better
+generalizability compared to others.
+
+Given that the average MIR value is 1.6, an RMSE of over 2 suggests that the models are not accurately capturing the variability in the data and are making predictions that are significantly different from the observed values.
+
+Each model's performance underlines the complex interplay of various
+ecological and environmental factors in predicting EEE transmission
+cycles.
 
@@
 
-Likely the biggest limitation \## Conclusions While all models
+## Conclusions
+
+While all models
 demonstrated improvement over null model, none of them were robust
@@
 of observations, other sources of data will be incorporated. Including
-virus activity from surrounding regions and areas from which birds
-migrate to Massachusetts.
+virus activity from surrounding regions and incorporating bird population data in to the models could provide more insight into the relationship between the predictors and virus activity.
 

### 2024-04-30 products/manuscript/supplement/Supplementary-Material.qmd
edits to readme and mauscript

---
title: "Supplement to Factors Influencing Abundance & Virus Isolation in Mosquitoes"

subtitle: "Schematic of workflow"
author: Andrew Ruiz
date: "2024-04-30"
format:
  pdf:
    toc: false
    number-sections: true
    highlight-style: github
---
#### *The image should appear upon rendering*
```{mermaid width=700 height=500}
flowchart TB
    id1["Research Questions"] --> id2["Data Collection"]
    id2 --> id3["Data Loading"]
    id3 --> id4["Data Cleaning & Preprocessing"]
    id4 --> id5["Data Exploration & Visualization"]
    id5 --> id6["Variable Selection"]
    id6 --> id7["Model Development"]
    id7 --> id7a["Model Training (Random Seed 123)"]
    id7a --> id7b["Model Comparison\n(GAM, RF, LASSO, GAMS, GBM)"]
    id7b --> id8["Model Evaluation\n(RMSE Calculation)"]
    id8 -.-> id1
    id8 --> id9["Result Interpretation"]

    classDef mainProcess fill:#fff,stroke:#333,stroke-width:2px;
    class id1,id2,id3,id4,id5,id6,id7,id7a,id7b,id8,id9 mainProcess
```

### 2024-05-03 R/analysis-code/analysis.qmd
Edits to the manuscript and code cleaning

@@
 library(lubridate)     # for working with dates and times, offering functions to parse, manipulate, and format date-time objects
-library(readr)         # for reading and writing rectangular data in text format, providing faster and more consistent reading of delimited files
+library(readr)         # for reading & writing rectangular data in text format
 library(stargazer)     # for creating LaTeX and ASCII tables for descriptive statistics and regression models
@@
 ## 1 Part Random Forest
+### Load the dataset, define predictors, and prepare the data for modeling.
+```{r}
 ```{r}
@@
 cv_folds_continuous <- vfold_cv(train_data_continuous, v = 5)
-
+```
+### Define the RF model specifications, create recipe, and setup workflow
+```{r}
 # Define the Random Forest model specification
@@
   add_model(rf_model_spec_continuous)
-
-# Tune the model
+```
+
+## Tune the model, extract the best hyperparameters, and fit the final model
+```{r}
+```{r}
 tuning_results_continuous <- tune_grid(
@@
 
+# View the best hyperparameters
+print(best_hps_continuous)
+
 # Fit the final model with the best hyperparameters on the full training dataset
@@
 )
-
+```
+
+### Extract the fitted model, calculate feature importance
+```{r}
 # Extract the fit using the appropriate tidymodels function
@@
 print(importance_plot)
+```
 
 ### Calculate RMSE for the training and testing sets
-
+```{r}
 # Predict on the training and testing set
@@
 cat("Testing Set RMSE:", test_rmse_result_rf$.estimate, "\n")
-
+```
+
+### Bootstrapping to evaluate model robustness
+```{r}
 boot_fit_and_evaluate <- function(data, indices) {
@@
 ### There still seems to be overfitting. The RMSE on the test set is higher than the training set.
+### Additionally, LASSO models are not ideal for capturing non-linear relationships.
 ### We will try a different model
@@
 ```{r}
-library(tidymodels)
-library(mgcv)
-library(readr) # For read_csv
-
 # Load the dataset
@@
 ```
+
 # Gradient Boosting Machine (GBM)
-```{r}
-
-library(tidymodels)
-library(dplyr)
-library(yardstick)
-library(ranger) # For Random Forest
-library(boot) # For bootstrapping
-
+### Gradient Boosting Machine (GBM) is an ensemble learning technique that builds models sequentially, with each new model correcting errors made by the previous ones.
+It uses decision trees as the base learners and optimizes a loss function, making it effective for both regression and classification tasks.
+```{r}
 # Load the dataset
@@
 )
+```
 
 ### Calculate RMSE for the training and testing sets
-
+```{r}
 # Predict on the training and testing set
@@
 cat("Testing Set RMSE:", test_rmse_result_gbm$.estimate, "\n")
-
+```
+
+### Bootstrap to evaluate GBM robustness
+```{r}
 # Define a function to fit the model and calculate RMSE
@@
 
-# SUmmarize the RSME values
+# Summarize the RSME values
 ```{r}

### 2024-05-03 products/manuscript/Manuscript.qmd
Edits to the manuscript and code cleaning

@@
 
-Eastern equine encephalitis (EEE) is a rare but serious disease caused
+Eastern equine encephalitis (EEE) is a rare but severe disease caused
 by an Alphavirus transmitted to humans by infected mosquitoes. With a
@@
 involves two cycles: an enzootic cycle between Culiseta melanura
-mosquitoes and birds, and an epizootic cycle where bridge vector
-mosquitoes transmit the virus to humans and other mammals. While C.
-melanura poses minimal direct threat to humans, isolations of EEE virus
-in these mosquitoes serve as early indicators of virus circulation.
-Conversely, isolations in bridge vector species signal heightened
-transmission risk. This research aims to identify significant
-relationships between variables to inform dissertation prospectus
-development and guide methodological approaches.
+mosquitoes and birds and an epizootic cycle where bridge vector
+mosquitoes transmit the virus to humans and other mammals.
+While *C.
+melanura* poses minimal direct threat to humans, isolations of EEE virus in these mosquitoes serve as early indicators of virus circulation.
+In contrast, isolations in bridge vector species signal a heightened transmission risk.
+This project aims to identify significant relationships between variables to inform the development of a dissertation prospectus and guide methodological approaches..
 :::
 
-# Research question
+# Research Question
 
-Which environmental,ecological, and temporal factors influence the
-transmission cycles for EEE in southeastern Massachusetts, and which
-models best describe the relationship between these factors?
+In southeastern Massachusetts, the transmission dynamics of Eastern Equine Encephalitis (EEE) remain inadequately understood, particularly how environmental, ecological, and temporal factors interplay to affect outbreak patterns.
+This gap in knowledge hinders effective predictive modeling and response strategies, crucial for preventing and mitigating the impact of EEE.
+
+**Which environmental,ecological, and temporal factors influence the transmission cycles for EEE in southeastern Massachusetts, and which models best describe the relationship between these factors?**
 
@@
 to humans by the bite of an infected mosquito. Human infections are
-rare, but serious [@lindseyEasternEquineEncephalitis2018]. Humans and
+rare but serious [@lindseyEasternEquineEncephalitis2018]. Humans and
 other mammals are dead-end hosts for EEE. About 94% of human EEE
@@
 consequences can be severe. Mortality rates associated with symptomatic
-infections range from 30 to 50% [@morensEasternEquineEncephalitis2019].
+cases range from 30 to 50% [@morensEasternEquineEncephalitis2019].
 Those who survive often experience lifelong disability that includes
@@
 [@massachusettsdepartmentofpublichealthArbovirusSurveillancePlan2024].
-On average, there are 11 cases reported in the US annually. Most occur
+On average, 11 cases reported in the US annually. Most occur
 in Massachusetts and Florida [@vanderkelenRiskIndexModel2014]. Until
@@
 recommends preventive measures. Recommended measures include outreach
-and education to increase personal protective practices . Additionally,
+and education to increase personal protective practices. Additionally,
 MDPH may recommend that cities and towns in the affected areas cancel
@@
 west, Plymouth County, MA to the east, and Norfolk County, MA to the
-north. Naushon Island and Martha's Vineyard are across Buzzard's Bay to
-the Southeast.
+north. Naushon Island and Martha's Vineyard are across Buzzard's Bay.
 
@@
 crucial habitats for the larval growth of Culiseta melanura (MEL) and
-also provide shelter during the winter months.
+also provide shelter during the winter.
 
@@
 
-**Which environmental,ecological, and temporal factors influence the
-transmission cycles for EEE in southeastern Massachusetts, and which
-models best describe the relationship between these factors?** From
-prior research we know that the mosquito lifecyle is influenced by a
-number of factors. Temperature plays a major role in the mosquito life
-cyle virus transmission dynamics. As temperature increases, the rate of
-larval development also increases and the time between when a mosquito
-ingests virus in a bloodmeal and when it can transmit it decreases. The
-result is that the mosquito population can grow at faster rate and they
-will have more time to transmit the virus. However, MEL larvae develop
-in well-insulated crypts that are less sensitive to ambient air
-temperature changes, so the effect of temperature changes on MEL
-development speed may be less pronounced than other vector species that
-develop in open, standing water, like ditches and containers.
-Precipitation increases the amount of standing water and increases the
-chances of larval survival during the mosquito season and also during
-the winter months for those species that overwinter in the larval stage.
-For EEE, the main enzootic vector, MEL, overwinters as larvae.
-
-While these relationships are well known, they are not well described
-for EEE, especially in terms of quantifying the role these two factors
-play in EEE transmission. Specifically, there's a gap in the
-understanding of how these factors influence the number of EEE vectors
-mosquitoes collected in traps and the rate of virus infection among
-those mosquitoes.
+*Which environmental, ecological, and temporal factors influence the transmission cycles for EEE in southeastern Massachusetts, and which models best describe the relationship between these factors?* From prior research, we know that the mosquito lifecycle is influenced by several factors.
+Temperature plays a significant role in the mosquito life cycle and virus transmission dynamics.
+As temperature increases, the rate of larval development also increases, and the time between when a mosquito ingests the virus in a blood meal and when it can transmit it decreases.
+The result is that the mosquito population can grow at a faster rate, and they will have more time to transmit the virus.
+However, MEL larvae develop in well-insulated crypts that are less sensitive to ambient air temperature changes, so the effect of temperature changes on MEL development speed may be less pronounced than other vector species that develop in open, standing water, like ditches and containers.
+Precipitation increases the amount of standing water and the chances of larval survival during the mosquito season and winter for those species that overwinter in the larval stage.
+For EEE, the primary enzootic vector, MEL, overwinters as larvae.
+While these relationships are well known, they are not well described for EEE, especially in quantifying the role these two factors play in EEE transmission.
+Specifically, there’s a gap in understanding how these factors influence the number of EEE vector mosquitoes collected in traps and the rate of virus infection among those mosquitoes
 
@@
 
-The mosquito surveillance dataset encompasses counts by species from all
-mosquito traps deployed in Bristol County, MA, from 2007 to 2023. The
-raw dataset includes over 45,000 records and includes temporal and
-spatial variables. The BCMCP, perform mosquito surveillance from June
-through October annually.
+The mosquito surveillance dataset encompasses counts by species from all mosquito traps deployed in Bristol County, MA, from 2007 to 2023.
+The raw dataset includes over 45,000 records and includes temporal and spatial variables.
+The BCMCP performs mosquito surveillance from June through October annually.
+During this interval, CDC light traps baited with carbon dioxide are placed at various sentinel sites weekly.
+Designed to attract mosquitoes in search of blood meals, light traps serve as a reliable measure of the actively feeding mosquito population, thereby establishing them as the standard method for EEE surveillance.
+Some of the sentinel sites have been used for over 50 years.
+Additionally, the county employs gravid traps specifically aimed at attracting and capturing gravid Culex mosquitoes looking for egg-laying sites.
+While these traps are primarily used for West Nile Virus (WNV) surveillance, they also hold significant value for EEE monitoring.
+Given the role of Culex species as bridge vectors for EEE, their capture in gravid traps provides critical insight into the risk of virus transmission.
+The traps are installed in the afternoon and retrieved the following morning.
+After the traps are collected, the next are stored in a freezer to preserve and kill the mosquitoes.
+The mosquitoes are sorted by species using a dissecting scope.
+Counts for each species are recorded in the database, and a portion of the vector species are driven to the Massachusetts State Laboratory in Boston.
+The lab tests each mosquito pool (or batch) for WNV and EEEv using PCR.
+Results are shared with BCMCP on the same day of testing.
+Virus isolation results are recorded in the collection records.
+In addition to mosquito counts by species, the dataset includes the town or city in which they were collected, the collection date, and whether the mosquitoes were submitted for virus testing.
 
-During this period, CDC light traps, baited with carbon dioxide, are
-strategically placed at various sentinel sites weekly. These traps are
-designed to attract mosquitoes seeking blood meals and are recognized as
-a reliable method for monitoring the actively feeding mosquito
-population. Consequently, they are established as the standard approach
-for Eastern Equine Encephalitis (EEE) surveillance. Notably, some of
-these sentinel sites have been in continuous use for over 50 years.
-
-Additionally, the county deploys gravid traps, which are specifically
-aimed at attracting and capturing gravid *Culex* mosquitoes looking for
-egg-laying sites. While these traps are primarily used for West Nile
-Virus (WNV) surveillance, they hold significant value for EEE monitoring
-as well. Given the role of *Culex* species as bridge vectors for EEE,
-their capture in gravid traps provides critical insight into the risk of
-virus transmission.
-
-The traps are set up in the afternoon and collected the following
-morning. Once retrieved, the captured mosquitoes are immediately stored
-on dry ice and then in a freezer. This process kills the mosquito and
-preserves them. The mosquitoes are sorted by species under a dissecting
-scope. Counts for each species are recorded in the database and a
-portion of the vector species are driven to the Massachusetts State
-Laboratory in Boston. The lab tests each mosquito pool (or batch) for
-WNV and EEEv using PCR. Results are shared with BCMCP on the same day of
-testing. Virus isolation results are recorded in the collection records.
-
-In addition to mosquito counts by species, the dataset includes the town
-or city in which they were collected, the collection date, and whether
-the mosquitoes were submitted for virus testing.
-
-### Minimun infection rate (MIR)
+### Minimum infection rate (MIR)
 
@@
 MIR = \left( \frac{\text{Number of positive mosquito pools}}{\text{Total number of mosquitoes tested}} \right) \times 1000
-$$ While MIR does not directly measure the risk to humans, a higher MIR
-in mosquitoes that commonly bite humans can indicate a higher risk of
-transmission. A study pubulihsed in 2007 indicated that in years with
-human cases in Massachusetts between 1979 and 2004, the MIR was ≥1.0 by
-week 31 (mid-late July) [@hachiyaHumanEasternEquine2007]. MIR allows for
-the comparison of infection rates across different geographical areas or
-time periods. This is crucial for identifying areas or times of higher
-risk and for evaluating the effectiveness of control measures.
+$$ While MIR does not directly measure the risk to humans, a higher MIR in mosquitoes that commonly bite humans can indicate a higher risk of transmission.
+A study published in 2007 indicated that in years with human cases in Massachusetts between 1979 and 2004, the MIR was ≥1.0 by week 31 (mid-late July) [@hachiyaHumanEasternEquine2007].
+MIR allows for comparing infection rates across different geographical areas or time periods.
+This is crucial for identifying areas or times of higher risk and evaluating control measures' effectiveness.
 
@@
 
-The surveillance dataset is structured to conform with state and federal
-reporting requirements. As a result, the format of the data must be
-transformed to in order to make it ready for analysis. Once the data
-structure is corrected, it can be joined to other relevant datasets
-using spatial or temporal attributes.
+The surveillance dataset is structured to conform with state and federal reporting requirements.
+As a result, the data format must be transformed to make it ready for analysis.
+Once the data structure is corrected, it can be joined to other relevant datasets using spatial or temporal attributes.
 
@@
 variables, such as the number of hours above a certain temperature
-threshold. or degree-days. Degree-days is a measure of the amount of
+threshold or degree-days. Degree-days is a measure of the amount of
 heat accumulated over a period of time. It is calculated by summing the
@@
 provide a versatile framework for capturing complex, non-linear
-relationships between response variables and multiple predictors, which
-are often encountered in ecological data sets.
+relationships between response variables and multiple predictors, often encountered in ecological data sets.
 

### 2024-05-03 R/Library_script.R
README updates

@@ -1,0 +1,45 @@
+library(here)
+
+# Function to extract library names from a line of code
+extract_libraries <- function(line) {
+  matches <- gregexpr("library\\(([^)]+)\\)|require\\(([^)]+)\\)", line)
+  unlist(lapply(regmatches(line, matches), function(x) gsub("library\\(|require\\(|\\)", "", x)))
+}
+
+# List of paths to the Quarto documents you are editing
+qmd_paths <- list(
+  here("R", "analysis-code", "analysis.qmd"),
+  here("R", "eda-code", "EDA_unified.qmd"),
+  here("R", "processing-code", "MADAproject_pt2_ruiz.qmd")
+)
+
+# Function to process each qmd file and extract libraries
+process_qmd <- function(qmd_path) {
+  qmd_content <- readLines(qmd_path)
+  library_lines <- grep("library\\(|require\\(", qmd_content, value = TRUE)
+  library_names <- unique(unlist(lapply(library_lines, extract_libraries)))
+  library_names <- gsub("\"|'", "", library_names)  # Remove potential quotes
+  library_names <- trimws(library_names)  # Trim whitespace
+  library_names
+}
+
+# Apply the function to each qmd file and combine the results
+all_libraries <- unique(unlist(lapply(qmd_paths, process_qmd)))
+
+# Sort libraries alphabetically
+sorted_libraries <- sort(all_libraries)
+
+# Get versions of these libraries
+library_versions <- sapply(sorted_libraries, function(name) {
+  if (name %in% rownames(installed.packages())) {
+    as.character(packageVersion(name))
+  } else {
+    NA
+  }
+})
+
+# Combine names and versions into a data frame
+library_info <- data.frame(Library = sorted_libraries, Version = library_versions, stringsAsFactors = FALSE)
+
+# Print the library information in alphabetical order
+print(library_info)

### 2024-05-03 R/analysis-code/analysis.qmd
More README edits
Added script to exrtact libraries and versions for a list in the readme

@@
 
-# Define a more refined tuning grid for LASSO with a wider range of penalty values
+# Define tuning grid for LASSO with a wide range of penalty values
 tuning_grid_lasso <- expand.grid(
-  penalty = 10^seq(-10, 1, length.out = 50)  # Increase the range and resolution of the penalty parameter
-)
+  penalty = 10^seq(-10, 1, length.out = 50)
 
@@
 
-# Tune the model using the refined tuning grid and increased folds
+# Tune the model using the tuning grid and folds
 tuning_results_lasso <- tune_grid(
@@
 )
+```
+
+### Extract the fitted model and calculate RMSE for the training and testing sets
+```{r}
 
@@
 autoplot(tuning_results_lasso)
-
 ```

### 2024-05-03 R/Library_script.R

@@ -20,7 +20,7 @@ process_qmd <- function(qmd_path) {
   library_names <- unique(unlist(lapply(library_lines, extract_libraries)))
   library_names <- gsub("\"|'", "", library_names)  # Remove potential quotes
   library_names <- trimws(library_names)  # Trim whitespace
-  library_names
+  unique(library_names)  # Ensure unique names
 }
 
 # Apply the function to each qmd file and combine the results
@@ -41,5 +41,8 @@ library_versions <- sapply(sorted_libraries, function(name) {
 # Combine names and versions into a data frame
 library_info <- data.frame(Library = sorted_libraries, Version = library_versions, stringsAsFactors = FALSE)
 
-# Print the library information in alphabetical order
-print(library_info)
+# Save to CSV file
+write.csv(library_info, here("R", "library_versions.csv"), row.names = FALSE)
+
+# Message to inform about CSV creation
+cat("Library and version information saved to: ", here("results", "library_versions.csv"), "\n")
